---
title: "GRN"
params: 
  file_p2: "Matrix/CLL_annotated_Seurat_LogNorm_P2.rds"
  file_p3: "Matrix/CLL_annotated_Seurat_LogNorm_P3.rds"
  file_step2_p2: "Matrix/step_2_reg_p2.csv"
  file_step3_p2: "Matrix/step_3_aucell_p2.csv"
  file_step2_p3: "Matrix/step_2_reg_p3.csv"
  file_step3_p3: "Matrix/step_3_aucell_p3.csv"
format: html
editor: visual
---

## Data and libraries initialization

```{r}
#| message: false
library(SeuratObject)
library(SeuratDisk)
library(sp)
library(Seurat)
library(hdf5r)
library(dplyr)
library(patchwork)
library(ggplot2)
library(stringr)
library(pheatmap)
library(viridis)
library(gridExtra)
library(SeuratExtend)
library(AUCell)
library(clusterProfiler)
library(org.Hs.eg.db)
source("functions.R")
```

Read the data pre-processed

```{r}
data_p2<-readRDS(params$file_p2)
Idents(data_p2) <- data_p2$final_annot
data_p2<-subset(data_p2, subset = (final_annot %in% c("B","B intermediate","B memory","B naive","Macrophage")))

data_p3<-readRDS(params$file_p3)
Idents(data_p3) <- data_p3$final_annot
data_p3<-subset(data_p3, subset = (final_annot %in% c("B","B intermediate","B memory","B naive","Macrophage")))
```

```{r}
SaveLoom(data, filename="TimeCourse.loom")
```

### Apply pySCENIC in the Genotoul cluster with the script "pySCENIC.sh"

Import pySCENIC output

```{r}
# TF and targets
regulons_p2 <- read.csv(params$file_step2_p2)
regulons_p3 <- read.csv(params$file_step2_p3)

# Loom output in seurat object
#scenic_output<-"/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/out_TimeCourse.loom"
#data <- ImportPyscenicLoom(scenic_output, seu = data)
```

Bridge annotation P2

```{r}
umap_coords <- Embeddings(data_p2, "umap")
umap_df <- as.data.frame(umap_coords)
colnames(umap_df)
umap_df$cell <- rownames(umap_df)
colnames(umap_df)

x_min <- 2.5
x_max <- 6
y_min <- 4.5
y_max <- 7

umap_df$selected <- ifelse(
  umap_df$umap_1 >= x_min & umap_df$umap_1 <= x_max &
  umap_df$umap_2 >= y_min & umap_df$umap_2 <= y_max,
  "Intermediate population",  
  NA      
)

# Add "bridge" on final annotation
umap_df <- umap_df %>%
  mutate(
    bridge_annot = ifelse(is.na(selected), data_p2$final_annot[cell], selected)
  )

data_p2$bridge_annot <- umap_df$bridge_annot[match(colnames(data_p2), umap_df$cell)]

table(data_p2$bridge_annot)

# Plot UMAP
umap_df$bridge_annot <- data_p2$bridge_annot[match(umap_df$cell, colnames(data_p2))]

plotumap<-ggplot(umap_df, aes(x = umap_1, y = umap_2, color = bridge_annot)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("Intermediate population" = "red", "B" = "grey", "Macrophages" = "grey", "B naive" = "grey", "B memory" = "grey", "B intermediate" = "grey")) +
  theme_minimal() +
  ggtitle("Bridge B cells and macrophages")
plotumap
```

P3

```{r}
umap_coords <- Embeddings(data_p3, "umap")
umap_df <- as.data.frame(umap_coords)
colnames(umap_df)
umap_df$cell <- rownames(umap_df)
colnames(umap_df)

x_min <- 3.5
x_max <- 7.5
y_min <- -2.5
y_max <- 0

umap_df$selected <- ifelse(
  umap_df$umap_1 >= x_min & umap_df$umap_1 <= x_max &
  umap_df$umap_2 >= y_min & umap_df$umap_2 <= y_max,
  "Intermediate population",  
  NA      
)

# Add "bridge" or final annotation
umap_df <- umap_df %>%
  mutate(
    bridge_annot = ifelse(is.na(selected), data_p3$final_annot[cell], selected)
  )

data_p3$bridge_annot <- umap_df$bridge_annot[match(colnames(data_p3), umap_df$cell)]

table(data_p3$bridge_annot)

# Plot UMAP
umap_df$bridge_annot <- data_p3$bridge_annot[match(umap_df$cell, colnames(data_p3))]

plotumap<-ggplot(umap_df, aes(x = umap_1, y = umap_2, color = bridge_annot)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("Intermediate population" = "red", "B" = "grey", "Macrophages" = "grey", "B naive" = "grey", "B memory" = "grey", "B intermediate" = "grey")) +
  theme_minimal() +
  ggtitle("Bridge B cells and macrophages")
plotumap
```

## Patient 2 analysis 

```{r}
tf_auc <- read.csv(params$file_step3_p2, row.names = 1, check.names=FALSE)
colnames(tf_auc) <- gsub("[(+)]", "", colnames(tf_auc))
celltypes<-data_p3@meta.data[["bridge_annot"]]
celltypes<-ifelse(celltypes=="B intermediate"|celltypes=="B naive"|celltypes=="B memory"|celltypes=="B","B cells", celltypes)
tf_zscore <- CalcStats(tf_auc, f = celltypes, order = "p", n = 15, t = TRUE) # change the n (n*cluster=TF number)
plot<-Heatmap(tf_zscore, lab_fill = "zscore")+ ggtitle("Transcription factor activity (z-score) across B cells,\n intermediate population, and macrophages, P2")
plot
ggsave("Plots/heatmap_tf_P2.jpeg", plot, width = 1500, height = 2000, units = "px")
```

```{r}
celltypes<-data_p2@meta.data[["bridge_annot"]]
tf_zscore <- CalcStats(tf_auc, f = celltypes, order = "p", n = 6, t = TRUE) # change the n (n*cluster=TF number)
plot<-Heatmap(tf_zscore, lab_fill = "zscore")
plot
```

Waterfall plot (CLL-macro vs B, CLL-macro vs macro)

```{r}
plot<-WaterfallPlot(t(tf_auc), f = celltypes,
  ident.1 = "Intermediate population",      # First group of cells
  ident.2 = "Macrophage",     # Second group of cells
  exp.transform = FALSE,      # Disable transformation of expression data
  top.n = 20,                  # Display the top 20 most differentially active TFs
  length="logFC", 
  log.base="2", 
  color="p"
) + ggtitle("Differential transcription factor activity: \nintermediate population vs macrophages, P2")
plot
ggsave("Plots/waterfallplot_macroxbridge_p2.jpeg", plot, width = 2000, height = 2000, units = "px")
```

```{r}
plot<-WaterfallPlot(t(tf_auc), f = celltypes,
  ident.1 = "Intermediate population",      # First group of cells
  ident.2 = "B cells",     # Second group of cells
  exp.transform = FALSE,      # Disable transformation of expression data
  top.n = 20,                  # Display the top 20 most differentially active TFs
  length="logFC", 
  log.base="2", 
  color="p"
) + ggtitle("Differential transcription factor activity: \nintermediate population vs B cells, P2")
plot
ggsave("Plots/waterfallplot_Bxbridge_p2.jpeg", plot, width = 2000, height = 2000, units = "px")
```

TF Regulon Activity (AUCell)

```{r}
# AUC score dataframe
auc_mtx <- read.csv(params$file_step3_p2, row.names = 1, check.names=FALSE)
auc_mtx <- t(auc_mtx)
auc_mtx <- as.data.frame(auc_mtx)
rownames(auc_mtx) <- gsub("[(+)]", "", rownames(auc_mtx))
auc_mtx[] <- lapply(auc_mtx, as.numeric)
auc_mtx <- as.matrix(auc_mtx)

data_p2[['TF']] <- CreateAssayObject(data = auc_mtx)
DefaultAssay(data_p2) <- 'RNA'

plot<-DimPlot2(
  data_p2,
  features = c("tf_CEBPD", "tf_NFIX"),
  ncol = 2,
  cols=list("tf_CEBPD" = "D", "tf_NFIX" = "D"),
  theme = NoAxes(), 
  reduction = "umap",
  label.size=2
)
plot
ggsave("Plots/tf_activity_P2.jpeg", plot, width = 2000, height = 1000, units = "px")
```

Heatmap

```{r}
groups <- data_p2@meta.data[["bridge_annot"]]
identical(colnames(auc_mtx), colnames(data_p2))
annotation_df <- data.frame(Cell_types = data_p2@meta.data[["bridge_annot"]])
rownames(annotation_df) <- colnames(auc_mtx)

# Keep the most variable TFs
var_auc <- apply(auc_mtx, 1, var)
top_var_regulons <- names(sort(var_auc, decreasing = TRUE))[1:20]
auc_mat_top <- auc_mtx[top_var_regulons, ]

my_colors <- list(
  Cell_types = c(B = "darkgreen","B intermediate" = "darkgreen","B memory" = "darkgreen","B naive" = "darkgreen", "CD4 T" = "lightblue", "CD8 T" = "lightblue", "T" = "lightblue", "Macrophage"= "red","Monocyte"="orange", "Neutrophil"="brown", "NK"="darkblue", "Erythroblast"="grey" )
)

# Scaled matrix 
auc_mat_scaled <- t(scale(t(auc_mat_top)))

# Heatmap
pheatmap(
  auc_mat_top,
  cluster_cols = TRUE,
  cluster_rows = TRUE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  annotation_col = annotation_df, 
  annotation_colors = my_colors)
```

## Patient 3 analysis 

```{r}
tf_auc <- read.csv(params$file_step3_p3, row.names = 1, check.names=FALSE)
colnames(tf_auc) <- gsub("[(+)]", "", colnames(tf_auc))
celltypes<-data_p3@meta.data[["bridge_annot"]]
celltypes<-ifelse(celltypes=="B intermediate"|celltypes=="B naive"|celltypes=="B memory"|celltypes=="B","B cells", celltypes)
tf_zscore <- CalcStats(tf_auc, f = celltypes, order = "p", n = 15, t = TRUE) # change the n (n*cluster=TF number)
plot<-Heatmap(tf_zscore, lab_fill = "zscore")+ ggtitle("Transcription factor activity (z-score) across B cells,\n intermediate population, and macrophages, P3")
plot
ggsave("Plots/heatmap_tf_P3.jpeg", plot, width = 1500, height = 2000, units = "px")
```

```{r}
celltypes<-data_p3@meta.data[["bridge_annot"]]
tf_zscore <- CalcStats(tf_auc, f = celltypes, order = "p", n = 6, t = TRUE) # change the n (n*cluster=TF number)
plot<-Heatmap(tf_zscore, lab_fill = "zscore")
plot
```

Waterfall plot (CLL-macro vs B, CLL-macro vs macro)

```{r}
plot<-WaterfallPlot(t(tf_auc), f = celltypes,
  ident.1 = "Intermediate population",      # First group of cells
  ident.2 = "Macrophage",     # Second group of cells
  exp.transform = FALSE,      # Disable transformation of expression data
  top.n = 20,                  # Display the top 20 most differentially active TFs
  length="logFC", 
  log.base="2", 
  color="p"
) + ggtitle("Differential transcription factor activity: \nintermediate population vs macrophages, P3")
plot
ggsave("Plots/waterfallplot_macroxbridge_p3.jpeg", plot, width = 2000, height = 2000, units = "px")
```

```{r}
plot<-WaterfallPlot(t(tf_auc), f = celltypes,
  ident.1 = "Intermediate population",      # First group of cells
  ident.2 = "B cells",     # Second group of cells
  exp.transform = FALSE,      # Disable transformation of expression data
  top.n = 20,                  # Display the top 20 most differentially active TFs
  length="logFC", 
  log.base="2", 
  color="p"
) + ggtitle("Differential transcription factor activity: \nintermediate population vs B cells, P3")
plot
ggsave("Plots/waterfallplot_Bxbridge_p3.jpeg", plot, width = 2000, height = 2000, units = "px")
```

TF Regulon Activity (AUCell)

```{r}
# AUC score dataframe
auc_mtx <- read.csv(params$file_step3_p3, row.names = 1, check.names=FALSE)
auc_mtx <- t(auc_mtx)
auc_mtx <- as.data.frame(auc_mtx)
rownames(auc_mtx) <- gsub("[(+)]", "", rownames(auc_mtx))
auc_mtx[] <- lapply(auc_mtx, as.numeric)
auc_mtx <- as.matrix(auc_mtx)

data_p3[['TF']] <- CreateAssayObject(data = auc_mtx)
DefaultAssay(data_p3) <- 'RNA'

plot<-DimPlot2(
  data_p3,
  features = c("tf_CEBPD", "tf_NFIX"),
  ncol = 2,
  cols=list("tf_CEBPD" = "D", "tf_NFIX" = "D"),
  theme = NoAxes(), 
  reduction = "umap",
  label.size=2
)
plot
ggsave("Plots/tf_activity_P3.jpeg", plot, width = 2000, height = 1000, units = "px")
```

Heatmap

```{r}
groups <- data_p3@meta.data[["bridge_annot"]]
identical(colnames(auc_mtx), colnames(data_p3))
annotation_df <- data.frame(Cell_types = data_p3@meta.data[["bridge_annot"]])
rownames(annotation_df) <- colnames(auc_mtx)

# Keep the most variable TFs
var_auc <- apply(auc_mtx, 1, var)
top_var_regulons <- names(sort(var_auc, decreasing = TRUE))[1:20]
auc_mat_top <- auc_mtx[top_var_regulons, ]

my_colors <- list(
  Cell_types = c(B = "darkgreen","B intermediate" = "darkgreen","B memory" = "darkgreen","B naive" = "darkgreen", "CD4 T" = "lightblue", "CD8 T" = "lightblue", "T" = "lightblue", "Macrophage"= "red","Monocyte"="orange", "Neutrophil"="brown", "NK"="darkblue", "Erythroblast"="grey" )
)

# Scaled matrix 
auc_mat_scaled <- t(scale(t(auc_mat_top)))

# Heatmap
pheatmap(
  auc_mat_top,
  cluster_cols = TRUE,
  cluster_rows = TRUE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  annotation_col = annotation_df, 
  annotation_colors = my_colors)
```

```{r}
TF_list<-tf_zscore[order(tf_zscore$`Intermediate population`, decreasing = T),]
TF_list<-head(TF_list, 15)
TF_list<-rownames(TF_list)

enrich_tf <- enrichGO(
  gene          = TF_list,     # noms des TFs les plus actifs
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",        # Biological Process
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.05
)
```

Binaryzation

```{r}

```
