---
title: "Deconvolution"
params: 
  Data_bulk: "~/Documents/Singlecell-R/Scripts/Data"
  Data_CITEseq: "~/Documents/Singlecell-R/Scripts/Matrix/CLL_Seurat_merge_p2p3.rds"
  Output: "~/Documents/Singlecell-R/Scripts/Plots/"
format: html
editor: visual
---

Goal: To investigate the transdifferentiation population in various bulk datasets from CLL, Richter’s transformation (RT) and other conditions, with a view to linking this population to clinical data. To investigate using bulk data from PBMCs as well as lymph nodes.

```{r}
#| message: false
library(Seurat)
library(ggplot2)
library(dplyr)
#library(multideconv)
devtools::load_all("/home/a.blanc-boekholt/Documents/Singlecell-R/multideconv")
#source("functions.R")

# Seurat object combining the two patients
data_bulk<-readRDS(params$Data_bulk)
data_cite<-readRDS(params$Data_CITEseq)
data_cite<-subset(data_cite, subset = (Annotation %in% c("B intermediate", "B naive", "B memory", "B", "Bridge","Monocyte","Macrophage")))
```

Create custom signatures based on the CITE-seq data (Patients 2 and 3) for the "intermediate population".

```{r}
# Create metacells because ~110k cells 
data_cite$sample_time <- paste(data_cite$patient, data_cite$timepoint, sep = "_")
metacells = create_metacells(data_cite,
                             labels_column = "Annotation", 
                             samples_column = "sample_time", 
                             exclude_cells = NULL,
                             min_cells = 50, 
                             k = 15, 
                             max_shared = 15, 
                             n_workers = 1, 
                             min_meta = 10)

metacells_seurat = Seurat::CreateSeuratObject(metacells[["Counts"]], meta.data = metacells[["Metadata"]])

# Calculate the top markers
Idents(data_cite) <- "Annotation"
markers <- FindAllMarkers(data_cite,assay = "RNA",slot = "data", only.pos = TRUE, min.pct = 0.1, logfc.threshold = 0.8)
# Filter significant markers
top_markers <- markers[markers$p_val_adj < 0.05, ]
# Split marker genes by cluster
top_markers_list <- split(top_markers$gene, top_markers$cluster)

# Pseudobulk of my single-cell data
pseudobulk = create_sc_pseudobulk(metacells_seurat, cells_labels = "Annotation", sample_labels = "sample_time", normalized = TRUE, file_name = "Pseudobulk_deconvolution")

# Create a signature using Multideconv 
metacell_metadata<-metacells_seurat@meta.data
signatures = create_sc_signatures(metacells[["Counts"]], 
                                  metacell_metadata, 
                                  cells_labels = "Annotation", 
                                  sample_labels = "sample_time", 
                                  bulk_rna = pseudobulk, 
                                  cell_markers = top_markers_list, 
                                  name_signature = "CITE_CLL", 
                                  credentials.mail = "sushant.parab@unito.it", 
                                  credentials.token = "2834897837d47d520bffc8610c69b8cf",
                                  methods_sig = c("CIBERSORTx"))

# Verify the signature
#1) deconvolution pseudobulk 
deconv_pseudo = compute.deconvolution(raw.counts = pseudobulk,
                               normalized = FALSE,
                               #signatures_select = c("", ""),
                               return = TRUE, credentials.mail = "sushant.parab@unito.it", 
                               credentials.token = "2834897837d47d520bffc8610c69b8cf",
                               file_name = "pseudobulk")
#2) # Counting real cell proportion 
cells_groundtruth <- table(metacells_seurat$sample_time, metacells_seurat$Annotation)
cells_groundtruth <- prop.table(cells_groundtruth, margin = 1)
cells_groundtruth <- as.data.frame.matrix(cells_groundtruth)
colnames(cells_groundtruth) <- gsub(" ", ".", colnames(cells_groundtruth))
colnames(cells_groundtruth) <- dplyr::recode(
  colnames(cells_groundtruth),
  "B cell" = "B.cells",
  "NK cell" = "NK.cells",
  "T cell" = "T.cells",
  "Fibroblast" = "Fibroblasts",
  "pDC" = "Dendritic.cells",
   "Neuroendocrine" = "Neuroendocrine.cells",
  "Schwann" = "Schwann.cells",
)

#3) Compute deconvolution benchmark
benchmark = compute.benchmark(deconv_pseudo, 
                              cells_groundtruth, 
                              corr_type = "spearman",
                              scatter = FALSE, 
                              plot = TRUE, 
                              pval = 0.05, 
                              file_name = "Deconvolution", 
                              width = 10, 
                              height = 15)

```

Deconvolution of CLL bulk dataset with custom signature

```{r}
# Deconvolution 
deconv_bulk = compute.deconvolution(raw.counts = data_bulk,
                               normalized = FALSE,
                               #signatures_select = c("", ""),
                               return = TRUE,
                               file_name = "Deconvolution_bulk_CLL",  doParallel = TRUE, workers = 2)

deconv_subgroups = compute.deconvolution.analysis(deconvolution = deconv_bulk,
                                                  corr = 0.7,
                                                  seed = 123,
                                                  file_name = "Bulk_CLL",
                                                  return = TRUE)
```
