---
title: "Deconvolution"
params: 
  Data_bulk: "~/Documents/Singlecell-R/Scripts/Data/Dataset CLL-map (Knisbacher, B. A., et al. 2022)/cllmap_rnaseq_counts_full.tsv"
  Clinical: "~/Documents/Singlecell-R/Scripts/Data/Dataset CLL-map (Knisbacher, B. A., et al. 2022)/participants.nwx2hb.tsv"
  Data_CITEseq: "~/Documents/Singlecell-R/Scripts/Matrix/CLL_Seurat_P2P3.rds"
  Output: "~/Documents/Singlecell-R/Scripts/Plots/"
format: html
editor: visual
---

Goal: To investigate the transdifferentiation population in various bulk datasets from CLL, Richter’s transformation (RT) and other conditions, with a view to linking this population to clinical data. To investigate using bulk data from PBMCs as well as lymph nodes.

```{r}
#| message: false
library(Seurat)
library(ggplot2)
library(dplyr)
#library(multideconv)
devtools::load_all("/home/a.blanc-boekholt/Documents/Singlecell-R/multideconv")
#source("functions.R")

# CITE seq data patients 2 and 3 merged --> for the Intermediate population signatures 
data_cite<-readRDS(params$Data_CITEseq)
#data_cite<-subset(data_cite, subset = Annotation != "Erythroblast")
data_cite$Annotation<-ifelse(data_cite$Annotation %in% c("B intermediate", "B naive", "B memory", "B"), "B cells", data_cite$Annotation)
data_cite$Annotation<-ifelse(data_cite$Annotation %in% c("CD4 T", "CD8 T", "T"), "T cells", data_cite$Annotation)
```

## 1) Create custom signatures based on the CITE-seq data (Patients 2 and 3) for the "intermediate population".

```{r}
# Create metacells because ~110k cells 
data_cite$sample_time <- paste(data_cite$patient, data_cite$timepoint, sep = "_")
metacells = create_metacells(data_cite,
                             labels_column = "Annotation", 
                             samples_column = "sample_time", 
                             exclude_cells = NULL,
                             min_cells = 50, 
                             k = 15, 
                             max_shared = 15, 
                             n_workers = 1, 
                             min_meta = 10)

metacells_seurat = Seurat::CreateSeuratObject(metacells[["Counts"]], meta.data = metacells[["Metadata"]])
metacells_seurat <- NormalizeData(metacells_seurat, normalization.method = "LogNormalize")

# Calculate the top markers
Idents(metacells_seurat) <- "Annotation"
markers <- FindAllMarkers(metacells_seurat, assay = "RNA", slot = "data", only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.8, max.cells.per.ident = 1000)
# Filter significant markers
top_markers <- markers[markers$p_val_adj < 0.05, ]
# Split marker genes by cluster
top_markers_list <- split(top_markers$gene, top_markers$cluster)

# Evaluate/Checking specificity of the signatures 
fc_col <- intersect(c("avg_log2FC", "logFC", "avg_logFC"), names(top_markers))[1]

markers_scored <- top_markers %>%
  mutate(
    fc = .data[[fc_col]],
    specificity = (pct.1 - pct.2) * pmax(fc, 0),
    expr_diff = pct.1 - pct.2,
    support = pct.1 * fc,
    robust = ifelse(p_val_adj < 0.05 & pct.1 > 0.25, TRUE, FALSE)
  ) %>%
  arrange(cluster, desc(specificity), desc(fc), desc(expr_diff))

best_markers <- markers_scored %>%
  group_by(cluster) %>%
  slice_head(n = 40) %>%
  ungroup()
# Split marker genes by cluster
top_markers_list <- split(best_markers$gene, best_markers$cluster)


DotPlot(object = metacells_seurat,
    features = unique(best_markers$gene),
    group.by = "Annotation") +
  RotatedAxis()+labs(y = NULL, x = "Genes") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

Validate signatures

```{r}
# Pseudobulk of my single-cell data
pseudobulk = create_sc_pseudobulk(metacells_seurat, cells_labels = "Annotation", sample_labels = "sample_time", normalized = TRUE, file_name = "Pseudobulk_deconvolution")

# Create a signature using Multideconv 
metacell_metadata<-metacells_seurat@meta.data
signatures = create_sc_signatures(metacells[["Counts"]], 
                                  metacell_metadata, 
                                  cells_labels = "Annotation", 
                                  sample_labels = "sample_time", 
                                  bulk_rna = pseudobulk, 
                                  cell_markers = top_markers_list, 
                                  name_signature = "CITE_CLL", 
                                  credentials.mail = "sushant.parab@unito.it", 
                                  credentials.token = "2834897837d47d520bffc8610c69b8cf",
                                  methods_sig = c("BSeqsc"))

# Verify the signature
#1) deconvolution pseudobulk 
deconv_pseudo = compute.deconvolution(raw.counts = pseudobulk,
                               normalized = FALSE,
                               signatures_select = c("BSeqSC-CITE_CLL-scRNAseq"),
                               return = TRUE, credentials.mail = "sushant.parab@unito.it", 
                               credentials.token = "2834897837d47d520bffc8610c69b8cf",
                               file_name = "pseudobulk")

#2) # Counting real cell proportion 
cells_groundtruth <- table(metacells_seurat$sample_time, metacells_seurat$Annotation)
cells_groundtruth <- prop.table(cells_groundtruth, margin = 1)
cells_groundtruth <- as.data.frame.matrix(cells_groundtruth)
colnames(cells_groundtruth) <- dplyr::recode(
  colnames(cells_groundtruth),
  "B cells" = "B.cells",
  "Macrophage" = "Macrophages.cells",
  "Monocyte" = "Monocytes",
  "Neutrophil" = "Neutrophils",
  "NK" = "NK.cells",
  "T cells" = "T.cells",
  "CD4 T" = "CD4.T",
  "CD8 T" = "CD8.T"
)

#3) Compute deconvolution benchmark
benchmark = compute.benchmark(deconv_pseudo, 
                              cells_groundtruth, 
                              corr_type = "pearson",
                              scatter = FALSE, 
                              plot = TRUE, 
                              pval = 0.05, 
                              file_name = "Deconvolution", 
                              width = 10, 
                              height = 5)
```

Optimization signature

```{r}

# Autre methode 
Idents(data_cite) <- "Annotation"
neighbours <- c("B cells", "Monocyte", "Macrophage")

bridge_vs_each <- lapply(neighbours, function(ct) {
  if (!(ct %in% levels(Idents(data_cite)))) return(NULL)
  FindMarkers(data_cite, 
              ident.1 = "Bridge", ident.2 = ct,
              assay = "RNA", slot = "data",
              only.pos = TRUE, 
              min.pct = 0.25, 
              logfc.threshold = 0.8,
              min.diff.pct = 0.15)
  })
names(bridge_vs_each) <- neighbours
bridge_vs_each <- Filter(Negate(is.null), bridge_vs_each)

# Garder seulement les gènes significatifs contre TOUS les voisins simultanément
# Liste des gènes significatifs pour chaque comparaison
gene_lists <- lapply(bridge_vs_each, function(x) {
  rownames(subset(x, p_val_adj < 0.05))
})

# Compter dans combien de comparaisons chaque gène est significatif
gene_counts <- table(unlist(gene_lists))
# Garder les gènes présents dans au moins 4 comparaisons
bridge_specific <- names(gene_counts[gene_counts >= 1])

length(bridge_specific)



#Idents(data_cite) <- "Annotation"

celltypes <- c("B cells", "Bridge", "Monocyte", "Macrophage")

signature_list <- list()

for (ct in celltypes) {

  neighbours <- setdiff(celltypes, ct)

  markers_vs_each <- lapply(neighbours, function(nb) {

    FindMarkers(
      object = data_cite,
      ident.1 = ct,
      ident.2 = nb,
      assay = "RNA",
      slot = "data",
      only.pos = TRUE,
      min.pct = 0.25,
      logfc.threshold = 0.8,
      min.diff.pct = 0.15
    )

  })

  names(markers_vs_each) <- neighbours

  ## Gènes significatifs pour chaque comparaison
  gene_lists <- lapply(markers_vs_each, function(x) {
    rownames(subset(x, p_val_adj < 0.05))
  })

  ## Nombre de comparaisons où chaque gène est significatif
  gene_counts <- table(unlist(gene_lists))

  ## Gènes significatifs contre TOUS les voisins
  signature_list[[ct]] <- names(
    gene_counts[gene_counts == length(neighbours)]
  )

}

#Autres
Idents(data_cite) <- "Annotation"
celltypes <- levels(Idents(data_cite))
top_markers_list <- list()

for(ct in celltypes){
    markers <- FindMarkers(
        data_cite,
        ident.1 = ct,
        assay = "RNA",
        slot = "data",
        only.pos = TRUE,
        min.pct = 0.25,
        logfc.threshold = 0.25)

    markers <- markers[markers$p_val_adj < 0.05,]
    markers <- markers[order(markers$avg_log2FC,decreasing=TRUE),]
    top_markers_list[[ct]] <- rownames(head(markers,50))

}

# Combien de gènes Bridge partage avec chaque voisin ?
neighbours <- c("B intermediate", "B naive", "B memory", "B", "Monocyte", "Macrophage")

overlap_check <- sapply(neighbours, function(ct) {
     if (is.null(top_markers_list[[ct]])) return(NA)
     n_overlap <- length(intersect(top_markers_list$Bridge, top_markers_list[[ct]]))
     n_overlap / length(top_markers_list$Bridge)
 })
print(overlap_check)  # proportion de gènes Bridge partagés avec chaque type
```

## 2) Deconvolution of CLL bulk dataset with custom signature

#### PBMC: Dataset CLL-map (Knisbacher, B. A., et al. 2022)

708 samples + clinical data

```{r}
# Bulk Data for the deconvolution 
data_bulk<-read.table(params$Data_bulk, header=T)
data_bulk<-data_bulk[,-1]
cols_to_remove <- c("CRC.0007","CRC.0011","CRC.0028", "CRC.0033", "DFCI.5053", "JB.0010", "GCLL.0136")
data_bulk <- data_bulk[, !colnames(data_bulk) %in% cols_to_remove]
data_bulk$variance <- apply(data_bulk[, -1], 1, var)
# keep gene with the higher variance
data_bulk <- data_bulk %>%
  group_by(Description) %>%
  slice_max(variance, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(-variance)
data_bulk <-data.frame(data_bulk)
genes <- data_bulk$Description
data_bulk$Description <- NULL
rownames(data_bulk) <- genes
data_bulk[] <- lapply(data_bulk, as.numeric)
data_bulk <- as.matrix(data_bulk)

# Clinical data 
clinical<-read.table(params$Clinical, sep="\t", header=TRUE, fill = TRUE)
clinical$participant_id<-gsub("-", ".", clinical$participant_id)

id<-intersect(colnames(data_bulk),clinical$participant_id)

clinical<-subset(clinical, participant_id %in% id)
data_bulk <- data_bulk[, colnames(data_bulk) %in% id]
```

```{r}
# Deconvolution 
deconv_bulk = compute.deconvolution(raw.counts = data_bulk,
                               normalized = TRUE, 
                               methods = c("Epidish","DWLS", "CBSX"),
                               signatures_select = c("BSeqSC-CITE_CLL-scRNAseq"),
                               return = TRUE, credentials.mail = "marcelo.hurtado@inserm.fr",
                               credentials.token = "b72981d4d0554bcfffc37472ccd8588d",
                               file_name = "Deconvolution_bulk_CLL")

deconv_subgroups = compute.deconvolution.analysis(deconvolution = deconv_bulk,
                                                  corr = 0.7,
                                                  seed = 123,
                                                  file_name = "Bulk_CLL",
                                                  return = TRUE)

deconv_cite<-deconv_bulk[,colnames(deconv_bulk) %in% c("DWLS_BSeqSC.CITE_CLL.scRNAseq_Bridge","CBSX_BSeqSC.CITE_CLL.scRNAseq_Bridge", "Epidish_BSeqSC.CITE_CLL.scRNAseq_Bridge","DeconRNASeq_BSeqSC.CITE_CLL.scRNAseq_Bridge")]
correlation_bridge<-cor(deconv_cite,use = "complete.obs")
colnames(correlation_bridge)<-gsub("_BSeqSC.CITE_CLL.scRNAseq_", ".", colnames(correlation_bridge))
rownames(correlation_bridge)<-gsub("_BSeqSC.CITE_CLL.scRNAseq_", ".", rownames(correlation_bridge))
```

```{r}
# Heatmap proportion 
pheatmap(t(deconv_bulk), cluster_rows = TRUE, cluster_cols = TRUE, show_colnames = FALSE, annotation_col = annotation)
# Heatmap z-score proportion 
mat_z <- t(scale(t(deconv_bulk)))
pheatmap(t(mat_z), cluster_rows = TRUE, cluster_cols = TRUE, show_colnames = FALSE, annotation_col = annotation)


# Without B cells pop 
deconv_bcells<-deconv_bulk[, !colnames(deconv_bulk) %in% c("DWLS_BSeqSC.CITE_CLL.scRNAseq_B.cells","CBSX_BSeqSC.CITE_CLL.scRNAseq_B.cells", "Epidish_BSeqSC.CITE_CLL.scRNAseq_B.cells","DeconRNASeq_BSeqSC.CITE_CLL.scRNAseq_B.cells")]
# Heatmap z-score proportion 
mat_z <- t(scale(t(deconv_bcells)))
rownames(clinical) <- clinical$participant_id
annotation <- clinical[colnames(t(mat_z)), c("IGL_expression", "vital_status", "tumor_molecular_subtype", "cll_epitype","IGHV_mut","treatment_status_at_samp")]
pheatmap(t(mat_z), cluster_rows = TRUE, cluster_cols = TRUE, show_colnames = FALSE, annotation_col = annotation)


# Comparison clinical data 
# Fusion clinical data and deconv results 
data<-merge()
# Mean 3 deconv methods 

# treated and untreated, alive and dead, mutation, rai (cancer stage)
comparisons <- list(c("Treated", "Untreated"))
plot<-ggplot(data, aes(x = treatment_status_at_samp, y = Bridge)) + geom_violin(trim = FALSE, alpha = 0.6) + geom_boxplot(width = 0.1, outlier.shape = NA) + stat_compare_means(comparisons = comparisons, method = "wilcox.test", size=6) + theme_classic() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size=16), axis.text.y=element_text(size = 14), axis.title.x = element_text(size = 18),axis.title.y = element_text(size = 18), legend.title = element_text(size = 18),
legend.text = element_text(size = 18) ) + xlab("Treatment status") + ylab("Intermediate population deconvolution")
plot

# Survival curve
library(survival)
library(survminer)

# Définir OS_event, status = 1 décès/événement, 0 censuré --> OS_event

# Kaplan-Meirl curve (Are the survival curves for the two groups significantly different? test log-rank)
clinical$Bridge_group <- ifelse(clinical$Bridge_mean >= median(clinical$Bridge_mean_bin),"High","Low")
surv_obj <- Surv(time = clinical$death_days_from_samp,event = clinical$OS_event)
fit <- survfit(surv_obj ~ Bridge_group, data = clinical)

ggsurvplot(fit, data = clinical,
    pval = TRUE,
    risk.table = TRUE,
    conf.int = TRUE,
    palette = c("#2E9FDF", "#E7B800"))

# Cox model (Is the proportion of Intermediate population associated with the risk of death?)
cox <- coxph(Surv(death_days_from_samp, OS_event) ~ Bridge_mean, data = clinical)

summary(cox)
cox <- coxph(Surv(death_days_from_samp, OS_event) ~
        Bridge_mean , data = clinical)
```
