---
title: "Boolean model"
params: 
  patient: "P2"
  file: "~/Documents/Singlecell-R/Scripts/Matrix/CLL_Seurat_P2.rds"
  Matrix: "~/Documents/Singlecell-R/Scripts/Matrix/"
  Output: "~/Documents/Singlecell-R/Scripts/Plots/"
format: html
editor: visual
---

## Prepare the data for the boolean model

On the patient 2, on the CLL cells and the intermediate population. The data composition is analysed across all CLL cells and myeloid cells to observe the evolution through the bridge. But then the macrostates definition is only based on CLL cells and the bridge.

1- Explore the population composition in cell type, timepoint to define macrostates

2- Define macroostates based on the Stream2 trajectory (in jupyter notebook), the cell type and the timepoint

```{r}
#| message: false
library(Seurat)
library(ggplot2)
library(SingleR)
library(celldex)
library(dplyr)
library(tidyr)
library(dorothea)
library(readr)
library(stringr)
library(VennDiagram)
library(grid)
library(ggpubr)
library(pROC)
library(patchwork)
source("functions.R")
```

```{r}
data_patient<-readRDS(params$file)
```

## Data

CLL cells + bridge

```{r}
data_cll<-subset(data_patient, subset = (Annotation %in% c("B intermediate", "B naive", "B memory", "B", "Bridge")))
data_cll<-run_dim_red(data_cll)
```

CLL cells + myeloid cells

```{r}
data<-subset(data_patient, subset = (Annotation %in% c("B intermediate", "B naive", "B memory", "B", "Bridge","Monocyte","Macrophage")))
data<-run_dim_red(data)
data$timepoint <- factor(data$timepoint, levels = c("D1", "D4", "D8", "D11", "D14"))

plot<-DimPlot(data, group.by = "Annotation", reduction="umap")
plot1<-DimPlot(data, group.by = "timepoint", reduction="umap")
p<-plot+plot1
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/umap_cll_myel_annot.jpeg"), p, width = 4500, height = 2000, units = "px")
```

## 1- Exploring the population composition

Precise B cells annotation

```{r}
ref <- celldex::HumanPrimaryCellAtlasData()
norm_counts <- LayerData(data, assay = "RNA", layer = 'data')
ref <- ref[,grepl('B_cell', ref$label.fine, ignore.case = TRUE)]
unique(ref$label.fine)
annotation<-SingleR(test = norm_counts,
                  ref = ref, 
                  labels = ref$label.fine,
                  de.method = 'wilcox')
data <- AddMetaData(data, annotation$pruned.labels, col.name = 'B_fine')

data$B_annotation <- ifelse(
  data$Annotation %in% c("Bridge", "Macrophage", "Monocyte"),
  data$Annotation,
  data$B_fine
)

data$B_annotation <- ifelse(
  data$B_annotation %in% c("B_cell:CXCR4-_centrocyte", "B_cell:CXCR4+_centroblast"),
  "B_cell:Germinal_center",
  data$B_annotation)
data$B_annotation <- ifelse(
  data$B_annotation %in% c("Pro-B_cell_CD34+", "Pre-B_cell_CD34-"),
  "B_progenitor",
  data$B_annotation)
plot<-DimPlot(data, group.by = "B_annotation", reduction="umap")
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/umap_cll_myel_annotB.jpeg"), plot, width = 3000, height = 2000, units = "px")
```

```{r}
data$color <- ifelse(
  data$B_annotation %in% c("B_cell:Naive"),
  "B_cell:Naive",
  "Other")
DimPlot(data, group.by = "color",
        cols = c("B_cell:Naive" = "red", "Other" = "gray"),
        pt.size = 0.8,
        label = FALSE, reduction="umap")
```

Clustering to separate the different CLL population and define trajectory

```{r}
data<-FindClusters(data, resolution = 0.1)
DimPlot(data, reduction="umap")
```

Proportion in timepoint and cell type (B naive and B intermediate)

```{r}
df <- data@meta.data
png("~/Documents/Singlecell-R/Scripts/Plots/barplot_time_clus.png", width = 900, height = 500)
df %>%
  filter(seurat_clusters %in% c(0, 1, 2),
         Annotation %in% c("B intermediate", "B naive", "B memory","B")) %>%
  group_by(seurat_clusters, timepoint, Annotation) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters, timepoint) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = timepoint, y = prop, fill = Annotation)) +
  geom_bar(stat = "identity", position = "fill") +
  facet_wrap(~ seurat_clusters) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(
    "B intermediate" = "#1f78b4",
    "B naive" = "#33a02c",
    "B memory" = "darkblue",
    "B"="darkgreen"
  )) +
  labs(y = "Cell proportion", fill = "Cell type") +
  theme_classic()+theme(axis.title = element_text(size = 18),axis.text = element_text(size = 18),legend.title = element_text(size = 18),legend.text = element_text(size = 18),strip.text = element_text(size = 18)
  )
dev.off()
```

```{r}
DimPlot(data, reduction="umap", group.by = "timepoint", cols =c("D1"="red", "D4"="grey", "D8"="grey", "D11"="grey", "D14"="darkblue"), pt.size = 0.1)
```

```{r}
df <- data@meta.data
umap_coords <- as.data.frame(data@reductions[["umap"]]@cell.embeddings)
umap_coords$cell <- rownames(umap_coords)
df$cell <- rownames(df)
df2 <- df %>%
  left_join(umap_coords, by = "cell")
rownames(df2)<-df2$cell

df2 <- df2 %>%
  filter(seurat_clusters %in% c(0, 1, 2)) %>%
  mutate(cluster_side = case_when(
    seurat_clusters == 0 & umap_1 <= 0 ~ "0_left",
    seurat_clusters == 0 & umap_1 >  0 ~ "0_right",
    seurat_clusters == 1 & umap_1 <= 0 ~ "1_left",
    seurat_clusters == 1 & umap_1 >  0 ~ "1_right",
    seurat_clusters == 2 ~ "2",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(cluster_side))

png("~/Documents/Singlecell-R/Scripts/Plots/prop_time_clus.png", width = 900, height = 500)
df2 %>%
  dplyr::count(cluster_side, timepoint) %>%
  group_by(cluster_side) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = timepoint, y = factor(cluster_side, levels = c("0_left", "0_right", "1_left", "1_right", "2")), fill = prop)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(prop, 2)), size = 6) +
  scale_fill_gradient(low = "white", high = "red", limits = c(0, 1)) +
  labs(x = "Timepoint", y = "Clusters", fill = "Proportion") +
  theme_classic()+theme(axis.title = element_text(size = 18),axis.text = element_text(size = 18),legend.title = element_text(size = 18),legend.text = element_text(size = 14),strip.text = element_text(size = 18))
dev.off()

png("~/Documents/Singlecell-R/Scripts/Plots/prop_time_clus_bis.png", width = 900, height = 500)
df2 %>%
  dplyr::count(cluster_side, timepoint) %>%
  group_by(timepoint) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = timepoint, y = factor(cluster_side, levels = c("0_left", "0_right", "1_left", "1_right", "2")), fill = prop)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(prop, 2)), size = 6) +
  scale_fill_gradient(low = "white", high = "red", limits = c(0, 1)) +
  labs(x = "Timepoint", y = "Clusters", fill = "Proportion") +
  theme_classic()+theme(axis.title = element_text(size = 18),axis.text = element_text(size = 18),legend.title = element_text(size = 18),legend.text = element_text(size = 14),strip.text = element_text(size = 18))
dev.off()

df2 %>%
  group_by(cluster_side, timepoint, Annotation) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(cluster_side, timepoint) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = timepoint, y = prop, fill = Annotation)) +
  geom_bar(stat = "identity", position = "fill") +
  facet_wrap(~ factor(cluster_side, levels = c("0_left", "0_right", "1_left", "1_right", "2"))) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(
    "B intermediate" = "#1f78b4",
    "B naive" = "#33a02c", 
    "B memory" = "darkblue",
    "B" = "darkgreen"
  )) +
  labs(y = "Proportion des cellules", fill = "Type B") +
  theme_classic() +
  theme(
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 18),
    legend.title = element_text(size = 18),
    legend.text = element_text(size = 18),
    strip.text = element_text(size = 18)
  )
```

```{r}
df_long <- data@meta.data %>%
  mutate(type = "Timepoint", value = timepoint) %>%
  bind_rows(
    data@meta.data %>%
      mutate(type = "Celltype", value = Annotation)
  )

meta_sub <- df_long %>%
  dplyr::filter(seurat_clusters %in% c("0", "1", "2"))

ggplot(meta_sub, aes(x = seurat_clusters, fill = value)) +
  geom_bar(position = "fill") +
  facet_wrap(~type, scales = "free") +
  ylab("Proportion") +
  xlab("Cluster") +
  theme_minimal()
```

```{r}
meta_sub <- data@meta.data %>%
  filter(seurat_clusters %in% c(0,1,2),
         Annotation %in% c( "B naive", "B intermediate"))

df_counts <- meta_sub %>%
  group_by(seurat_clusters, timepoint, Annotation) %>%
  summarise(n = n(), .groups = "drop")

df_prop <- df_counts %>%
  group_by(seurat_clusters, timepoint) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

df_wide <- df_prop %>%
  pivot_wider(names_from = Annotation, values_from = prop, values_fill = 0)

png("~/Documents/Singlecell-R/Scripts/Plots/prop_time_celltype_clus.png", width = 900, height = 500)
ggplot(df_prop, aes(x = seurat_clusters, y = timepoint, fill = prop)) +
  geom_tile(color = "white") +
  facet_wrap(~Annotation) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  geom_text(aes(label = round(prop, 2)), size = 6) +
  theme_minimal() +
  xlab("Cluster") +
  ylab("Timepoint") +theme_classic()+theme(axis.title = element_text(size = 18),axis.text = element_text(size = 18),legend.title = element_text(size = 18),legend.text = element_text(size = 14),strip.text = element_text(size = 18))
  ggtitle("Celltypes B proportion per clusters and timepoints")
dev.off()
```

## 2- Macrostates definition

Based on : Initial states are D1 and intermediate B cells in the different CLL clusters (clusters 0 and 1). Terminal states are D14 and naive B cells for clusters 1, D14 and B intermediate in cluster 2, and all the bridge (no timepoint selection)

Constraints/Trajectory : I1 –\> T1, I1 –\> T3, I2 –\>T2, I2-\>T3

```{r}
data$macrostates <- ifelse(
  data$seurat_clusters == "0" & data$timepoint == "D1" & data$Annotation == "B intermediate",
  "I1",
  ifelse(
    data$Annotation == "Bridge",
    "T1", 
    ifelse(
      data$seurat_clusters == "2" & data$timepoint == "D14" & data$Annotation == "B intermediate",
      "T3",
      ifelse(
        data$seurat_clusters == "1" & data$timepoint == "D1" & data$Annotation == "B intermediate",
        "I2",
        ifelse(
        data$seurat_clusters == "1" & data$timepoint == "D14" & data$Annotation == "B naive",
        "T2", "None")))))

Idents(data) <- data$macrostates
cells_red  <- WhichCells(data, idents = c("I1","I2"))
cells_blue <- WhichCells(data, idents = c("T1","T2","T3"))
cells_grey <- setdiff(colnames(data), c(cells_red, cells_blue))
plot<-DimPlot(data,reduction = "umap",cells.highlight = list(Initial = cells_red, Terminal = cells_blue),cols.highlight = c("Initial" = "darkblue", "Terminal" = "red"), sizes.highlight = 0.1, pt.size = 0.01) + ggtitle("Macrostates based on the timepoint, cluster and cell type")
ggsave("~/Documents/Singlecell-R/Scripts/Plots/macrostate.jpeg", plot, width = 3000, height = 2000, units = "px")
```

```{r}
data_cll<-subset(data, subset = (Annotation %in% c("B intermediate", "B naive", "B memory", "B", "Bridge")))
data_cll<-run_dim_red(data_cll)
Convert_Seurat_Anndata(data_cll, "~/Documents/Singlecell-R/Scripts/Matrix/cll_raw_macro", raw=TRUE)
```

## 3 - Influence graph for BoNesis: DorothEA and infered GRN

```{r}
# DoRothEA 
data(dorothea_hs, package = "dorothea")

dorothea_base <- dorothea_hs %>%
  filter(confidence %in% c("A", "B","C")) %>%
  transmute(
    source = tf,
    target,
    sign = ifelse(mor == 1, "+1", "-1"),
    origin = "dorothea")

# GRN SCENIC
# Parameters
NES_MIN          <- 3.0   # strength of the argument
QVALUE_MAX       <- 0.05  # reliability of annotation (0 = directly annotated)
ORTHO_MIN        <- 0.8   # minimal orthological identity
TARGET_SCORE_MIN <- 0.5   # minimum target size

scenic_grn <- read.csv(paste0(params$Matrix,"step_2_reg_p2.csv"), skip = 3) 
colnames(scenic_grn) <- c("TF", "MotifID", "AUC", "NES", "MotifQvalue","OrthologousIdentity", "Annotation", "Context","TargetGenes", "RankAtMax")

scenic_filt <- scenic_grn %>%
  filter(
    NES                 >= NES_MIN,
    MotifQvalue         <= QVALUE_MAX,
    OrthologousIdentity >= ORTHO_MIN)
cat("Regulons before filtering :", nrow(scenic_grn), "\n")
cat("Regulons after filtering  :", nrow(scenic_filt), "\n")

parse_targets_with_scores <- function(tg_string) {
  matches <- str_match_all(tg_string, "'([A-Z0-9\\-\\.]+)',\\s*([0-9\\.]+)")[[1]]
  if (nrow(matches) == 0) return(tibble(target = character(), score = numeric()))
  tibble(
    target = matches[, 2],
    score  = as.numeric(matches[, 3])
  )
}

scenic_edges <- scenic_filt %>%
  dplyr::select(TF, TargetGenes) %>%
  rowwise() %>%
  mutate(targets = list(parse_targets_with_scores(TargetGenes))) %>%
  unnest(targets) %>%
  dplyr::select(source = TF, target, score) %>%
  group_by(source, target) %>%
  summarise(score = max(score), .groups = "drop") %>%
  filter(score >= TARGET_SCORE_MIN)

cat("Edges after targets filtering:", nrow(scenic_edges), "\n")

# Keep edges SCENIC missing in DoRothEA
scenic_new <- scenic_edges %>%
  anti_join(dorothea_base, by = c("source", "target"))

# Sign "+" for SCENIC
scenic <- scenic_new %>%
  mutate(sign = "+1", origin = "scenic_default")

# Combine DoRothEA + SCENIC
influence_graph <- bind_rows(
  dorothea_base,
  scenic
) %>%
  dplyr::select(source, target, sign)

# Recap
cat("Edges DoRothEA     :", nrow(dorothea_base), "\n")
cat("Edges SCENIC       :", nrow(scenic), "\n")
cat("Total              :", nrow(influence_graph), "\n")

write_csv(influence_graph, paste0(params$Matrix,"influence_graph_bonesis.csv"))
```

```{r}
# Nodes
nodes_scenic   <- unique(c(scenic_edges$source, scenic_edges$target))
nodes_dorothea <- unique(c(dorothea_base$source, dorothea_base$target))
tf_scenic <- unique(scenic_edges$source)
dorothea_filtered <- dorothea_base[dorothea_base$source %in% tf_scenic, ]

# Edges
edges_scenic   <- paste(scenic_edges$source, scenic_edges$target,   sep = "__")
edges_dorothea <- paste(dorothea_filtered$source, dorothea_filtered$target, sep = "__")

venn_nodes <- venn.diagram(
  x = list(SCENIC = nodes_scenic, DoRothEA = nodes_dorothea),
  filename = NULL,
  category.names = c("SCENIC", "DoRothEA"),
  fill = c("#4E79A7", "#F28E2B"),
  alpha = 0.5,
  cex = 2,
  cat.cex = 1.8,
  main = "Commons nodes (TF + Target)",print.mode = c("raw", "percent"),
  main.cex = 1.5
)

plot <- venn.diagram(
  x = list(SCENIC = edges_scenic, DoRothEA = edges_dorothea),
  filename = NULL,
  category.names = c("SCENIC", "DoRothEA"),
  fill = c("#76B7B2", "#E15759"),
  alpha = 0.5,
  cex = 2,
  cat.cex = 1.8,cat.pos = c(180, 180),
  main = "Commons edges",print.mode = c("raw", "percent"),
  main.cex = 1.5
)
ggsave(paste0(params$Output,"venn_edges.jpeg"), plot, width = 3000, height = 2000, units = "px")
```

```{r}
dorothea_net <- get_dorothea(organism = "human", levels = c("A", "B", "C"))
scenic <- read.csv(paste0(params$Matrix,"step_2_reg_p2.csv"), skip = 3) 
colnames(scenic) <- c("TF", "MotifID", "AUC", "NES", "MotifQvalue","OrthologousIdentity", "Annotation", "Context","TargetGenes", "RankAtMax")

scenic <- scenic %>%
   dplyr::select(TF, TargetGenes, NES) %>%
   rowwise() %>%
   mutate(targets = list(parse_targets_with_scores(TargetGenes))) %>%
   unnest(targets) %>%
   dplyr::select(TF, target, score, NES) %>%
   group_by(TF, target) %>%
   summarise(
      score = max(score),
      NES = max(NES),
      .groups = "drop")

scenic <- scenic %>%
  mutate(in_dorothea = paste(TF, target) %in% paste(dorothea_base$source, dorothea_base$target),
         status = ifelse(in_dorothea, "Known (DoroTHEA)", "New (SCENIC)"))
```

```{r}
# Wilcoxon 
wilcox_NES <- wilcox.test(NES ~ status, data = scenic)
cat("Wilcoxon NES p-value:", wilcox_NES$p.value, "\n")

# Effect size (Cohen's d ou rank-biserial)
library(rstatix)
scenic %>% wilcox_effsize(NES ~ status)

wilcox_aff <- wilcox.test(affiliation_score ~ status, data = scenic)
cat("Wilcoxon affiliation p-value:", wilcox_aff$p.value, "\n")

scenic %>% wilcox_effsize(score ~ status)
```

```{r}
p1<- ggplot(scenic, aes(x = status, y = NES, fill = status)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white") +
  stat_compare_means(method = "wilcox.test", label = "p.signif", 
                     comparisons = list(c("Known (DoroTHEA)", "New (SCENIC)"))) +
  scale_fill_manual(values = c("Known (DoroTHEA)" = "#2196F3", 
                                "New (SCENIC)" = "#FF9800")) +
  labs(title = "NES",
       x = "", y = "NES") +
  theme_bw() + theme(legend.position = "none")

p2<- ggplot(scenic, aes(x = status, y = score, fill = status)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white") +
  stat_compare_means(method = "wilcox.test", label = "p.signif",
                     comparisons = list(c("Known (DoroTHEA)", "New (SCENIC)"))) +
  scale_fill_manual(values = c("Known (DoroTHEA)" = "#2196F3", 
                                "New (SCENIC)" = "#FF9800")) +
  labs(title = "Affiliation score",
       x = "", y = "Affiliation score") +
  theme_bw() + theme(legend.position = "none")

p3<- ggplot(scenic, aes(x = NES, y = score, color = status)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_density_2d(aes(group = status), linewidth = 0.4) +
  scale_color_manual(values = c("Known (DoroTHEA)" = "#2196F3", 
                                 "New (SCENIC)" = "#FF9800")) +
  labs(title = "NES vs Affiliation score",
       x = "NES", y = "Affiliation score", color = "") +
  theme_bw() + theme(legend.position = "bottom")


roc_NES <- roc(scenic$in_dorothea, scenic$NES, quiet = TRUE)
roc_aff <- roc(scenic$in_dorothea, scenic$score, quiet = TRUE)

cat("AUC NES:", auc(roc_NES), "\n")
cat("AUC affiliation:", auc(roc_aff), "\n")

# Plot ROC
p4<- ggroc(list(NES = roc_NES, Affiliation = roc_aff)) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "grey") +
  scale_color_manual(values = c("NES" = "#2196F3", "Affiliation" = "#FF9800")) +
  labs(title = "ROC : DoroTHEA interaction prediction",
       subtitle = paste0("AUC NES = ", round(auc(roc_NES), 3),
                         " | AUC Affiliation = ", round(auc(roc_aff), 3)),
       color = "") +
  theme_bw()

plot<-(p1 | p2) / (p3 | p4)+
  plot_annotation(title = "Validation SCENIC, comparison with DoroTHEA",
                  theme = theme(plot.title = element_text(size = 16, face = "bold")))
ggsave(paste0(params$Output,"score_scenic.jpeg"), plot, width = 3000, height = 2000, units = "px")
```
