---
title: "GRN CLL dataset"
params: 
  data: "~/Documents/Singlecell-R/Scripts/Data/Dataset CLL-map (Knisbacher, B. A., et al. 2022)/cllmap_rnaseq_tpms_603samps_deseqLog10_batchcorr.tsv"
  data_ln: "~/Documents/Singlecell-R/Scripts/Data/Dataset CLL lymph node (Sun C, et al., 2023)/GSE161711_log2CPM_counts.txt"
  clinical_ln: "~/Documents/Singlecell-R/Scripts/Data/Dataset CLL lymph node (Sun C, et al., 2023)/SraRunTable.csv"
  sample_ln: "~/Documents/Singlecell-R/Scripts/Data/Dataset CLL lymph node (Sun C, et al., 2023)/Sample_name.txt"
  network: "/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/network_cll.txt"
format: html
editor: visual
---

## GRN CLL RNA seq bulk

### **Lymph node sample**

Dataset taken from the publication: Sun C, et al., 2023 doi:10.1182/bloodadvances.

\(1\) bulk RNA-seq (n = 29); (2) WES (n = 14) of CD19+ tumor cells from paired PB and LN samples; (3) scRNA-seq of paired PB and LN samples (n = 5); (4) bulk RNA-seq of unsorted CLL LN samples (n = 34); and (5) bulk RNA-seq of unsorted normal LN samples (n = 4).

```{r}
# Lymph node CLL samples
data_ln<-read.table(params$data_ln,header=TRUE)
clinical_ln<-read.csv(params$clinical_ln,header=TRUE)
sample_ln<-read.table(params$sample_ln,header=F)
colnames(sample_ln)<-c("GEO_Accession..exp.","Sample_names")
clinical_ln<-merge(clinical_ln, sample_ln, by="GEO_Accession..exp.")
#clinical_ln<-subset(clinical_ln, tissue == "lymph node" & diagnosis== "CLL")
clinical_ln <- clinical_ln %>%
  filter(grepl("LN_mix", Sample_names))
clinical_ln<-subset(clinical_ln, tissue == "lymph node" & diagnosis != "Normal")

# Filter some gene type 
data_ln<-subset(data_ln, gene_type =="protein_coding")

# Keep CLL and Lymph node mix sample  (n=34)
data_ln <- data_ln[, c(-1,-2,-4,-5,-6,-7,-8,-9)]
data_ln$variance <- apply(data_ln[, -1], 1, var)
# keep gene with the higher variance
data_ln <- data_ln %>%
  group_by(gene_name) %>%
  slice_max(variance, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(-variance)
data_ln <- as.matrix(data_ln)
data_ln <- data_ln[, grepl("LN_mix", colnames(data_ln)) | colnames(data_ln) == "gene_name"]
data_ln <- data_ln[, grepl("CLL", colnames(data_ln)) | colnames(data_ln) == "gene_name"]
```

Gene name issus (MARCH and SEP)

```{r}
fix_gene_names <- function(x) {
  x <- as.character(x)
  for (i in 1:20) {
    x <- sub(paste0("^", i, "-Mar$"), paste0("MARCH", i), x, ignore.case = TRUE)
    x <- sub(paste0("^", i, "-Sep$"), paste0("SEPT", i), x, ignore.case = TRUE)
  }
  x
}
data_ln[, "gene_name"] <- fix_gene_names(data_ln[, "gene_name"])
```

```{r}
data_ln <- as.matrix(data_ln)
write.table(data_ln, file = "~/Documents/Singlecell-R/Scripts/Matrix/matrix_lymph_node.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)
```

### **PBMC sample**

Dataset from : Knisbacher, B. A., et al. 2022,<https://doi.org/10.1038/s41588-022-01140-w>

603 samples of PBMC traitement naive

```{r}
library(dplyr)
data<-read.table(params$data, sep="\t", header=TRUE)
clinical<-read.table(params$clinical, sep="\t", header=TRUE, fill = TRUE)
sample<-read.table(params$sample, sep="\t", header=TRUE)
```

```{r}
# Change the header and rownames format 
data<-read.table(params$data, sep="\t", header=TRUE)
data<-data[,-1]
data <- data %>%
  dplyr::rename(gene = Description)
data$variance <- apply(data[, -1], 1, var)
# keep gene with the higher variance
data <- data %>%
  group_by(gene) %>%
  slice_max(variance, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(-variance)
genes <- data$gene
data$gene <- NULL
rownames(data) <- genes
data <- as.matrix(data)
write.table(data, file = "~/Documents/Singlecell-R/Scripts/Matrix/matrix.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)
```

```{bash}
# ARACNe pipeline 
for i in {1..100}
do
java -Xmx5G -jar /home/ablanc/work/scie-panc/ARACNe-AP/dist/aracne.jar -e matrix.txt -o outputFolder --tfs TF_list.txt --pvalue 1E-8 --calculateThreshold --seed $i
done

java -Xmx5G -jar /home/ablanc/work/scie-panc/ARACNe-AP/dist/aracne.jar -o outputFolder --consolidate
```

```{r}
library(viper)
grn<-read.table("/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/network_cll.txt", header=F)
grn<-grn[-1,-4]
write.table(grn, "/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/network_cll.txt", sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)
regulon <- aracne2regulon("/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/network_cll.txt", data,  format = "3col")

# 
edges <- bind_rows(
  lapply(names(regulon), function(tf) {
    data.frame(
      source = tf,
      target = names(regulon[[tf]]$tfmode),
      sign = ifelse(regulon[[tf]]$tfmode > 0, 1L, -1L),
      tfmode = regulon[[tf]]$tfmode,
      likelihood = regulon[[tf]]$likelihood,
      stringsAsFactors = FALSE)
  }))

regulon_size <- sapply(regulon, function(x) length(x$tfmode))
hist(regulon_size)

likelihoods <- unlist(lapply(regulon, function(x) x$likelihood))
hist(likelihoods, breaks = 100)
summary(likelihoods)

edges<-edges[,c(-4,-5)]

write.csv(edges, "/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/GRN_CLL.csv", row.names = FALSE, quote = FALSE, col.names = TRUE)
```

Visualize the GRN network

```{r}
top_edges <- network[order(network$MI, decreasing = TRUE), ][1:200, ]
g <- graph_from_data_frame(d = top_edges[, c("Regulator", "Target")], directed = TRUE)

all_regulators <- unique(top_edges$Regulator)
all_targets <- unique(top_edges$Target)

both_nodes <- intersect(all_regulators, all_targets)

E(g)$MI <- top_edges$MI

V(g)$type <- case_when(
  V(g)$name %in% both_nodes ~ "Both",
  V(g)$name %in% all_regulators ~ "Regulator",
  V(g)$name %in% all_targets ~ "Target",
  TRUE ~ "Other"
)

plot<-ggraph(g, layout = "fr") +
  geom_edge_link0(aes(edge_width = MI), alpha = 0.6, edge_colour = "grey") +
  geom_node_point(aes(color = type), size = 5, show.legend = TRUE) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_color_manual(values = c("Regulator" = "orange", "Target" = "blue", "Both" = "purple", "Other"= "grey")) +
  scale_edge_width(name = "MI") +
  guides(edge_width = guide_legend(override.aes = list(linetype = 1, shape = NA))) + 
  theme_void() + ggtitle("ARACNe GRN CLL lymph node, orange = regulators, blue = targets and purple= both")
plot
```
