---
title: "GRN CLL dataset"
params: 
  data: "~/Documents/Singlecell-R/Scripts/Data/Dataset CLL-map (Knisbacher, B. A., et al. 2022)/cllmap_rnaseq_tpms_603samps_deseqLog10_batchcorr.tsv"
  data_ln: "~/Documents/Singlecell-R/Scripts/Data/Dataset CLL lymph node (Sun C, et al., 2023)/GSE161711_log2CPM_counts.txt"
format: html
editor: visual
---

## GRN CLL RNA seq bulk

### **Lymph node sample**

Dataset taken from the publication: Sun C, et al., 2023 doi:10.1182/bloodadvances.

Bulk RNA sequencing (RNA-seq) of (1) immunomagnetically purified CD19+ cells from peripheral blood (n = 29), (2) CD19+ cells from lymph node single cell suspensions (n = 29), and (3) lymph node biopsies (n = 38). All samples were matched in 29 patients with CLL. RNA-seq of lymph node biopsies only were done in 5 patients and 4 normal controls

```{r}
data_ln<-read.table(params$data_ln,header=TRUE)
```

### **PBMC sample**

Dataset from : Knisbacher, B. A., et al. 2022,<https://doi.org/10.1038/s41588-022-01140-w>

603 samples of PBMC traitement naive

```{r}
library(dplyr)
data<-read.table(params$data, sep="\t", header=TRUE)
clinical<-read.table(params$clinical, sep="\t", header=TRUE, fill = TRUE)
sample<-read.table(params$sample, sep="\t", header=TRUE)
# Keep Untreated sample 
clinical<- subset(clinical, treatment_status_at_samp=="Untreated" & prior_treatment_category=="Untreated")
clinical$participant_id <- gsub("-", ".", clinical$participant_id)
# Keep ID sample from data 
samples <- unique(clinical$participant_id)
fixed_cols <- c("Description")
sample_cols <- intersect(colnames(data), samples)
data <- data[, c(fixed_cols, sample_cols)]
data <- data %>%
  dplyr::rename(gene = Description)
write.table(data, file = "~/Documents/Singlecell-R/Scripts/Matrix/matrix.txt", sep = "\t", quote = FALSE, col.names = NA)
```

```{r}
data<-read.table(params$data, sep="\t", header=TRUE)
data<-data[,-1]
data <- data %>%
  dplyr::rename(gene = Description)
data$variance <- apply(data[, -1], 1, var)
# keep gene with the higher variance
data <- data %>%
  group_by(gene) %>%
  slice_max(variance, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(-variance)
rownames(data) <-data$gene
write.table(data, file = "~/Documents/Singlecell-R/Scripts/Matrix/matrix.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)
```

```{bash}
# ARACNe pipeline 
for i in {1..100}
do
java -Xmx5G -jar /home/ablanc/work/scie-panc/ARACNe-AP/dist/aracne.jar -e matrix.txt -o outputFolder --tfs TF_list.txt --pvalue 1E-8 --calculateThreshold --seed $i
done

java -Xmx5G -jar /home/ablanc/work/scie-panc/ARACNe-AP/dist/aracne.jar -o outputFolder --consolidate
```
