---
title: "infer_CNA"
params: 
  patient: "P2"
  patientbis: "P3"
  patient_numbat: "p3"
  patient_infercnv: "p2"
  file: "~/Documents/Singlecell-R/Scripts/Matrix/CLL_Seurat_P2.rds"
  filebis: "~/Documents/Singlecell-R/Scripts/Matrix/CLL_Seurat_P3.rds"
  output: "~/Documents/Singlecell-R/Scripts/Plots/"
format: html
editor: visual
---

## Copy Number Alteration calculation

Goal : track the cancer cell thank to the CNA

```{r}
#| message: false
library(Seurat)
library(copykat)
library(ggplot2)
library(stringr)
library(numbat)
library(data.table)
library(dplyr)
library(glue)
library(ggtree)
library(tidygraph)
library(patchwork)
library(ggpubr)
library(infercnv)
library(biomaRt)
library(Matrix)
library(uwot)
source("functions.R")
```

```{r}
data_p2<-readRDS(params$file)
data_p3<-readRDS(params$filebis)
```

## CopyKat

Run copyKat (timepoint separately)

Patient 2

```{r}
timepoints <- c("D1", "D4", "D8", "D11", "D14")
for (tp in timepoints) {

  data_tp <- subset(data_p2, subset = timepoint == tp)
  counts_matrix <- GetAssayData(data_tp, assay = "RNA", layer = "counts")
  norm.cells <- colnames(data_tp)[data_tp$Annotation %in% c("Monocyte", "CD4 T", "CD8 T", "NK")]
  
  ck <- copykat(
    rawmat = counts_matrix,
    id.type = "S",
    ngene.chr = 5,
    win.size = 25,
    KS.cut = 0.1,
    sam.name = tp,
    distance = "euclidean",
    norm.cell.names = norm.cells,
    output.seg = "FALSE",
    plot.genes = "TRUE",
    genome = "hg20",
    n.cores = 1
  )
  
  copykat_results[[tp]] <- ck
  pred_results[[tp]] <- data.frame(ck$prediction)
  cna_results[[tp]] <- data.frame(ck$CNAmat)
}
```

Patient 3

```{r}
timepoints <- c("D1", "D4", "D8", "D11", "D14")
for (tp in timepoints) {

  data_tp <- subset(data_p3, subset = timepoint == tp)
  counts_matrix <- GetAssayData(data_tp, assay = "RNA", layer = "counts")
  norm.cells <- colnames(data_tp)[data_tp$Annotation %in% c("Monocyte", "NK")]
  
  ck <- copykat(
    rawmat = counts_matrix,
    id.type = "S",
    ngene.chr = 5,
    win.size = 25,
    KS.cut = 0.1,
    sam.name = tp,
    distance = "euclidean",
    norm.cell.names = norm.cells,
    output.seg = "FALSE",
    plot.genes = "TRUE",
    genome = "hg20",
    n.cores = 1
  )
  
  copykat_results[[tp]] <- ck
  pred_results[[tp]] <- data.frame(ck$prediction)
  cna_results[[tp]] <- data.frame(ck$CNAmat)
}
```

Exploring the results

Add the results in the Seurat Object, day sep

```{r}
pred_D1 <- data.frame(copykat_D1$prediction)
data$copykat_pred <- NA
rownames(pred.res_D1) <- pred_D1$cell.names
data$copykat_pred <- pred_D1[colnames(data), "copykat.pred"]
```

```{r}
DimPlot(data, group.by = "copykat_pred", reduction = "umap", pt.size=0.1, label.size = 2.5) + ggtitle(str_wrap("UMAP", width = 60)) + theme(plot.title = element_text(size = 15, hjust = 0.5))

```

Timepoints combined

Patient 2

```{r}
copykat.res <- do.call(rbind, pred_results)data_p2$copykat_pred <- NA
rownames(copykat.res) <- copykat.res$cell.names
data_p2$copykat_pred <- copykat.res[colnames(data_p2), "copykat.pred"]

title=paste("UMAP CNA, patient", params$patient)
plot<-DimPlot(data_p2, group.by = "copykat_pred", reduction = "umap", pt.size=0.05, label.size = 2.5) + ggtitle(str_wrap(title, width = 60)) + theme(plot.title = element_text(size = 10, hjust = 0.5))
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/cna_", params$patient, ".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

Patient 3

```{r}
copykat.res <- do.call(rbind, pred_results)data_p3$copykat_pred <- NA
rownames(copykat.res) <- copykat.res$cell.names
data_p3$copykat_pred <- copykat.res[colnames(data_p3), "copykat.pred"]

title=paste("UMAP CNA, patient", params$patientbis)
plot<-DimPlot(data_p3, group.by = "copykat_pred", reduction = "umap", pt.size=0.05, label.size = 2.5) + ggtitle(str_wrap(title, width = 60)) + theme(plot.title = element_text(size = 10, hjust = 0.5))
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/cna_", params$patientbis, ".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

## Numbat

Patient 2

Barcodes files

```{r}
timepoints <- c("D1", "D4", "D8", "D11", "D14")
for (tp in timepoints) {
  data_sub<-subset(data_p2, subset = timepoint == tp)
  counts_matrix <- GetAssayData(data_sub, assay = "RNA", layer = "counts")
  barcodes <- colnames(counts_matrix)
  barcodes <- sub("^D[0-9]+_", "", barcodes)
  write.table(barcodes, paste0("~/Documents/Singlecell-R/Scripts/Data/barcodes_", tp, ".tsv"), sep="\t",row.names = FALSE, col.names = FALSE)
}
```

Run the pile up and phasing on Numbat Docker in the HPC : run the script Numbat.sh

Combining the pileup & phasing out (because one per timepoint) + add the timepoint prefix on the ID cell

```{r}
files <- c(
  "/home/ablanc/work/NetBIO2/CNA/Numbat_p2/Results_pileup_phasing/D1_allele_counts.tsv.gz",
  "/home/ablanc/work/NetBIO2/CNA/Numbat_p2/Results_pileup_phasing/D4_allele_counts.tsv.gz",
  "/home/ablanc/work/NetBIO2/CNA/Numbat_p2/Results_pileup_phasing/D8_allele_counts.tsv.gz",
  "/home/ablanc/work/NetBIO2/CNA/Numbat_p2/Results_pileup_phasing/D11_allele_counts.tsv.gz",
  "/home/ablanc/work/NetBIO2/CNA/Numbat_p2/Results_pileup_phasing/D14_allele_counts.tsv.gz"
)

timepoints <- sub(".*/(D\\d+)_allele_counts.*", "\\1", files)

df_allele_list <- lapply(seq_along(files), function(i) {
  fread(files[i]) %>%
    mutate(
      timepoint = as.integer(sub("D", "", timepoints[i])),
      cell = paste0(timepoints[i], "_", cell)
    )
})

df_allele <- bind_rows(df_allele_list)

head(df_allele)
```

Count matrix

```{r}
if (params$patient_numbat =="p2") {
  data<-data_p2
} else {
  data<-data_p3
}

counts_matrix <- GetAssayData(data, assay = "RNA", layer = "counts")
#colnames(counts_matrix) <- sub("^D[0-9]+_", "", colnames(counts_matrix))
```

Reference created (or choose the given one). The reference is important because it's the comparison for calculate the CNV based on the gene expression. It's better to use cells in our sample as reference as normal cells. Here we used reference cell all the cellltype (except B cells) at D1.

```{r}
celltype<-unique(data$final_annot)
celltype <- celltype[!grepl("^B", celltype)]
data_ref<-subset(data, subset = timepoint=="D1"&(final_annot %in% celltype))
cell_annot <- data.frame(
  cell = colnames(data_ref),
  group = data_ref$final_annot,
  stringsAsFactors = FALSE
)
counts_ref <- GetAssayData(data_ref, assay = "RNA", layer = "counts")

ref_internal <- aggregate_counts(counts_ref, cell_annot)
```

### Running Numbat

In run it on the HPC, I used 8 CPU and 200 G, time : \~15h.

```{r}
df_allele<-read.table("~/Documents/Singlecell-R/Scripts/Matrix/all_allele_counts.tsv", sep=",", header=T)
df_allele <- fread("~/Documents/Singlecell-R/Scripts/Matrix/all_allele_counts.tsv.gz")
```

```{r}
out = run_numbat(
    counts_matrix, # gene x cell integer UMI count matrix 
    ref_internal, # reference expression profile, a gene x cell type normalized expression level matrix
    df_allele, # allele dataframe generated by pileup_and_phase script
    genome = "hg38",
    t = 1e-5,
    ncores = 8,
    ncores_nni = 8,
    skip_nj = TRUE,
    plot = TRUE,
    max_entropy = 0.7,
    out_dir = '/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/Resultats_numbat/Results_numbat_ref_internal_p2'
)
```

### Interpreting Numbat results

Put the Numbat folder result and then use the parameter \$patient_numbat in the script header.

```{r}
nb = Numbat$new(out_dir = paste0("/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/Resultats_numbat/Results_numbat_ref_internal_",params$patient_numbat))
```

```{r}
mypal = c('1' = 'gray', '2' = "#377EB8", '3' = "#4DAF4A", '4' = "#984EA3", '5' ="#ea4258", '6'="orange")
  
nb$plot_phylo_heatmap(
  clone_bar = TRUE, 
  p_min = 0.9,
  pal_clone = mypal
)
```

```{r}
nb$bulk_clones %>% 
  filter(n_cells > 50) %>%
  plot_bulks(
    min_LLR = 5, # filtering CNVs by evidence
    legend = TRUE
  )
```

```{r}
head(nb$joint_post) %>% select(cell, CHROM, seg, cnv_state, p_cnv, p_cnv_x, p_cnv_y)
```

```{r}
# Add metadata to the seurat object 
meta_subset <- nb$clone_post[, c("cell", "clone_opt", "GT_opt", "p_opt", "p_cnv", "compartment_opt","p_cnv_x","p_cnv_y")]
rownames(meta_subset) <- meta_subset$cell
meta_subset$cell <- NULL
head(colnames(data))
head(rownames(meta_subset))
data <- AddMetaData(data, metadata = meta_subset)
length(intersect(colnames(data), rownames(meta_subset)))
```

```{r}
plot<-DimPlot(data, group.by = "clone_opt", reduction = "umap", pt.size=0.1, label.size = 2.5, cols =c("gray","#377EB8","#4DAF4A", "#984EA3", "#ea4258","orange") ) + ggtitle(str_wrap("Clonal assignments", width = 60)) + theme(plot.title = element_text(size = 15, hjust = 0.5))
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/umap_clone_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")

plot<-DimPlot(data, group.by = "compartment_opt", reduction = "umap", pt.size=0.1, label.size = 2.5, cols =c("lightblue","#ea4258") ) + ggtitle(str_wrap("Tumor and normal cells annotation", width = 60)) + theme(plot.title = element_text(size = 15, hjust = 0.5))
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/umap_tumor_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")

plot<-FeaturePlot(data, features="p_cnv", reduction="umap")+ ggtitle(str_wrap("Tumor versus normal probability", width = 60)) + theme(plot.title = element_text(size = 15, hjust = 0.5))
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/umap_proba_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

```{r}
df <- data@meta.data
tumor_prop <- df %>%
  group_by(Annotation, compartment_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Annotation) %>%
  mutate(prop = n / sum(n))
totals <- df %>%
  count(Annotation)

p1<-ggplot(tumor_prop, aes(x = Annotation, y = prop, fill = compartment_opt)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(data = totals,
            aes(x = Annotation, y = 1.05, label = paste0("n=", n)),
            inherit.aes = FALSE,
            size = 3) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1.1)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") +
  xlab("Cell Type") +
  labs(fill = "Cell annotation\n(tumor/normal)")

p2 <- ggplot(counts, aes(x = Annotation, y = n)) +
  geom_bar(stat = "identity", fill = "grey50") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Number of cells") +
  xlab("")
plot<-p1 / p2
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/propor_tumor_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

```{r}
df$clone_opt <- as.factor(df$clone_opt)
clone_prop <- df %>%
  group_by(Annotation, clone_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Annotation) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(clone_prop, aes(x = Annotation, y = prop, fill = clone_opt)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Cell Type") + labs(fill = "Cell clone")
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/propor_clone_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

```{r}
GT_prop <- df %>%
     group_by(Annotation, GT_opt) %>%
     summarise(n = n(), .groups = "drop") %>%
     group_by(Annotation) %>%
     mutate(prop = n / sum(n))
plot<-ggplot(GT_prop, aes(x = Annotation, y = prop, fill = GT_opt)) +
     geom_bar(stat = "identity", position = "fill") +
     scale_y_continuous(labels = scales::percent) +
     theme_classic() +
     theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
     ylab("Cell proportion") + xlab("Cell Type") + labs(fill = "Alteration type\n(chromosomes)")
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/propor_mut_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")

df$timepoint <- factor(
  df$timepoint,
  levels = c("D1", "D4", "D8", "D11", "D14")
)

time_GT_prop <- df %>%
     group_by(timepoint, GT_opt) %>%
     summarise(n = n(), .groups = "drop") %>%
     group_by(timepoint) %>%
     mutate(prop = n / sum(n))

ggplot(time_GT_prop, aes(x = timepoint, y = prop, fill = GT_opt)) +
     geom_bar(stat = "identity", position = "fill") +
     scale_y_continuous(labels = scales::percent) +
     theme_classic() +
     theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
     ylab("Cell proportion") + xlab("Cell Type") + labs(fill = "Alteration type\n(chromosomes)")

data_p2$GT_11c <- ifelse(data_p2$GT_opt %in% c("7b,11c","12c_bdel", "7b,11c,12c_del"), "11c", "other")
DimPlot(data, group.by = "GT_11c", reduction = "umap", pt.size=0.1, label.size = 2.5, cols =c("#ea4258","lightblue") ) + ggtitle(str_wrap("Tumor and normal cells annotation", width = 60)) + theme(plot.title = element_text(size = 15, hjust = 0.5))
```

Timepoint by clones

```{r}
clone_prop <- df %>%
     group_by(timepoint, clone_opt) %>%
     summarise(n = n(), .groups = "drop") %>%
     group_by(timepoint) %>%
     mutate(prop = n / sum(n))
clone_prop$timepoint <- factor(clone_prop$timepoint,levels = c("D1","D4","D8","D11","D14"))

plot<-ggplot(clone_prop, aes(x = timepoint, y = prop, color = factor(clone_opt), group = clone_opt)) +
     geom_line(size = 1.2) +
     geom_point(size = 3) +
     scale_y_continuous(labels = scales::percent) +
     theme_classic() +
     labs(x = "Timepoint",y = "Cell proportion",color = "Clone")
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/propor_time_clone_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

Tumor phylogeny

```{r}
nb$plot_sc_tree(
  label_size = 3, 
  branch_width = 0.5, 
  tip_length = 0.5,
  pal_clone = mypal, 
  tip = TRUE
)

nb$plot_mut_history(pal = mypal)
```

Look at the T cells annotated as tumor cells

```{r}
data_T<-subset(data, Annotation %in% c("CD8 T","CD4 T"))
```

```{r}
VlnPlot(data_T, features = "CD19.1", assay="ADT")
VlnPlot(subset(data_p2, Annotation %in% c("CD8 T","CD4 T", "T") & compartment_opt=="normal"), features = "CD19.1", assay="ADT")
```

```{r}
data_T$Tcells_tumor <- paste(data_T$Annotation, data_T$compartment_opt, sep = "_")
adt=c("CD19.1", "CD5.1","CD20","CD47.1","CD4.1","CD8","CD56","CD279 (PD-1)")
plot <- DoHeatmap(data_T, features = adt, size = 4, assay = "ADT", group.by = "Tcells_tumor") +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16)
  )
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/heatmap_Tadt_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

Explore the tumor T cells (B cell genes CD79A, CD79B and protein makers CD19 and CD20)

```{r}
comparisons <- list(c("CD4 T normal","CD4 T tumor"),c("CD8 T normal","CD8 T tumor"))
df <- FetchData(data, vars = c("CD79A", "Annotation", "compartment_opt"))
df <- df %>% filter(Annotation %in% c("CD4 T", "CD8 T"))
df$group <- paste(df$Annotation, df$compartment_opt, sep = " ")
plot <- ggplot(df, aes(x = group, y = CD79A, fill = compartment_opt)) +geom_violin(trim = FALSE, alpha = 0.6) +stat_compare_means(comparisons = comparisons, method = "wilcox.test") +theme_classic() +theme(legend.position = "none",axis.text.x = element_text(angle = 45, hjust = 1, size=12),axis.text.y = element_text(size = 12),axis.title.x = element_text(size = 12),axis.title.y = element_text(size = 12)) +xlab("Cell type") +ylab("CD79A expression")
plot

df <- FetchData(data, vars = c("CD79B", "Annotation", "compartment_opt"))
df <- df %>% filter(Annotation %in% c("CD4 T", "CD8 T"))
df$group <- paste(df$Annotation, df$compartment_opt, sep = " ")
plot1 <- ggplot(df, aes(x = group, y = CD79B, fill = compartment_opt)) +geom_violin(trim = FALSE, alpha = 0.6) +stat_compare_means(comparisons = comparisons, method = "wilcox.test") +theme_classic() +theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12),axis.text.y = element_text(size = 12),axis.title.x = element_text(size = 12),axis.title.y = element_text(size = 12)) +xlab("Cell type") +ylab("CD79B expression")
plot1
p<-plot+plot1
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/vlnplot_cd79_",params$patient_numbat,".jpeg"), p, width = 3000, height = 1500, units = "px")

df <- FetchData(data, vars = c("adt_CD19.1", "Annotation", "compartment_opt"), assay="ADT")
df <- df %>% filter(Annotation %in% c("CD4 T", "CD8 T"))
df$group <- paste(df$Annotation, df$compartment_opt, sep = " ")
plot <- ggplot(df, aes(x = group, y = adt_CD19.1, fill = compartment_opt)) +geom_violin(trim = FALSE, alpha = 0.6) +stat_compare_means(comparisons = comparisons, method = "wilcox.test") +theme_classic() +theme(legend.position = "none",axis.text.x = element_text(angle = 45, hjust = 1, size=12),axis.text.y = element_text(size = 12),axis.title.x = element_text(size = 12),axis.title.y = element_text(size = 12)) +xlab("Cell type") +ylab("CD19 protein expression")
plot

df <- FetchData(data, vars = c("adt_CD20", "Annotation", "compartment_opt"), assay="ADT")
df <- df %>% filter(Annotation %in% c("CD4 T", "CD8 T"))
df$group <- paste(df$Annotation, df$compartment_opt, sep = " ")
plot1 <- ggplot(df, aes(x = group, y = adt_CD20, fill = compartment_opt)) +geom_violin(trim = FALSE, alpha = 0.6) +stat_compare_means(comparisons = comparisons, method = "wilcox.test") +theme_classic() +theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12),axis.text.y = element_text(size = 12),axis.title.x = element_text(size = 12),axis.title.y = element_text(size = 12)) +xlab("Cell type") +ylab("CD20 protein expression")
plot1
p<-plot+plot1
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/vlnplot_pro_",params$patient_numbat,".jpeg"), p, width = 3000, height = 1500, units = "px")
```

```{r}
common_cells <- intersect(colnames(data), colnames(data_p2))
annot_small <- data@meta.data[common_cells, "compartment_opt"]
annot_full  <- data_p2@meta.data[common_cells, "compartment_opt"]
data$annot_full <- NA
data$annot_full[common_cells] <- 
+     data@meta.data[common_cells, "compartment_opt"]
head(data@meta.data[, c("compartment_opt", "annot_full")])
```

```{r}
cna_complete <- data_tot$compartment_opt
cna_myeloid <- data_sub$compartment_opt
common_cells <- intersect(colnames(data_tot), colnames(data_sub))
cna_entire.pop <- cna_complete[common_cells]
cna_myeloid.pop <- cna_myeloid[common_cells]
both_tumor <- sum(cna_entire.pop == "tumor" & cna_myeloid.pop == "tumor")
table( cna_myeloid.pop,cna_entire.pop)
```

## inferCNV

Object preparation

```{r}
data <- get(paste0("data_", params$patient_infercnv))
#data<- subset(data, subset = Annotation %in% c("B","B intermediate","B memory","B naive","Bridge","CD4 T","CD8 T","Erythroblast","Macrophage","Monocyte","Neutrophil","NK"))
#data<- subset(data, subset = Annotation %in% c("B","B intermediate","B memory","B naive","Bridge"))
# raw cout
counts_matrix <- GetAssayData(data, assay = "RNA", layer = "counts")
# cell annotation
#annotations <- data.frame(cell = colnames(data), group = "tumor")
meta <- data@meta.data %>%
  dplyr::select(Annotation)
meta$cellid <- rownames(meta)
meta <- meta[, ncol(meta):1]
meta <- as.data.frame(meta)
meta$Type <- ifelse(meta$Annotation %in% c("CD4 T","CD8 T","Monocyte","Neutrophil","NK"), "Normal", "Tumor")
meta <- meta[,-2]
write.table(meta,"cellAnnotations.txt",sep = "\t",quote = FALSE,col.names = FALSE, row.names = FALSE)

# gene order 
mart <- useEnsembl(
  biomart = "genes",
  dataset = "hsapiens_gene_ensembl")

gene_order <- getBM(
  attributes = c(
    "hgnc_symbol",
    "chromosome_name",
    "start_position",
    "end_position"),mart = mart)
gene_order <- gene_order[gene_order$chromosome_name %in% c(as.character(1:22), "X", "Y"),]
gene_order$chromosome_name <- paste0("chr",gene_order$chromosome_name)
gene_order <- gene_order[gene_order$hgnc_symbol != "" & !is.na(gene_order$hgnc_symbol), ]
gene_order <- gene_order[!duplicated(gene_order$hgnc_symbol), ]
gene_order <- as.data.frame(gene_order)
write.table(gene_order,"gene_ordering_file.txt",sep = "\t",quote = FALSE,col.names = FALSE,row.names = FALSE)

# create the infercnv object
infercnv_obj = CreateInfercnvObject(raw_counts_matrix=counts_matrix,
                                    annotations_file="cellAnnotations.txt",
                                    delim="\t",
                                    gene_order_file="gene_ordering_file.txt",
                                    ref_group_names=c("Normal"))  
# ref_group_names=NULL : no reference cell as normal, take the average of all the celltype
```

Run infercnv

```{r}
# perform infercnv operations to reveal cnv signal
infercnv_obj = infercnv::run(infercnv_obj,
                             cutoff=0.1,  # use 1 for smart-seq, 0.1 for 10x-genomics
                             out_dir=paste0(params$output,"infercnv_res"),
                             cluster_by_groups=T,   # cluster
                             denoise=T,
                             HMM=T)
```

```{r}
# Charger les prédictions par région
cnv_regions <- read.table(paste0(params$output,"infercnv_res/HMM_CNV_predictions.HMMi6.leiden.hmm_mode-subclusters.Pnorm_0.5.pred_cnv_regions.dat"),
  header=T, sep="\t")

# Charger les probabilités d'être normal (Bayesian Network)
# C'est le fichier le plus fiable pour tumeur vs normal
cnv_state_probs <- read.table(paste0(params$output,"infercnv_res/BayesNetOutput.HMMi6.leiden.hmm_mode-subclusters/CNV_State_Probabilities.dat"),
  header=T, sep="\t")

cell_groupings <- read.table(
  paste0(params$output,"infercnv_res/17_HMM_predHMMi6.leiden.hmm_mode-subclusters.cell_groupings"),
  header=T, sep="\t")
```

## XClone from the CellSNP lite results

```{r}
# Xclone
meta <- data@meta.data %>%
  dplyr::select(Annotation)
meta$cell_id <- rownames(meta)
meta <- meta[, ncol(meta):1]
meta <- as.data.frame(meta)
meta$cell_type <- ifelse(meta$Annotation %in% c("CD4 T","CD8 T","Monocyte","Neutrophil","NK"), "Normal", "Tumor")
meta$sample <- sub("_.*", "", meta$cell_id)
meta$cell <- sub(".*_", "", meta$cell_id)
write.table(meta,"Xclone.tsv",sep = "\t",quote = FALSE,col.names = TRUE, row.names = FALSE)

```

## Cell SNP lite - AF clustering

```{r}
timepoints <- c("D1", "D4", "D8", "D11", "D14")
base_path <- "/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/Cell_snp_lite"

# Step 1 : AF per timepoint
# Apply to each tiempoint
af_list <- lapply(timepoints, process_timepoint)
names(af_list) <- timepoints

# Step 2 : combine matrix
# Find common SNPs to all tiempoints  
snp_ids <- lapply(af_list, rownames)
common_snps <- Reduce(intersect, snp_ids)
cat("\nSNPs common to all  timepoints :", length(common_snps), "\n")
AF_combined <- do.call(cbind, lapply(af_list, function(af) af[common_snps, ]))
cat("Matrix combined :", nrow(AF_combined), "SNPs x", ncol(AF_combined), "cells\n")

# Steps 3 : imputation + PCA + UMAP
# Imputation per mean of the SNP
AF_imputed <- AF_combined
for (i in 1:nrow(AF_combined)) {
  AF_imputed[i, is.na(AF_combined[i, ])] <- mean(AF_combined[i, ], na.rm = TRUE)
}

# PCA
pca_res <- prcomp(t(AF_imputed), center = TRUE, scale. = FALSE)
n_pcs <- min(20, ncol(pca_res$x))
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

# Scree plot
plot(var_explained[1:20], type = "b", pch = 19,
     xlab = "PC", ylab = "% explained variance",
     main = "Scree plot")

# UMAP
umap_res <- umap(pca_res$x[, 1:n_pcs],
                 n_neighbors = 15, min_dist = 0.3, metric = "euclidean")

# Final dataframe
tp_labels <- rep(timepoints, times = sapply(af_list, ncol))

umap_df <- data.frame(
  UMAP1     = umap_res[, 1],
  UMAP2     = umap_res[, 2],
  PC1       = pca_res$x[, 1],
  PC2       = pca_res$x[, 2],
  timepoint = tp_labels
)

# PCA
ggplot(umap_df, aes(x = PC1, y = PC2, color = timepoint)) +
  geom_point(size = 0.8, alpha = 0.7) +
  theme_classic() +
  labs(title = "PCA SNP",
       x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
       y = paste0("PC2 (", round(var_explained[2], 1), "%)"))

# UMAP colorée par timepoint
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = timepoint)) +
  geom_point(size = 0.8, alpha = 0.7) +
  theme_classic() +
  labs(title = "UMAP SNP")

## 1. Hierarchical clustering : 
dist_mat <- dist(pca_res$x[, 1:n_pcs])
hc <- hclust(dist_mat, method = "ward.D2")
# 1st : visualize
plot(hc, labels = FALSE, main = "Dendrogram (cells)")
# Choose the number of clusters/clones
k <- 4
clone_labels <- cutree(hc, k = k)
umap_df$clone <- factor(clone_labels)
# visualize
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = clone)) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  theme_classic() +
  labs(title = "Clones SNP")

## 2. K-means clustering : 
# "elbow" method
set.seed(42)
wss <- sapply(1:15, function(k) {
  kmeans(pca_res$x[, 1:n_pcs], centers = k, nstart = 25, iter.max = 100)$tot.withinss
})

elbow_df <- data.frame(k = 1:15, wss = wss)
plot_elbow <- ggplot(elbow_df, aes(x = k, y = wss)) +
  geom_line() + geom_point(size = 2) +
  scale_x_continuous(breaks = 1:15) +
  theme_classic() +
  labs(title = "Elbow plot – K-means",
       x = "Clusters number (k)", y = "WSS (intra inertia)")
ggsave(paste0(params$output, "kmeans_elbow_", params$patient_infercnv, ".jpeg"),plot_elbow, width = 3000, height = 1500, units = "px")

# K-means 
k <- 4 
set.seed(42)
km_res <- kmeans(pca_res$x[, 1:n_pcs], centers = k, nstart = 50, iter.max = 200)

umap_df$clone_kmeans <- factor(km_res$cluster)

plot_km <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = clone_kmeans)) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  theme_classic() +
  labs(title = "Clones SNP – K-means")
ggsave(paste0(params$output, "umap_clone_kmeans_", params$patient_infercnv, ".jpeg"),plot_km, width = 3000, height = 1500, units = "px")
```

```{r}
# Add to the seurat object the clones 
AF_clones<-km_res[["cluster"]]
AF_clones<-data.frame(AF_clones)
data <- get(paste0("data_", params$patient_infercnv))
data <- AddMetaData(data, metadata = AF_clones)
DimPlot(data, group.by = "AF_clones", reduction = "umap", pt.size=0.1, label.size = 2.5 ) + ggtitle(str_wrap("Clones clustering based on allelic frequency", width = 60)) + theme(plot.title = element_text(size = 15, hjust = 0.5))
```

```{r}

tumor_prop<-data@meta.data
tumor_prop_filtered <- tumor_prop %>%
  filter(Annotation %in% c("B","B intermediate","B memory","B naive","Macrophage","Bridge", "Monocyte")) %>%
  mutate(
    Annotation = recode(
      Annotation,
      "Bridge" = "Intermediate population"))
tumor_prop_filtered$AF_clones<-as.factor(tumor_prop_filtered$AF_clones)

ggplot(tumor_prop_filtered,
       aes(x = Annotation,
           fill = AF_clones)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent,
                     limits = c(0, 1)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1,size=13)) +
  ylab("Timepoint proportion") + xlab(" ")+
  labs(fill = "Clones")
```
