---
title: "infer_CNA"
params: 
  patient: "P2"
  patientbis: "P3"
  patient_numbat: "p2"
  file: "~/Documents/Singlecell-R/Scripts/Matrix/CLL_Seurat_P2.rds"
  filebis: "~/Documents/Singlecell-R/Scripts/Matrix/CLL_Seurat_P3"
format: html
editor: visual
---

## Copy Number Alteration calculation

Goal : track the cancer cell thank to the CNA

```{r}
#| message: false
library(Seurat)
library(copykat)
library(ggplot2)
library(stringr)
library(numbat)
library(data.table)
library(dplyr)
library(glue)
library(ggtree)
library(tidygraph)
library(patchwork)
```

```{r}
data_p2<-readRDS(params$file)
data_p3<-readRDS(params$filebis)
```

## CopyKat

Run copyKat (timepoint separately)

Patient 2

```{r}
timepoints <- c("D1", "D4", "D8", "D11", "D14")
for (tp in timepoints) {

  data_tp <- subset(data_p2, subset = timepoint == tp)
  counts_matrix <- GetAssayData(data_tp, assay = "RNA", layer = "counts")
  norm.cells <- colnames(data_tp)[data_tp$Annotation %in% c("Monocyte", "CD4 T", "CD8 T", "NK")]
  
  ck <- copykat(
    rawmat = counts_matrix,
    id.type = "S",
    ngene.chr = 5,
    win.size = 25,
    KS.cut = 0.1,
    sam.name = tp,
    distance = "euclidean",
    norm.cell.names = norm.cells,
    output.seg = "FALSE",
    plot.genes = "TRUE",
    genome = "hg20",
    n.cores = 1
  )
  
  copykat_results[[tp]] <- ck
  pred_results[[tp]] <- data.frame(ck$prediction)
  cna_results[[tp]] <- data.frame(ck$CNAmat)
}
```

Patient 3

```{r}
timepoints <- c("D1", "D4", "D8", "D11", "D14")
for (tp in timepoints) {

  data_tp <- subset(data_p3, subset = timepoint == tp)
  counts_matrix <- GetAssayData(data_tp, assay = "RNA", layer = "counts")
  norm.cells <- colnames(data_tp)[data_tp$Annotation %in% c("Monocyte", "NK")]
  
  ck <- copykat(
    rawmat = counts_matrix,
    id.type = "S",
    ngene.chr = 5,
    win.size = 25,
    KS.cut = 0.1,
    sam.name = tp,
    distance = "euclidean",
    norm.cell.names = norm.cells,
    output.seg = "FALSE",
    plot.genes = "TRUE",
    genome = "hg20",
    n.cores = 1
  )
  
  copykat_results[[tp]] <- ck
  pred_results[[tp]] <- data.frame(ck$prediction)
  cna_results[[tp]] <- data.frame(ck$CNAmat)
}
```

Exploring the results

Add the results in the Seurat Object, day sep

```{r}
pred_D1 <- data.frame(copykat_D1$prediction)
data$copykat_pred <- NA
rownames(pred.res_D1) <- pred_D1$cell.names
data$copykat_pred <- pred_D1[colnames(data), "copykat.pred"]
```

```{r}
DimPlot(data, group.by = "copykat_pred", reduction = "umap", pt.size=0.1, label.size = 2.5) + ggtitle(str_wrap("UMAP", width = 60)) + theme(plot.title = element_text(size = 15, hjust = 0.5))

```

Timepoints combined

Patient 2

```{r}
copykat.res <- do.call(rbind, pred_results)data_p2$copykat_pred <- NA
rownames(copykat.res) <- copykat.res$cell.names
data_p2$copykat_pred <- copykat.res[colnames(data_p2), "copykat.pred"]

title=paste("UMAP CNA, patient", params$patient)
plot<-DimPlot(data_p2, group.by = "copykat_pred", reduction = "umap", pt.size=0.05, label.size = 2.5) + ggtitle(str_wrap(title, width = 60)) + theme(plot.title = element_text(size = 10, hjust = 0.5))
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/cna_", params$patient, ".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

Patient 3

```{r}
copykat.res <- do.call(rbind, pred_results)data_p3$copykat_pred <- NA
rownames(copykat.res) <- copykat.res$cell.names
data_p3$copykat_pred <- copykat.res[colnames(data_p3), "copykat.pred"]

title=paste("UMAP CNA, patient", params$patientbis)
plot<-DimPlot(data_p3, group.by = "copykat_pred", reduction = "umap", pt.size=0.05, label.size = 2.5) + ggtitle(str_wrap(title, width = 60)) + theme(plot.title = element_text(size = 10, hjust = 0.5))
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/cna_", params$patientbis, ".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

## Numbat

Patient 2

Barcodes files

```{r}
timepoints <- c("D1", "D4", "D8", "D11", "D14")
for (tp in timepoints) {
  data_sub<-subset(data_p2, subset = timepoint == tp)
  counts_matrix <- GetAssayData(data_sub, assay = "RNA", layer = "counts")
  barcodes <- colnames(counts_matrix)
  barcodes <- sub("^D[0-9]+_", "", barcodes)
  write.table(barcodes, paste0("~/Documents/Singlecell-R/Scripts/Data/barcodes_", tp, ".tsv"), sep="\t",row.names = FALSE, col.names = FALSE)
}
```

Run the pile up and phasing on Numbat Docker in the HPC : run the script Numbat.sh

Combining the pileup & phasing out (because one per timepoint) + add the timepoint prefix on the ID cell

```{r}
files <- c(
  "/home/ablanc/work/NetBIO2/CNA/Numbat_p2/Results_pileup_phasing/D1_allele_counts.tsv.gz",
  "/home/ablanc/work/NetBIO2/CNA/Numbat_p2/Results_pileup_phasing/D4_allele_counts.tsv.gz",
  "/home/ablanc/work/NetBIO2/CNA/Numbat_p2/Results_pileup_phasing/D8_allele_counts.tsv.gz",
  "/home/ablanc/work/NetBIO2/CNA/Numbat_p2/Results_pileup_phasing/D11_allele_counts.tsv.gz",
  "/home/ablanc/work/NetBIO2/CNA/Numbat_p2/Results_pileup_phasing/D14_allele_counts.tsv.gz"
)

timepoints <- sub(".*/(D\\d+)_allele_counts.*", "\\1", files)

df_allele_list <- lapply(seq_along(files), function(i) {
  fread(files[i]) %>%
    mutate(
      timepoint = as.integer(sub("D", "", timepoints[i])),
      cell = paste0(timepoints[i], "_", cell)
    )
})

df_allele <- bind_rows(df_allele_list)

head(df_allele)
```

Count matrix

```{r}
counts_matrix <- GetAssayData(data_p2, assay = "RNA", layer = "counts")
#colnames(counts_matrix) <- sub("^D[0-9]+_", "", colnames(counts_matrix))
```

Reference created (or choose the given one)

```{r}
data_ref<-subset(data_p2, subset = (Annotation %in% c("CD4 T", "CD8 T", "NK", "Monocyte")))
cell_annot <- data.frame(
  cell = colnames(data_ref),
  group = data_ref$Annotation,
  stringsAsFactors = FALSE
)
counts_ref <- GetAssayData(data_ref, assay = "RNA", layer = "counts")

ref_internal <- aggregate_counts(counts_ref, cell_annot)
```

### Running Numbat

In run it on the HPC, I used 8 CPU and 200 G, time : \~15h.

```{r}
df_allele<-read.table("~/Documents/Singlecell-R/Scripts/Matrix/all_allele_counts.tsv", sep=",", header=T)
df_allele <- fread("~/Documents/Singlecell-R/Scripts/Matrix/all_allele_counts.tsv.gz")
```

```{r}
out = run_numbat(
    counts_matrix, # gene x cell integer UMI count matrix 
    ref_internal, # reference expression profile, a gene x cell type normalized expression level matrix
    df_allele, # allele dataframe generated by pileup_and_phase script
    genome = "hg38",
    t = 1e-5,
    ncores = 8,
    ncores_nni = 8,
    skip_nj = TRUE,
    plot = TRUE,
    max_entropy = 0.7,
    out_dir = '/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/Resultats_numbat/Results_numbat_ref_internal_p2'
)
```

### Interpreting Numbat results 

Put the Numbat folder result and then use the parameter \$patient_numbat in the script header.

```{r}
nb = Numbat$new(out_dir = paste0("/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/Resultats_numbat/Results_numbat_ref_internal_",params$patient_numbat))

if (params$patient_numbat =="p2") {
  data<-data_p2
} else {
  data<-data_p3
}
```

```{r}
mypal = c('1' = 'gray', '2' = "#377EB8", '3' = "#4DAF4A", '4' = "#984EA3", '5' ="#ea4258", '6'="orange")
  
nb$plot_phylo_heatmap(
  clone_bar = TRUE, 
  p_min = 0.9,
  pal_clone = mypal
)
```

```{r}
nb$bulk_clones %>% 
  filter(n_cells > 50) %>%
  plot_bulks(
    min_LLR = 5, # filtering CNVs by evidence
    legend = TRUE
  )
```

```{r}
head(nb$joint_post) %>% select(cell, CHROM, seg, cnv_state, p_cnv, p_cnv_x, p_cnv_y)
```

```{r}
# Add metadata to the seurat object 
meta_subset <- nb$clone_post[, c("cell", "clone_opt", "GT_opt", "p_opt", "p_cnv", "compartment_opt")]
rownames(meta_subset) <- meta_subset$cell
meta_subset$cell <- NULL
head(colnames(data))
head(rownames(meta_subset))
data <- AddMetaData(data, metadata = meta_subset)
length(intersect(colnames(data), rownames(meta_subset)))
```

```{r}
plot<-DimPlot(data, group.by = "clone_opt", reduction = "umap", pt.size=0.1, label.size = 2.5, cols =c("gray","#377EB8","#4DAF4A", "#984EA3", "#ea4258","orange") ) + ggtitle(str_wrap("Clonal assignments", width = 60)) + theme(plot.title = element_text(size = 15, hjust = 0.5))
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/umap_clone_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")

plot<-DimPlot(data, group.by = "compartment_opt", reduction = "umap", pt.size=0.1, label.size = 2.5, cols =c("lightblue","#ea4258") ) + ggtitle(str_wrap("Tumor and normal cells annotation", width = 60)) + theme(plot.title = element_text(size = 15, hjust = 0.5))
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/umap_tumor_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")

plot<-FeaturePlot(data, features="p_cnv", reduction="umap")+ ggtitle(str_wrap("Tumor versus normal probability", width = 60)) + theme(plot.title = element_text(size = 15, hjust = 0.5))
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/umap_proba_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

```{r}
df <- data@meta.data
tumor_prop <- df %>%
  group_by(Annotation, compartment_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Annotation) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(tumor_prop, aes(x = Annotation, y = prop, fill = compartment_opt)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Cell Type") + labs(fill = "Cell annotation\n(tumor/normal)")
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/propor_tumor_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

```{r}
df$clone_opt <- as.factor(df$clone_opt)
clone_prop <- df %>%
  group_by(Annotation, clone_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Annotation) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(clone_prop, aes(x = Annotation, y = prop, fill = clone_opt)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Cell Type") + labs(fill = "Cell clone")
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/propor_clone_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

```{r}
GT_prop <- df %>%
     group_by(Annotation, GT_opt) %>%
     summarise(n = n(), .groups = "drop") %>%
     group_by(Annotation) %>%
     mutate(prop = n / sum(n))
plot<-ggplot(GT_prop, aes(x = Annotation, y = prop, fill = GT_opt)) +
     geom_bar(stat = "identity", position = "fill") +
     scale_y_continuous(labels = scales::percent) +
     theme_classic() +
     theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
     ylab("Cell proportion") + xlab("Cell Type") + labs(fill = "Alteration type\n(chromosomes)")
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/propor_mut_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")

df$timepoint <- factor(
  df$timepoint,
  levels = c("D1", "D4", "D8", "D11", "D14")
)

time_GT_prop <- df %>%
     group_by(timepoint, GT_opt) %>%
     summarise(n = n(), .groups = "drop") %>%
     group_by(timepoint) %>%
     mutate(prop = n / sum(n))

ggplot(time_GT_prop, aes(x = timepoint, y = prop, fill = GT_opt)) +
     geom_bar(stat = "identity", position = "fill") +
     scale_y_continuous(labels = scales::percent) +
     theme_classic() +
     theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
     ylab("Cell proportion") + xlab("Cell Type") + labs(fill = "Alteration type\n(chromosomes)")

data_p2$GT_11c <- ifelse(data_p2$GT_opt %in% c("7b,11c","12c_bdel", "7b,11c,12c_del"), "11c", "other")
DimPlot(data, group.by = "GT_11c", reduction = "umap", pt.size=0.1, label.size = 2.5, cols =c("#ea4258","lightblue") ) + ggtitle(str_wrap("Tumor and normal cells annotation", width = 60)) + theme(plot.title = element_text(size = 15, hjust = 0.5))
```

Timepoint by clones

```{r}
clone_prop <- df %>%
     group_by(timepoint, clone_opt) %>%
     summarise(n = n(), .groups = "drop") %>%
     group_by(timepoint) %>%
     mutate(prop = n / sum(n))
clone_prop$timepoint <- factor(clone_prop$timepoint,levels = c("D1","D4","D8","D11","D14"))

plot<-ggplot(clone_prop, aes(x = timepoint, y = prop, color = factor(clone_opt), group = clone_opt)) +
     geom_line(size = 1.2) +
     geom_point(size = 3) +
     scale_y_continuous(labels = scales::percent) +
     theme_classic() +
     labs(x = "Timepoint",y = "Cell proportion",color = "Clone")
plot
ggsave(paste0("~/Documents/Singlecell-R/Scripts/Plots/propor_time_clone_",params$patient_numbat,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

Tumor phylogeny

```{r}
nb$plot_sc_tree(
  label_size = 3, 
  branch_width = 0.5, 
  tip_length = 0.5,
  pal_clone = mypal, 
  tip = TRUE
)

nb$plot_mut_history(pal = mypal)
```
