---
title: "V(D)J analysis"
params: 
  patient: "P2"
  file: "~/Documents/Singlecell-R/Scripts/Matrix/CLL_Seurat_P2.rds"
  folder_trust4: "~/Documents/Singlecell-R/Scripts/Matrix/VDJ_TRUST4/"
  output: "~/Documents/Singlecell-R/Scripts/Plots/"
format: html
editor: visual
---

## V(D)J analysis: track clonotypes

Analysis done with TRUST4 on the BAM, timepoint separately

Explore the results after the TRUST4 run in the script bash VDJ_TRSUT4.sh

Libraries and data

```{r}
#| message: false
library(Seurat)
library(ggplot2)
library(dplyr)
library(tidyr)
library(pheatmap)
source("functions.R")
```

```{r}
#| echo: false

#Load the TRUST4 output files 
days <- c("D1", "D4", "D8", "D11", "D14")

for (day in days) {
  df <- read.table(paste0(params$folder_trust4, "vdj_p2_", day, "_barcode_report.tsv"),header = FALSE, sep = "\t", stringsAsFactors = FALSE)
  colnames(df) <- c("Barcodes", "Celltype", "Chain_1", "Chain_2", "Secondary_chain_1", "Secondary_chain_2")
  df$Barcodes <- paste0(day, "_", df$Barcodes)
  assign(day, df)
}

for (day in days) {
  df <- read.table(paste0(params$folder_trust4, "vdj_p2_", day, "_barcode_airr.tsv"),header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  df$Barcodes <- paste0(day, "_", df$cell_id)
  assign(paste0(day,"_airr"), df)
}
```

```{r}
data<-readRDS(params$file)
barcodes <- colnames(data)
```

```{r}
# keep only the cells present in the Seurat object
for (day in days) {
  df <- get(day)
  df <- df[df$Barcodes %in% barcodes, ]
  assign(day, df)
}

days_airr <- c("D1_airr", "D4_airr", "D8_airr", "D11_airr", "D14_airr")
for (day in days_airr) {
  df <- get(day)
  df <- df[df$Barcodes %in% barcodes, ]
  assign(day, df)
}

# Add the TRUST4 results into the seurat object
combined_vdj <- do.call(rbind, lapply(days, get))
combined_vdj <- combined_vdj[combined_vdj$Barcodes %in% colnames(data), ]
rownames(combined_vdj) <- combined_vdj$Barcodes
combined_vdj$Barcodes <- NULL
data <- AddMetaData(data, metadata = combined_vdj)
```

### B and T annotation by TRUST4

```{r}
plot<-DimPlot(data, reduction="umap", group.by = "Celltype", cols =c("#ea4258","darkgreen","orange"),na.value = "lightblue") + ggtitle("VDJ annotation") + theme(plot.title = element_text(size = 15, hjust = 0.5))
ggsave(paste0(params$output,"vdj_umap.jpeg"), plot, width = 3000, height = 2000, units = "px")
```

Compare annotated B cells by TRUST4 and annotated tumoral cells by Numbat in T cells population

```{r}
data_T<-subset(data, subset= Annotation %in% c("T", "CD4 T", "CD8 T"))
# Compare B cells and tumoral cells from CNA analysis 
meta <- data_T@meta.data %>%
  as_tibble(rownames = "barcode") %>%
  select(barcode, Celltype, compartment_opt) %>%
  filter(
    !is.na(Celltype),
    !is.na(compartment_opt))

# Celltype × compartment_opt
ctab <- meta %>% count(Celltype, compartment_opt, name = "n_cells") %>% arrange(desc(n_cells))
ctab_wide <- ctab %>% pivot_wider(names_from = compartment_opt, values_from = n_cells, values_fill = 0)

# Plots
ctab_long <- ctab_wide %>%
  pivot_longer(!Celltype, names_to = "compartment_opt", values_to = "n_cells")

cols_fun <- colorRampPalette(c("white", "red"))

p_heatmap <- ggplot(ctab_long, aes(
  x = compartment_opt,
  y = Celltype,
  fill = n_cells)) +
  geom_tile(color = "gray95", size = 0.2) +
  scale_fill_gradientn(
    colors = cols_fun(100),
    na.value = "white") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 20),axis.text.y = element_text(size = 20),plot.title = element_text(size = 20, hjust = 0.5), legend.position = "right", legend.text = element_text(size = 20)) +labs(title = "Cell type V(D)J vs tumor annotation CNA",x = "Tumor annotation (CNA)",y = "Lymphoid annotation V(D)J",fill = "Cell number")

p_bar <- ggplot(ctab, aes(x = compartment_opt,y = n_cells,fill = Celltype)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 20),axis.text.y = element_text(size = 20),legend.text = element_text(size = 20), plot.title = element_text(size = 20, hjust = 0.5)) +
  labs(
    title = "Stacked barplot cell type V(D)J per tumor annotation CNA",
    x = "compartment_opt",
    y = "Cell fraction",
    fill = "Cell type")

ggsave(paste0(params$output,"vdj_cna_heatmap.jpeg"), p_heatmap, width = 3000, height = 2000, units = "px")
ggsave(paste0(params$output,"vdj_cna_bar.jpeg"), p_bar, width = 3000, height = 2000, units = "px")

plot<-ggplot(meta,
       aes(axis1 = compartment_opt,
           axis2 = Celltype)) +
  geom_alluvium(aes(fill = compartment_opt)) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size=6)+ labs(fill = "CNA annotation",title = "CNA and V(D)J annotation in the T population") +
  theme_minimal()+ theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 20),axis.text.y = element_text(size = 20),legend.text = element_text(size = 20), plot.title = element_text(size = 20, hjust = 0.5), legend.title = element_text(size = 20),)
ggsave(paste0(params$output,"vdj_cna_flux.jpeg"), plot, width = 3000, height = 2000, units = "px")
```

Compare annotated B cells by TRUST4 and annotated tumoral cells by Numbat in monocyte population

```{r}
data_myel<-subset(data, subset= Annotation %in% c("Monocyte"))
# Compare B cells and tumoral cells from CNA analysis 
meta <- data_myel@meta.data %>%
  as_tibble(rownames = "barcode") %>%
  select(barcode, Celltype, compartment_opt) %>%
  filter(
    !is.na(Celltype),
    !is.na(compartment_opt))

# Celltype × compartment_opt
ctab <- meta %>% count(Celltype, compartment_opt, name = "n_cells") %>% arrange(desc(n_cells))
ctab_wide <- ctab %>% pivot_wider(names_from = compartment_opt, values_from = n_cells, values_fill = 0)

# Plots
ctab_long <- ctab_wide %>%
  pivot_longer(!Celltype, names_to = "compartment_opt", values_to = "n_cells")

cols_fun <- colorRampPalette(c("white", "red"))

p_heatmap <- ggplot(ctab_long, aes(
  x = compartment_opt,
  y = Celltype,
  fill = n_cells)) +
  geom_tile(color = "gray95", size = 0.2) +
  scale_fill_gradientn(
    colors = cols_fun(100),
    na.value = "white") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 20),axis.text.y = element_text(size = 20),plot.title = element_text(size = 20, hjust = 0.5), legend.position = "right",legend.text = element_text(size = 20)) +labs(title = "Cell type V(D)J vs tumor annotation CNA",x = "Tumor annotation (CNA)",y = "Lymphoid annotation V(D)J",fill = "Cell number")

p_bar <- ggplot(ctab, aes(x = compartment_opt,y = n_cells,fill = Celltype)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 20),axis.text.y = element_text(size = 20),plot.title = element_text(size = 20, hjust = 0.5), legend.position = "right",legend.text = element_text(size = 20)) +
  labs(
    title = "Stacked barplot cell type V(D)J per tumor annotation CNA",
    x = "compartment_opt",
    y = "Cell fraction",
    fill = "Cell type")


ggsave(paste0(params$output,"vdj_cna_heatmap_myel.jpeg"), p_heatmap, width = 3000, height = 2000, units = "px")
ggsave(paste0(params$output,"vdj_cna_bar_myel.jpeg"), p_bar, width = 3000, height = 2000, units = "px")

plot<-ggplot(meta,
       aes(axis1 = compartment_opt,
           axis2 = Celltype)) +
  geom_alluvium(aes(fill = compartment_opt)) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size=6)+ labs(fill = "CNA annotation",title = "CNA and V(D)J annotation in the monocyte population") +
  theme_minimal()+ theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 20),axis.text.y = element_text(size = 20),legend.text = element_text(size = 20), plot.title = element_text(size = 20, hjust = 0.5), legend.title = element_text(size = 20),)
ggsave(paste0(params$output,"vdj_cna_flux_myel.jpeg"), plot, width = 3000, height = 2000, units = "px")
```

### Clonotypes and SHM (Somatic hypermutation rate of BCR)

```{r}
# Filter the dataframe TRUST4 output
combined_vdj <- do.call(rbind, lapply(days_airr, get))
combined_vdj <-combined_vdj[, -c(1,2,3,4, 10, 11, 12, 13, 14, 16, 17, 18, 19, 21, 22, 23)]
  
# Separate the IGH et IGK/L
heavy <- subset(combined_vdj, locus == "IGH")
light <- subset(combined_vdj, locus %in% c("IGK","IGL"))

# 1 chain per cell
heavy <- heavy %>%
group_by(Barcodes) %>%
slice_max(consensus_count, n=1)

light <- light %>%
group_by(Barcodes) %>%
slice_max(consensus_count, n=1)

# Combine IGH et IGK/L 
bcr <- left_join(light, heavy, by="Barcodes", suffix=c("_L","_H"))
```

Define clonotypes

junction = CDR3 (most variable zone)

Different clonotypes definition :

-   basic : junction on the heavy chain (IGH) or on the light chain (IGK/L)

-   more robust : junction on the heavy chain and light chain (IGK/L)

-   robust : variable zone + junction zone on the heavy and light chain

```{r}
# more light chain than heavy chain in the output 
# Add the TRUST4 results into the seurat object
bcr <-bcr[, -c(3, 5, 8, 12, 14, 17)]
bcr<-data.frame(bcr)
rownames(bcr) <- bcr$Barcodes
bcr$Barcodes <- NULL
data <- AddMetaData(data, metadata = bcr)
```

```{r}
plot<-DimPlot(data, reduction="umap", group.by = "j_call_L",na.value = "lightblue") + ggtitle("Germinal gene junction on light chain") + theme(plot.title = element_text(size = 15, hjust = 0.5))
ggsave(paste0(params$output,"vdj_clono_umap_light.jpeg"), plot, width = 3000, height = 2000, units = "px")
```

Combine Variable (v) and junction (j) on light chain

```{r}
data$v_j_Light<-paste(data$v_call_L, data$j_call_L, sep="_")

counts <- table(data@meta.data$v_j_Light)
keep_annotations <- names(counts[counts > 100])

data@meta.data$v_j_Light_filtered <- ifelse(
  data@meta.data$v_j_Light %in% keep_annotations,
  data@meta.data$v_j_Light,
  "Others (<100 cells)")


DimPlot(data, reduction = "umap", group.by = "v_j_Light_filtered",cols=c("red", "orange", "darkgreen", "blue","lightblue", "grey", "grey"), na.value = "lightblue") + ggtitle("Clonotypes VJ Light chain (>100 cells)") + theme(plot.title = element_text(size = 15, hjust = 0.5))
```

Combine junction (j) on light chain and heavy chain

```{r}
data$j_j_light_heavy<-paste(data$j_call_L, data$j_call_H, sep="_")

counts <- table(data@meta.data$j_j_light_heavy)
keep_annotations <- names(counts[counts > 100])

data@meta.data$j_j_light_heavy_filtered <- ifelse(
  data@meta.data$j_j_light_heavy %in% keep_annotations,
  data@meta.data$j_j_light_heavy,
  "Others (<100 cells)")


DimPlot(data, reduction = "umap", group.by = "j_j_light_heavy_filtered",cols=c("red", "orange", "darkgreen", "blue","lightblue","yellow", "purple", "grey", "grey"), na.value = "lightblue") + ggtitle("Clonotypes VJ Light chain (>100 cells)") + theme(plot.title = element_text(size = 15, hjust = 0.5))
```

CDR3 junction_aa

```{r}
tab <- table(data$junction_aa_L)
keep <- names(tab[tab > 30])
data_filt <- subset(data, junction_aa_L %in% keep)
tab<-table(data_filt$junction_aa_L)
cdr3 <- names(tab)

data$highlight_cdr3_L <- ifelse(
  data$junction_aa_L %in% cdr3,
  data$junction_aa_L,
  "Others"
)

plot<-DimPlot(data, reduction = "umap", group.by = "highlight_cdr3_L",cols=c("black","#12A8F5", "#E58B00", "#6A0DAD","#006400", "grey"), na.value = "lightblue") + ggtitle("Clonotypes CDR3 (main CDR3) on light chain") + theme(plot.title = element_text(size = 15, hjust = 0.5))
ggsave(paste0(params$output,"vdj_clono_cdr3_umap_light.jpeg"), plot, width = 3000, height = 2000, units = "px")

clone_matrix <- table(data$Annotation, 
                     ifelse(data$junction_aa_L %in% cdr3, data$junction_aa_L, "Other"))
mat <- clone_matrix[, colnames(clone_matrix) != "Other"]
mat_log <- log1p(mat)
mat_log <- mat_log[apply(mat_log, 1, var) > 0, ]
plot<-pheatmap(mat_log,scale = "row",main = "Main CDR3 per clusters",angle_col = 45,fontsize = 18,fontsize_row = 16,fontsize_col = 16)
ggsave(paste0(params$output,"vdj_clono_cdr3_heatmap_light.jpeg"), plot, width = 4000, height = 2500, units = "px")
```

```{r}
tab <- table(data$junction_aa_H)
keep <- names(tab[tab > 30])
data_filt <- subset(data, junction_aa_H %in% keep)
tab<-table(data_filt$junction_aa_H)
cdr3 <- names(tab)

data$highlight_cdr3_H <- ifelse(
  data$junction_aa_H %in% cdr3,
  data$junction_aa_H,
  "Others"
)

plot<-DimPlot(data, reduction = "umap", group.by = "highlight_cdr3",cols=c("#12A8F5", "#E58B00", "#6A0DAD","#006400",  "grey"), na.value = "lightblue") + ggtitle("Clonotypes CDR3 (main CDR3) on heavy chain") + theme(plot.title = element_text(size = 15, hjust = 0.5))
ggsave(paste0(params$output,"vdj_clono_cdr3_umap_heavy.jpeg"), plot, width = 3000, height = 2000, units = "px")

clone_matrix <- table(data$Annotation, 
                     ifelse(data$junction_aa_L %in% cdr3, data$junction_aa_L, "Other"))
mat <- clone_matrix[, colnames(clone_matrix) != "Other"]
mat_log <- log1p(mat)
mat_log <- mat_log[apply(mat_log, 1, var) > 0, ]
mat_scaled <- t(scale(t(mat_log)))
# NA/Inf
mat_scaled <- mat_scaled[apply(mat_scaled, 1, function(x) all(is.finite(x))), ]

pheatmap(mat_scaled,
         scale = "none",
         main = "Main CDR3 per clusters")
ggsave(paste0(params$output,"vdj_clono_cdr3_heatmap_heavy.jpeg"), plot, width = 3000, height = 2000, units = "px")
```

Subset myeloid and B pop

```{r}
# On the subset myeloid and B populations 
data_myel<-subset(data, subset = (Annotation %in% c("B intermediate", "B naive", "B memory", "B", "Bridge","Monocyte","Macrophage")))
data_myel<-run_dim_red(data_myel)

plot<-DimPlot(data_myel, reduction="umap", group.by = "j_call_L",na.value = "lightblue") + ggtitle("Germinal gene junction on light chain on B and myeloid population") + theme(plot.title = element_text(size = 15, hjust = 0.5))
ggsave(paste0(params$output,"vdj_clono_umap_light_myel.jpeg"), plot, width = 3000, height = 2000, units = "px")

plot<-DimPlot(data_myel, reduction = "umap", group.by = "highlight_cdr3_L",cols=c("black","#12A8F5", "#E58B00", "#6A0DAD","#006400", "grey"), na.value = "lightblue") + ggtitle("Clonotypes CDR3 on light chain on B and myeloid population") + theme(plot.title = element_text(size = 15, hjust = 0.5))
ggsave(paste0(params$output,"vdj_clono__cdr3_umap_light_myel.jpeg"), plot, width = 3000, height = 2000, units = "px")
```

```{r}
data<-FindClusters(data, resolution = 0.2)
DimPlot(data, reduction="umap")
```

SHM

```{r}

shm_df <- data.frame(
  v_call_L = data$v_identity_L,
  Annotation=data$Annotation,
  shm_rate = 1-data$v_identity_L  # ou calculez: 1 - v_call_L
)


ggplot(shm_df, aes(x = v_call_L*100)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = median(shm_rate*100)), 
             color = "red", linetype = "dashed", size = 1) +
  labs(x = "Taux de SHM (%)", y = "Nombre de clonotypes", 
       title = "Distribution du SHM par clonotype") +
  theme_minimal()

ggplot(shm_df, aes(x = v_call_L*100, fill = Annotation)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = tapply(shm_df$v_call_L*100, shm_df$macrostate, median, na.rm = TRUE),
             color = "red", linetype = "dashed") +
  facet_wrap(~macrostate, scales = "free_y") +
  labs(x = "Taux de SHM (%)", y = "Nombre de cellules") +
  theme_minimal()
```
