---
title: "V(D)J analysis"
params: 
  patient: "p3" # format p2 or p3
  file: "~/Documents/Singlecell-R/Scripts/Matrix/CLL_Seurat_P3.rds"
  folder_trust4: "~/Documents/Singlecell-R/Scripts/Matrix/VDJ_TRUST4/"
  output: "~/Documents/Singlecell-R/Scripts/Plots/"
format: html
editor: visual
---

## V(D)J analysis: track clonotypes

Analysis done with TRUST4 on the BAM, timepoint separately

Explore the results after the TRUST4 run in the script bash VDJ_TRSUT4.sh

Libraries and data

```{r}
#| message: false
library(Seurat)
library(ggplot2)
library(dplyr)
library(tidyr)
library(pheatmap)
library(ggalluvial)
library(tibble)
library(ggpubr)
library(ape)
library(ggtree)
library(stringdist)
source("functions.R")
```

```{r}
#| echo: false

#Load the TRUST4 output files 
days <- c("D1", "D4", "D8", "D11", "D14")

for (day in days) {
  df <- read.table(paste0(params$folder_trust4, "vdj_",params$patient,"_", day, "_barcode_report.tsv"),header = FALSE, sep = "\t", stringsAsFactors = FALSE)
  colnames(df) <- c("Barcodes", "Celltype", "Chain_1", "Chain_2", "Secondary_chain_1", "Secondary_chain_2")
  df$Barcodes <- paste0(day, "_", df$Barcodes)
  assign(day, df)
}

for (day in days) {
  df <- read.table(paste0(params$folder_trust4, "vdj_",params$patient,"_", day, "_barcode_airr.tsv"),header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  df$Barcodes <- paste0(day, "_", df$cell_id)
  assign(paste0(day,"_airr"), df)
}
```

```{r}
data<-readRDS(params$file)
barcodes <- colnames(data)
```

```{r}
# keep only the cells present in the Seurat object
for (day in days) {
  df <- get(day)
  df <- df[df$Barcodes %in% barcodes, ]
  assign(day, df)
}

days_airr <- c("D1_airr", "D4_airr", "D8_airr", "D11_airr", "D14_airr")
for (day in days_airr) {
  df <- get(day)
  df <- df[df$Barcodes %in% barcodes, ]
  assign(day, df)
}
```

### B and T annotation by TRUST4

```{r}
# Add the TRUST4 results into the seurat object
combined_vdj <- do.call(rbind, lapply(days, get))
combined_vdj <- combined_vdj[combined_vdj$Barcodes %in% colnames(data), ]
rownames(combined_vdj) <- combined_vdj$Barcodes
combined_vdj$Barcodes <- NULL
data <- AddMetaData(data, metadata = combined_vdj)
```

```{r}
plot<-DimPlot(data, reduction="umap", group.by = "Celltype", cols =c("#ea4258","darkgreen","orange"),na.value = "lightblue") + ggtitle("VDJ annotation") + theme(plot.title = element_text(size = 15, hjust = 0.5))
ggsave(paste0(params$output,"vdj_umap_",params$patient,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

Compare annotated B cells by TRUST4 and annotated tumoral cells by Numbat in T cells population

```{r}
data_T<-subset(data, subset= Annotation %in% c("T", "CD4 T", "CD8 T"))
# Compare B cells and tumoral cells from CNA analysis 
meta <- data_T@meta.data %>%
  as_tibble(rownames = "barcode") %>%
  select(barcode, Celltype, compartment_opt) %>%
  filter(
    !is.na(Celltype),
    !is.na(compartment_opt))

# Celltype × compartment_opt
ctab <- meta %>% count(Celltype, compartment_opt, name = "n_cells") %>% arrange(desc(n_cells))
ctab_wide <- ctab %>% pivot_wider(names_from = compartment_opt, values_from = n_cells, values_fill = 0)

# Plots
ctab_long <- ctab_wide %>%
  pivot_longer(!Celltype, names_to = "compartment_opt", values_to = "n_cells")

cols_fun <- colorRampPalette(c("white", "red"))

p_heatmap <- ggplot(ctab_long, aes(
  x = compartment_opt,
  y = Celltype,
  fill = n_cells)) +
  geom_tile(color = "gray95", size = 0.2) +
  scale_fill_gradientn(
    colors = cols_fun(100),
    na.value = "white") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 20),axis.text.y = element_text(size = 20),plot.title = element_text(size = 20, hjust = 0.5), legend.position = "right", legend.text = element_text(size = 20)) +labs(title = "Cell type V(D)J vs tumor annotation CNA",x = "Tumor annotation (CNA)",y = "Lymphoid annotation V(D)J",fill = "Cell number")

p_bar <- ggplot(ctab, aes(x = compartment_opt,y = n_cells,fill = Celltype)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 20),axis.text.y = element_text(size = 20),legend.text = element_text(size = 20), plot.title = element_text(size = 20, hjust = 0.5)) +
  labs(
    title = "Stacked barplot cell type V(D)J per tumor annotation CNA",
    x = "compartment_opt",
    y = "Cell fraction",
    fill = "Cell type")

ggsave(paste0(params$output,"vdj_cna_heatmap_",params$patient,".jpeg"), p_heatmap, width = 3000, height = 2000, units = "px")
ggsave(paste0(params$output,"vdj_cna_bar_",params$patient,".jpeg"), p_bar, width = 3000, height = 2000, units = "px")

plot<-ggplot(meta,
       aes(axis1 = compartment_opt,
           axis2 = Celltype)) +
  geom_alluvium(aes(fill = compartment_opt)) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size=6)+ labs(fill = "CNA annotation",title = "CNA and V(D)J annotation in the T population") +
  theme_minimal()+ theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 20),axis.text.y = element_text(size = 20),legend.text = element_text(size = 20), plot.title = element_text(size = 20, hjust = 0.5), legend.title = element_text(size = 20),)
ggsave(paste0(params$output,"vdj_cna_flux_",params$patient,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

Compare annotated B cells by TRUST4 and annotated tumoral cells by Numbat in monocyte population

```{r}
data_myel<-subset(data, subset= Annotation %in% c("Monocyte"))
# Compare B cells and tumoral cells from CNA analysis 
meta <- data_myel@meta.data %>%
  as_tibble(rownames = "barcode") %>%
  select(barcode, Celltype, compartment_opt) %>%
  filter(
    !is.na(Celltype),
    !is.na(compartment_opt))

# Celltype × compartment_opt
ctab <- meta %>% count(Celltype, compartment_opt, name = "n_cells") %>% arrange(desc(n_cells))
ctab_wide <- ctab %>% pivot_wider(names_from = compartment_opt, values_from = n_cells, values_fill = 0)

# Plots
ctab_long <- ctab_wide %>%
  pivot_longer(!Celltype, names_to = "compartment_opt", values_to = "n_cells")

cols_fun <- colorRampPalette(c("white", "red"))

p_heatmap <- ggplot(ctab_long, aes(
  x = compartment_opt,
  y = Celltype,
  fill = n_cells)) +
  geom_tile(color = "gray95", size = 0.2) +
  scale_fill_gradientn(
    colors = cols_fun(100),
    na.value = "white") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 20),axis.text.y = element_text(size = 20),plot.title = element_text(size = 20, hjust = 0.5), legend.position = "right",legend.text = element_text(size = 20)) +labs(title = "Cell type V(D)J vs tumor annotation CNA",x = "Tumor annotation (CNA)",y = "Lymphoid annotation V(D)J",fill = "Cell number")

p_bar <- ggplot(ctab, aes(x = compartment_opt,y = n_cells,fill = Celltype)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 20),axis.text.y = element_text(size = 20),plot.title = element_text(size = 20, hjust = 0.5), legend.position = "right",legend.text = element_text(size = 20)) +
  labs(
    title = "Stacked barplot cell type V(D)J per tumor annotation CNA",
    x = "compartment_opt",
    y = "Cell fraction",
    fill = "Cell type")


ggsave(paste0(params$output,"vdj_cna_heatmap_myel_",params$patient,".jpeg"), p_heatmap, width = 3000, height = 2000, units = "px")
ggsave(paste0(params$output,"vdj_cna_bar_myel_",params$patient,".jpeg"), p_bar, width = 3000, height = 2000, units = "px")

plot<-ggplot(meta,
       aes(axis1 = compartment_opt,
           axis2 = Celltype)) +
  geom_alluvium(aes(fill = compartment_opt)) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size=6)+ labs(fill = "CNA annotation",title = "CNA and V(D)J annotation in the monocyte population") +
  theme_minimal()+ theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 20),axis.text.y = element_text(size = 20),legend.text = element_text(size = 20), plot.title = element_text(size = 20, hjust = 0.5), legend.title = element_text(size = 20),)
ggsave(paste0(params$output,"vdj_cna_flux_myel_",params$patient,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

### Clonotypes and SHM (Somatic hypermutation rate of BCR)

```{r}
# Filter the dataframe TRUST4 output
combined_vdj <- do.call(rbind, lapply(days_airr, get))
combined_vdj <-combined_vdj[, -c(1,2,3,4, 10, 11, 12, 13, 14, 16, 17, 18, 19, 21, 22, 23)]
  
# Separate the IGH et IGK/L
heavy <- subset(combined_vdj, locus == "IGH")
light <- subset(combined_vdj, locus %in% c("IGK","IGL"))

# 1 chain per cell
heavy <- heavy %>%
group_by(Barcodes) %>%
slice_max(consensus_count, n=1)

light <- light %>%
group_by(Barcodes) %>%
slice_max(consensus_count, n=1)

# Combine IGH et IGK/L 
bcr <- left_join(light, heavy, by="Barcodes", suffix=c("_L","_H"))
```

Define clonotypes

junction = CDR3 (most variable zone)

Different clonotypes definition :

-   basic : junction on the heavy chain (IGH) or on the light chain (IGK/L)

-   more robust : junction on the heavy chain and light chain (IGK/L)

-   robust : variable zone + junction zone on the heavy and light chain

```{r}
# more light chain than heavy chain in the output 
# Add the TRUST4 results into the seurat object
bcr <-bcr[, -c(3, 5, 8, 12, 14, 17)]
bcr<-data.frame(bcr)
rownames(bcr) <- bcr$Barcodes
bcr$Barcodes <- NULL
data <- AddMetaData(data, metadata = bcr)
```

```{r}
plot<-DimPlot(data, reduction="umap", group.by = "j_call_L",na.value = "lightblue") + ggtitle("Germinal gene junction on light chain") + theme(plot.title = element_text(size = 15, hjust = 0.5))
ggsave(paste0(params$output,"vdj_clono_umap_light_",params$patient,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

Combine Variable (v) and junction (j) on light chain

```{r}
data$v_j_Light<-paste(data$v_call_L, data$j_call_L, sep="_")

counts <- table(data@meta.data$v_j_Light)
keep_annotations <- names(counts[counts > 100])

data@meta.data$v_j_Light_filtered <- ifelse(
  data@meta.data$v_j_Light %in% keep_annotations,
  data@meta.data$v_j_Light,
  "Others (<100 cells)")


DimPlot(data, reduction = "umap", group.by = "v_j_Light_filtered",cols=c("red", "orange", "darkgreen", "blue","lightblue", "grey", "grey"), na.value = "lightblue") + ggtitle("Clonotypes VJ Light chain (>100 cells)") + theme(plot.title = element_text(size = 15, hjust = 0.5))
```

Combine junction (j) on light chain and heavy chain

```{r}
data$j_j_light_heavy<-paste(data$j_call_L, data$j_call_H, sep="_")

counts <- table(data@meta.data$j_j_light_heavy)
keep_annotations <- names(counts[counts > 100])

data@meta.data$j_j_light_heavy_filtered <- ifelse(
  data@meta.data$j_j_light_heavy %in% keep_annotations,
  data@meta.data$j_j_light_heavy,
  "Others (<100 cells)")


DimPlot(data, reduction = "umap", group.by = "j_j_light_heavy_filtered",cols=c("red", "orange", "darkgreen", "blue","lightblue","yellow", "purple", "grey", "grey"), na.value = "lightblue") + ggtitle("Clonotypes VJ Light chain (>100 cells)") + theme(plot.title = element_text(size = 15, hjust = 0.5))
```

CDR3 junction_aa

```{r}
meta <- data@meta.data

# size for each clone
clone_sizes <- meta %>%
  filter(!is.na(junction_aa_L)) %>%
  count(junction_aa_L, name = "n_cells")

# clones classes
clone_sizes <- clone_sizes %>%
  mutate(
    class = case_when(
      n_cells == 1 ~ "1",
      n_cells <= 5 ~ "2-10",
      n_cells <= 10 ~ "11-30",
      n_cells <= 50 ~ "31-50",
      n_cells <= 100 ~ "51-100",
      n_cells <= 500 ~ "101-500",
      TRUE ~ "500+"))

# clones per classe
plot_df <- clone_sizes %>%count(class)
plot_df$class <- factor(plot_df$class,levels = c("1","2-10","11-30","31-50","51-100","101-500","500+"))

# plot
plot<-ggplot(plot_df, aes(x = class, y = n)) +geom_col(fill = "steelblue", width = 0.8) + labs( title = "Distribution clonotype size",x = "Number of cells carrying the clone", y = "Number of clones") + theme_classic(base_size = 16)
plot
ggsave(paste0(params$output,"vdj_clone_distribution_",params$patient,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

```{r}
tab <- table(data$junction_aa_L)
if (params$patient == "p2"){
  keep <- names(tab[tab > 30])
}
if (params$patient == "p3"){
  keep <- names(tab[tab > 10])
}
data_filt <- subset(data, junction_aa_L %in% keep)
tab<-table(data_filt$junction_aa_L)
cdr3 <- names(tab)

data$cdr3_L <- ifelse(
  data$junction_aa_L %in% cdr3,
  data$junction_aa_L,
  "Others"
)

if (params$patient == "p2"){
  plot<-DimPlot(data, reduction = "umap", group.by = "cdr3_L",cols=c("black","#12A8F5", "#E58B00", "#6A0DAD","#006400", "grey"), na.value = "lightblue") + ggtitle("Clonotypes CDR3 (main CDR3) on light chain") + theme(plot.title = element_text(size = 15, hjust = 0.5))
}
if (params$patient == "p3"){
  plot<-DimPlot(data, reduction = "umap", group.by = "cdr3_L",cols=c("#006400","#12A8F5", "grey","#E58B00"), na.value = "lightblue") + ggtitle("Clonotypes CDR3 (main CDR3) on light chain") + theme(plot.title = element_text(size = 15, hjust = 0.5))
}
ggsave(paste0(params$output,"vdj_clono_cdr3_umap_light_",params$patient,".jpeg"), plot, width = 3000, height = 2000, units = "px")

clone_matrix <- table(data$Annotation, 
                     ifelse(data$junction_aa_L %in% cdr3, data$junction_aa_L, "Other"))
mat <- clone_matrix[, colnames(clone_matrix) != "Other"]
mat_log <- log1p(mat)
mat_log <- mat_log[apply(mat_log, 1, var) > 0, ]
plot<-pheatmap(mat_log,scale = "row",main = "Main CDR3 per clusters",angle_col = 45,fontsize = 18,fontsize_row = 16,fontsize_col = 16)
ggsave(paste0(params$output,"vdj_clono_cdr3_heatmap_light_",params$patient,".jpeg"), plot, width = 4000, height = 2500, units = "px")
```

Subset myeloid and B pop

```{r}
# On the subset myeloid and B populations 
data_myel<-subset(data, subset = (Annotation %in% c("B intermediate", "B naive", "B memory", "B", "Bridge","Monocyte","Macrophage")))
data_myel<-run_dim_red(data_myel)

plot<-DimPlot(data_myel, reduction="umap", group.by = "j_call_L",na.value = "lightblue") + ggtitle("Germinal gene junction on light chain on B and myeloid population") + theme(plot.title = element_text(size = 15, hjust = 0.5))
ggsave(paste0(params$output,"vdj_clono_umap_light_myel_",params$patient,".jpeg"), plot, width = 3000, height = 2000, units = "px")

if (params$patient == "p2"){
  plot<-DimPlot(data_myel, reduction = "umap", group.by = "cdr3_L",cols=c("black","#12A8F5", "#E58B00", "#6A0DAD","#006400", "grey"), na.value = "lightblue") + ggtitle("Clonotypes CDR3 on light chain on B and myeloid population") + theme(plot.title = element_text(size = 15, hjust = 0.5))
}
if (params$patient == "p3"){
  plot<-DimPlot(data_myel, reduction = "umap", group.by = "cdr3_L",cols=c("#006400","#12A8F5", "grey","#E58B00"), na.value = "lightblue") + ggtitle("Clonotypes CDR3 on light chain on B and myeloid population") + theme(plot.title = element_text(size = 15, hjust = 0.5))
}
ggsave(paste0(params$output,"vdj_clono_cdr3_umap_light_myel_",params$patient,".jpeg"), plot, width = 3000, height = 2000, units = "px")

clone_matrix <- table(data_myel$Annotation, 
                     ifelse(data_myel$junction_aa_L %in% cdr3, data_myel$junction_aa_L, "Other"))
mat <- clone_matrix[, colnames(clone_matrix) != "Other"]
mat_log <- log1p(mat)
mat_log <- mat_log[apply(mat_log, 1, var) > 0, ]
plot<-pheatmap(mat_log,scale = "row",main = "Main CDR3 per clusters on B and myeloid population",angle_col = 45,fontsize = 18,fontsize_row = 16,fontsize_col = 16)
ggsave(paste0(params$output,"vdj_clono_cdr3_heatmap_myel_",params$patient,".jpeg"), plot, width = 4000, height = 2500, units = "px")
```

```{r}
if (params$patient == "p2"){
  clono_color=c("CAIWYSSTSHSDTHTW"="black","CMQALQTL"="#12A8F5", "CQQYNSYWTF"="#E58B00", "CQSADSSGTYVVF"="#6A0DAD","CSAWDTSLGVWVF"="#006400", "Others"="grey")
}
if (params$patient == "p3"){
  clono_color=c("CQQYYSYPRTF"="#12A8F5", "SYPRTF"="#E58B00","CQQDYSYPRTF"="#006400", "Others"="grey")
}

data<-FindClusters(data, resolution = 0.15)
DimPlot(data, reduction="umap")
data_myel<-FindClusters(data_myel, resolution = 0.15)
plot<-DimPlot(data_myel, reduction="umap",label=T, label.size = 6 )
ggsave(paste0(params$output,"umap_myel_clusters_",params$patient,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

Clonotype x time x celltype

```{r}
meta <- data_myel@meta.data

# Alluvial Plot
df_alluv <- meta %>%
  mutate(
    clonotype = as.character(cdr3_L),
    cluster   = as.character(seurat_clusters),
    timepoint = as.character(timepoint)
  ) %>%
  filter(
    !is.na(clonotype),
    !is.na(cluster),
    !is.na(timepoint),
    clonotype != "Others",
    clonotype != ""
  ) %>%
  count(clonotype, cluster, timepoint, name = "n")

df_alluv$timepoint <- factor(
  df_alluv$timepoint,
  levels = c("D1", "D4", "D8", "D11", "D14"))

plot <- ggplot(df_alluv,
       aes(axis1 = clonotype,
           axis2 = cluster,
           axis3 = timepoint,
           y = n)) +
  geom_alluvium(aes(fill = clonotype), width = 1/12) +
  geom_stratum(width = 1/8, fill = "grey85") +
  geom_text(stat = "stratum",
            aes(label = after_stat(stratum)),
            size = 5,
            check_overlap = FALSE) +
  scale_x_discrete(limits = c("Clonotype", "Cluster", "Timepoint"),
                   expand = c(0.3, 0.05)) +
  theme_minimal(base_size = 16) +
  theme(plot.margin = margin(t=10, r=10, b=10, l=60)) +
  ggtitle("Clone per cluster and per timepoint in myeloid and B population ")
plot
ggsave(paste0(params$output,"vdj_flow_myel_",params$patient,".jpeg"), plot, width = 3500, height = 2000, units = "px")

# Proportion
meta <- meta %>%
  mutate(
    clonotype = as.character(cdr3_L),
    timepoint = as.character(timepoint),
    cluster = as.character(seurat_clusters),
    celltype = as.character(Annotation))

prop <-meta %>%
  filter(!celltype %in% c("Monocyte", "B memory")) %>%
  mutate(timepoint = factor(timepoint, levels = c("D1", "D4", "D8", "D11", "D14")))

if (params$patient == "p3"){
  prop$clonotype <- factor(prop$clonotype,levels = c("CQQDYSYPRTF", "CQQYYSYPRTF", "SYPRTF", "Others"))
}
plot<-ggplot(prop, aes(x = timepoint, fill = clonotype)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::number) +
  scale_fill_manual(values = clono_color) +
  facet_wrap(~ celltype, nrow = 1) +
  theme_minimal(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y = "Fraction", fill = "Clonotype",
       title = "Clonal composition over time and cell type")
plot
ggsave(paste0(params$output,"vdj_prop_cell_time_myel_",params$patient,".jpeg"), plot, width = 3500, height = 1500, units = "px")
```

Fisher's test : Enrichment clonotype and clusters

```{r}
res <- list()

enrich <-meta %>%
  filter(!is.na(clonotype),clonotype != "Others")

for(cl in unique(enrich$clonotype)){
  for(clus in unique(enrich$cluster)){

    a <- sum(meta$clonotype == cl & meta$cluster == clus)
    b <- sum(meta$clonotype == cl & meta$cluster != clus)
    c <- sum(meta$clonotype != cl & meta$cluster == clus)
    d <- sum(meta$clonotype != cl & meta$cluster != clus)

    mat <- matrix(c(a,b,c,d), nrow=2)

    ft <- fisher.test(mat)

    res[[paste(cl,clus,sep="_")]] <- data.frame(
      clonotype = cl,
      cluster = clus,
      odds_ratio = ft$estimate,
      pvalue = ft$p.value)
  }
}

enrich <- bind_rows(res) %>%
  mutate(FDR = p.adjust(pvalue, method="BH")) %>%
  arrange(FDR)

enrich <- enrich %>%
  mutate(
    odds_ratio = ifelse(is.infinite(odds_ratio), NA, odds_ratio),
    odds_ratio = ifelse(odds_ratio <= 0, NA, odds_ratio),
    FDR = ifelse(FDR == 0, 1e-300, FDR))

mat <- enrich %>%
  mutate(
    score = sign(log2(odds_ratio)) * -log10(FDR),
    cluster = factor(cluster, levels = sort(unique(cluster)))
  ) %>%
  select(clonotype, cluster, score) %>%
  pivot_wider(
    names_from = cluster,
    values_from = score,
    values_fill = 0
  ) %>%
  column_to_rownames("clonotype") %>%
  as.matrix()


plot<-pheatmap(mat,main = "Fisher's test : clone-cluster enrichment",fontsize = 14,fontsize_row = 12, fontsize_col = 16, angle_col = 45, cluster_cols = TRUE, cluster_rows = TRUE, show_rownames = TRUE, show_colnames = TRUE)
ggsave(paste0(params$output,"vdj_heatmap_enrich_myel_",params$patient,".jpeg"), plot, width = 3000, height = 1500, units = "px")
```

Somatic Hypermutation (SHM)

SHM on the light chain (IGK/L). Here the SHM is calculated on the variable zone (V identity)

Variable zone is longer than junction zone, that's why it's better to calculate SHM on this zone

When SHM is high ( for example 100%) it means that the cells are naive, because no or not a lot of mutation. When SHM is lower it means that the cell has more mutation so more mature and old.

```{r}
meta$SHM_L<-100-meta$v_identity_L
df_plot <- meta %>%
  filter(Annotation != "Monocyte") %>%
  filter(cdr3_L != "Others") %>%
  mutate(Annotation_merged = case_when(
    Annotation %in% c("B", "B naive", "B intermediate", "B memory") ~ "B cells",
    TRUE ~ as.character(Annotation) ))

plot<-ggplot(df_plot, aes(x = cdr3_L, y = SHM_L, fill = Annotation_merged)) +
  geom_boxplot(outlier.shape = NA, width = 0.8) +
  theme_minimal() +
  #ylim(90, 100) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12), axis.text.y = element_text(size = 12) , axis.title=element_text(size = 12), legend.text=element_text(size = 12), legend.title=element_text(size = 12)) + 
  labs(title = "Somatic Hypermutation (SHM) per CDR3 IGK/L and per cell type",
       y = "Somatic hypermutation (%)",
       x = "CDR3 Sequence", fill = "Cell type")
plot
ggsave(paste0(params$output,"vdj_box_celltype_",params$patient,".jpeg"), plot, width = 3000, height = 1500, units = "px")

my_comparisons <- combn(unique(df_plot$cdr3_L), 2, simplify = FALSE)
plot<-ggplot(df_plot, aes(x = cdr3_L, y = SHM_L)) +
  geom_boxplot(outlier.shape = NA, width = 0.8) +
  theme_minimal() +
  #ylim(90, 110) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12), axis.text.y = element_text(size = 12) , axis.title=element_text(size = 12), plot.title=element_text(size = 18)) +
  labs(title = "Somatic Hypermutation (SHM) per CDR3",
       y = "Somatic hypermutation (%)",
       x = "CDR3 Sequence") +
 # global test : kruskal and wallis 
  stat_compare_means(method = "kruskal.test",label.y = 20) +
  # pairwise tests
  stat_compare_means(comparisons = my_comparisons, method = "wilcox.test",method.args = list(exact = FALSE),p.adjust.method = "BH",hide.ns = FALSE,label = "p.signif",step.increase = 0.04)
ggsave(paste0(params$output,"vdj_box_clone_",params$patient,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

```{r}
# Cells and sequences selection
df_tree <- data_myel@meta.data %>% 
  filter(!is.na(junction_aa_L)) %>% 
  group_by(junction_aa_L) %>% 
  slice(1) %>%
  ungroup()

# Distance matrix (Hamming on CDR3)
seqs <- df_tree$junction_aa_L
names(seqs) <- df_tree$junction_aa_L
dist_mat <- stringdistmatrix(seqs, seqs, method = "lv")
rownames(dist_mat) <- names(seqs)

# Build the tree
tree <- nj(as.dist(dist_mat))

# Plot
plot<-ggtree(tree, layout="circular") + geom_tiplab(size=2) + theme_tree2() + labs(title = "Clonotypes Phylogeny (Levenshtein distance)")
plot
ggsave(paste0(params$output,"vdj_tree_",params$patient,".jpeg"), plot, width = 3000, height = 2000, units = "px")
```

```{r}
tab <- table(data_myel$junction_aa_L)
keep <- names(tab[tab > 2])
data_filt <- subset(data_myel, junction_aa_L %in% keep)
tab<-table(data_filt$junction_aa_L)
cdr3 <- names(tab)

data_myel$cdr3_low_L <- ifelse(
  data_myel$junction_aa_L %in% cdr3,
  data_myel$junction_aa_L,
  "Others"
)

# Cells and sequences selection
df_tree <- data_myel@meta.data %>%
  filter(
    !is.na(cdr3_low_L),
    cdr3_low_L != "Others",
    !is.na(junction_aa_L),
    junction_aa_L != ""
  ) %>%
  group_by(cdr3_low_L) %>%
  slice(1) %>%
  ungroup()

# Distance matrix (Hamming on CDR3)
seqs <- df_tree$junction_aa_L
names(seqs) <- df_tree$junction_aa_L
dist_mat <- stringdistmatrix(seqs, seqs, method = "lv")
rownames(dist_mat) <- names(seqs)

# Build the tree
tree <- nj(as.dist(dist_mat))

# Plot
clones<-unique(data$cdr3_L)
tip_df <- data.frame(label = tree$tip.label) %>%
  mutate(highlight = ifelse(label %in% clones, "Dominant clones", "Non-dominant clones") )

plot<-ggtree(tree, layout = "dendrogram") %<+% tip_df +
  geom_tiplab(aes(color = highlight),size = 5, hjust = 0) +
  scale_color_manual(name = "Dominance",values = c("Dominant clones" = "red","Non-dominant clones" = "black")) + xlim_tree(8) +
  theme_tree() +theme(
    plot.margin = margin(20, 100, 80, 20),
    plot.clip = "off",
    legend.position = "right",
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold")) +
  labs(title = "Clonotypes Phylogeny (Levenshtein distance)")
plot
ggsave(paste0(params$output,"vdj_tree_small_",params$patient,".jpeg"), plot, width = 4000, height = 4000, units = "px")
```
