---
title: "GRN"
params: 
  file_p2: "/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/CLL_Seurat_P2.rds"
  file_p3: "/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/CLL_Seurat_P3.rds"
  file_step2_p2: "/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/step_2_reg_p2.csv"
  file_step3_p2: "/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/step_3_aucell_p2.csv"
  file_step2_p3: "/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/step_2_reg_p3.csv"
  file_step3_p3: "/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/step_3_aucell_p3.csv"
format: html
editor: visual
---

# Data and libraries initialization

```{r}
#| message: false
library(SeuratObject)
library(SeuratDisk)
library(sp)
library(Seurat)
library(hdf5r)
library(dplyr)
library(patchwork)
library(ggplot2)
library(stringr)
library(pheatmap)
library(viridis)
library(gridExtra)
library(SeuratExtend)
library(AUCell)
library(clusterProfiler)
library(org.Hs.eg.db)
library(SCENIC)
library(ComplexHeatmap)
library(decoupleR)
library(forcats) 
library(OmnipathR)
source("functions.R")
```

Read the data pre-processed

```{r}
data_p2<-readRDS(params$file_p2)
Idents(data_p2) <- data_p2$final_annot
data_p2<-subset(data_p2, subset = (final_annot %in% c("B","B intermediate","B memory","B naive","Macrophage")))

data_p3<-readRDS(params$file_p3)
Idents(data_p3) <- data_p3$final_annot
data_p3<-subset(data_p3, subset = (final_annot %in% c("B","B intermediate","B memory","B naive","Macrophage")))
```

```{r}
#SaveLoom(data, filename="TimeCourse.loom")
```

# SCENIC

### Apply pySCENIC in the Genotoul cluster with the script "pySCENIC.sh"

Import pySCENIC output

```{r}
# TF and targets
regulons_p2 <- read.csv(params$file_step2_p2)
regulons_p3 <- read.csv(params$file_step2_p3)

# Loom output in seurat object
#scenic_output<-"/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/out_TimeCourse.loom"
#data <- ImportPyscenicLoom(scenic_output, seu = data)
```

Bridge annotation P2

```{r}
umap_coords <- Embeddings(data_p2, "umap")
umap_df <- as.data.frame(umap_coords)
colnames(umap_df)
umap_df$cell <- rownames(umap_df)
colnames(umap_df)

x_min <- 2.5
x_max <- 6
y_min <- 4.5
y_max <- 7

umap_df$selected <- ifelse(
  umap_df$umap_1 >= x_min & umap_df$umap_1 <= x_max &
  umap_df$umap_2 >= y_min & umap_df$umap_2 <= y_max,
  "Intermediate population",  
  NA      
)

# Add "bridge" on final annotation
umap_df <- umap_df %>%
  mutate(
    bridge_annot = ifelse(is.na(selected), data_p2$final_annot[cell], selected)
  )

data_p2$bridge_annot <- umap_df$bridge_annot[match(colnames(data_p2), umap_df$cell)]

table(data_p2$bridge_annot)

# Plot UMAP
umap_df$bridge_annot <- data_p2$bridge_annot[match(umap_df$cell, colnames(data_p2))]

plotumap<-ggplot(umap_df, aes(x = umap_1, y = umap_2, color = bridge_annot)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("Intermediate population" = "red", "B" = "grey", "Macrophages" = "grey", "B naive" = "grey", "B memory" = "grey", "B intermediate" = "grey")) +
  theme_minimal() +
  ggtitle("Bridge B cells and macrophages")
plotumap
```

P3

```{r}
umap_coords <- Embeddings(data_p3, "umap")
umap_df <- as.data.frame(umap_coords)
colnames(umap_df)
umap_df$cell <- rownames(umap_df)
colnames(umap_df)

x_min <- 3.5
x_max <- 7.5
y_min <- -2.5
y_max <- 0

umap_df$selected <- ifelse(
  umap_df$umap_1 >= x_min & umap_df$umap_1 <= x_max &
  umap_df$umap_2 >= y_min & umap_df$umap_2 <= y_max,
  "Intermediate population",  
  NA      
)

# Add "bridge" or final annotation
umap_df <- umap_df %>%
  mutate(
    bridge_annot = ifelse(is.na(selected), data_p3$final_annot[cell], selected)
  )

data_p3$bridge_annot <- umap_df$bridge_annot[match(colnames(data_p3), umap_df$cell)]

table(data_p3$bridge_annot)

# Plot UMAP
umap_df$bridge_annot <- data_p3$bridge_annot[match(umap_df$cell, colnames(data_p3))]

plotumap<-ggplot(umap_df, aes(x = umap_1, y = umap_2, color = bridge_annot)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("Intermediate population" = "red", "B" = "grey", "Macrophages" = "grey", "B naive" = "grey", "B memory" = "grey", "B intermediate" = "grey")) +
  theme_minimal() +
  ggtitle("Bridge B cells and macrophages")
plotumap
```

## Patient 2 analysis

```{r}
tf_auc <- read.csv(params$file_step3_p2, row.names = 1, check.names=FALSE)
colnames(tf_auc) <- gsub("[(+)]", "", colnames(tf_auc))
celltypes<-data_p2@meta.data[["bridge_annot"]]
celltypes<-ifelse(celltypes=="B intermediate"|celltypes=="B naive"|celltypes=="B memory"|celltypes=="B","B cells", celltypes)
tf_zscore <- CalcStats(tf_auc, f = celltypes, order = "p", n = 15, t = TRUE) # change the n (n*cluster=TF number)
plot<-Heatmap(tf_zscore, lab_fill = "zscore")+ ggtitle("Transcription factor activity (z-score) across B cells,\n intermediate population, and macrophages, P2")
plot
ggsave("Plots/heatmap_tf_P2.jpeg", plot, width = 1500, height = 2000, units = "px")
```

```{r}
celltypes<-data_p2@meta.data[["bridge_annot"]]
tf_zscore <- CalcStats(tf_auc, f = celltypes, order = "p", n = 6, t = TRUE) # change the n (n*cluster=TF number)
plot<-Heatmap(tf_zscore, lab_fill = "zscore")
plot
```

Waterfall plot (CLL-macro vs B, CLL-macro vs macro)

```{r}
plot<-WaterfallPlot(t(tf_auc), f = celltypes,
  ident.1 = "Intermediate population",      # First group of cells
  ident.2 = "Macrophage",     # Second group of cells
  exp.transform = FALSE,      # Disable transformation of expression data
  top.n = 20,                  # Display the top 20 most differentially active TFs
  length="logFC", 
  log.base="2", 
  color="p"
) + ggtitle("Differential transcription factor activity: \nintermediate population vs macrophages, P2")+theme(axis.text.y = element_text(size = 12))
plot
ggsave("Plots/waterfallplot_macroxbridge_p2.jpeg", plot, width = 2000, height = 2000, units = "px")
```

```{r}
celltypes<-ifelse(celltypes=="B intermediate"|celltypes=="B naive"|celltypes=="B memory"|celltypes=="B","B cells", celltypes)
plot<-WaterfallPlot(t(tf_auc), f = celltypes,
  ident.1 = "Intermediate population",      # First group of cells
  ident.2 = "B cells",     # Second group of cells
  exp.transform = FALSE,      # Disable transformation of expression data
  top.n = 20,                  # Display the top 20 most differentially active TFs
  length="logFC", 
  log.base="2", 
  color="p"
) + ggtitle("Differential transcription factor activity: \nintermediate population vs B cells, P2")+theme(axis.text.y = element_text(size = 12))
plot
ggsave("Plots/waterfallplot_Bxbridge_p2.jpeg", plot, width = 2000, height = 2000, units = "px")
```

TF Regulon Activity (AUCell)

```{r}
# AUC score dataframe
auc_mtx <- read.csv(params$file_step3_p2, row.names = 1, check.names=FALSE)
auc_mtx <- t(auc_mtx)
auc_mtx <- as.data.frame(auc_mtx)
rownames(auc_mtx) <- gsub("[(+)]", "", rownames(auc_mtx))
auc_mtx[] <- lapply(auc_mtx, as.numeric)
auc_mtx <- as.matrix(auc_mtx)

data_p2[['TF']] <- CreateAssayObject(data = auc_mtx)
DefaultAssay(data_p2) <- 'RNA'

plot<-DimPlot2(
  data_p2,
  features = c("tf_CEBPA", "tf_NFIX"),
  ncol = 2,
  theme = NoAxes(), 
  reduction = "umap",
  label.size=2
)
plot
ggsave("Plots/tf_activity_P2.jpeg", plot, width = 2000, height = 1000, units = "px")
```

Heatmap

```{r}
groups <- data_p2@meta.data[["bridge_annot"]]
identical(colnames(auc_mtx), colnames(data_p2))
annotation_df <- data.frame(Cell_types = data_p2@meta.data[["bridge_annot"]])
rownames(annotation_df) <- colnames(auc_mtx)

# Keep the most variable TFs
var_auc <- apply(auc_mtx, 1, var)
top_var_regulons <- names(sort(var_auc, decreasing = TRUE))[1:20]
auc_mat_top <- auc_mtx[top_var_regulons, ]

my_colors <- list(
  Cell_types = c(B = "darkgreen","B intermediate" = "darkgreen","B memory" = "darkgreen","B naive" = "darkgreen", "Macrophage"= "red", "Intermediate population"="darkblue")
)

# Scaled matrix 
auc_mat_scaled <- t(scale(t(auc_mat_top)))

# Heatmap
pheatmap(
  auc_mat_top,
  cluster_cols = TRUE,
  cluster_rows = TRUE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  annotation_col = annotation_df, 
  annotation_colors = my_colors)
```

Heatmap : TF most variable

```{r}
cell_types <- data_p2@meta.data[["bridge_annot"]]

# Mean per cell type
auc_mtx_by_type <- sapply(unique(cell_types), function(ct) {
  cells_in_type <- which(cell_types == ct)
  rowMeans(auc_mtx[, cells_in_type, drop = FALSE])
})
auc_mtx_by_type <- t(auc_mtx_by_type)

# TF most variable
var_auc <- apply(auc_mtx_by_type, 2, var)
top_var_regulons <- names(sort(var_auc, decreasing = TRUE))[1:30]

auc_mat_top <- auc_mtx_by_type[, top_var_regulons, drop = FALSE]

plot<-pheatmap(
  auc_mat_top,
  cluster_cols = TRUE,
  cluster_rows = TRUE,
  show_colnames = TRUE,
  show_rownames = TRUE,
  color = colorRampPalette(c("white", "blue", "darkblue"))(100), 
  angle_col = 45, 
  fontsize_row = 7,
  fontsize_col = 7, 
  fontsize = 5,
  main="TF most variable depending on cell type")
plot
ggsave("Plots/heatmap_tf_P2.jpeg", plot, width = 2000, height = 800, units = "px")
```

UMAP based on TF activity

```{r}
DefaultAssay(data_p2) <- 'TF'
all.genes <- rownames(data_p2)
data_p2 <- ScaleData(data_p2, features = all.genes)
data_p2<-RunPCA(data_p2, assay="TF", reduction.name = "pca_tf", features = all.genes)
data_p2 <- RunUMAP(data_p2, dims = 1:10,  reduction.name = "umap_tf")
data_p2 <- RunTSNE(data_p2, dims = 1:10,  reduction.name = "tsne_tf")

cell_palette <- c(
  "Intermediate population" = "#1f77b4",
  "B intermediate" = "#ff7f0e",
  "B naive" = "#ffd92f",
  "B memory" = "#582d15ff",
  "Macrophage" = "#984ea3",
  "B" = "#f2a890ff")
plot<-DimPlot(data_p2, reduction = "umap_tf", group.by = "bridge_annot", cols = cell_palette) + ggtitle("UMAP based on TF activity, P2")+ theme(plot.title = element_text(size = 10, hjust = 0.5), legend.text = element_text(size =8 ))
plot
ggsave("Plots/umap_tf_P2.jpeg", plot, width = 2000, height = 1000, units = "px")
DimPlot(data_p2, reduction = "tsne_tf", group.by = "bridge_annot", cols = cell_palette) + ggtitle("t-SNE based on TF activity, P2")
```

```{r}
TF_list<-tf_zscore[order(tf_zscore$`Intermediate population`, decreasing = T),]
TF_list<-head(TF_list, 15)
TF_list<-rownames(TF_list)

enrich_tf <- enrichGO(
  gene          = TF_list,
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05
)
plot<-barplot(enrich_tf, showCategory = 20, title = "Top 20 enriched GO terms")
ggsave("Plots/enrich_P2.jpeg", plot, width = 2000, height = 3000, units = "px")
cnetplot(enrich_tf, categorySize="pvalue", foldChange=NULL)  # si foldChange disponible
plot<-goplot(enrich_tf)
ggsave("Plots/enrich_goplot_P2.jpeg", plot, width = 4000, height = 3000, units = "px")
```

TF differential active

```{r}
Idents(data_p2) <- data_p2$bridge_annot

markers_TF <- FindMarkers(
  data_p2,
  assay = "TF",
  ident.1 = "Intermediate population", ident.2 = c("Macrophage","B", "B naive","B intermediate","B memory"),
  test.use = "wilcox",
  logfc.threshold = 0
)
markers_TF <- subset(markers_TF, p_val_adj < 0.05)
head(markers_TF[order(markers_TF$avg_log2FC, decreasing = TRUE), ])   # TF activés
head(markers_TF[order(markers_TF$avg_log2FC, decreasing = FALSE), ])  # TF réprimés

top_TF <- rownames(markers_TF[order(abs(markers_TF$avg_log2FC), decreasing = TRUE), ])[1:20]
DoHeatmap(data_p2, features = top_TF, assay = "TF")

markers_TF$TF <- rownames(markers_TF)
ggplot(markers_TF, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
  geom_point() +
  geom_vline(xintercept = c(-0.25, 0.25), linetype = "dashed") +
  theme_minimal() +
  labs(x = "Variation d’activité (log2FC)", y = "-log10(p_adj)", title = "TF activés ou réprimés")

regulons_p2<-regulons_p2$Enrichment.6
```

```{r}
DefaultAssay(data_p2) <- 'TF'
plot<-VlnPlot(data_p2, features = c("CEBPA"),split.by = "timepoint")
plot<-VlnPlot(data_p2, features = c("CEBPB"))
ggsave("PAX5.jpeg", plot, width = 3000, height = 2000, units = "px")
```

```{r}
selectedResolution <- "Annotation_B" # select resolution
# Split the cells by cluster:
cellClusters<-data_p2@meta.data %>%
  dplyr::select(Annotation) %>%
  dplyr::mutate(Annotation=dplyr::recode(Annotation, "Bridge" = "Intermediate population")) 

cellClusters$Annotation_B <- ifelse(cellClusters$Annotation %in% c("B memory", "B intermediate", "B naive", "B"),"B cells",cellClusters$Annotation)

cellsPerCluster <- split(rownames(cellClusters), cellClusters[,selectedResolution]) 
regulonAUC <- read.csv("/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/step_3_aucell_p2.csv", row.names = 1, check.names=FALSE)
regulonAUC <-t(regulonAUC)
regulonAUC <- regulonAUC[onlyNonDuplicatedExtended(rownames(regulonAUC)),]
# Calculate average expression:
regulonAUC_mat <- as.matrix(regulonAUC)

regulonActivity_byCellType <- sapply(
  cellsPerCluster,
  function(cells) {
    rowMeans(regulonAUC_mat[, cells, drop = FALSE])
  })

# Scale expression:
regulonActivity_byCellType_Scaled <- t(scale(t(regulonActivity_byCellType), center = T, scale=T))

# Plots
options(repr.plot.width=8, repr.plot.height=10) # To set the figure size in Jupyter
hm <- draw(ComplexHeatmap::Heatmap(regulonActivity_byCellType_Scaled, name="Regulon activity",
                       row_names_gp=grid::gpar(fontsize=5), show_row_names = FALSE, column_names_rot = 45)) # row font size


regulonOrder <- rownames(regulonActivity_byCellType_Scaled)[row_order(hm)] # to save the clustered regulons for later

tf_interest <- c("SPI1(+)", "CEBPD(+)", "CEBPA(+)","CEBPB(+)","ETV6(+)","NFIX(+)")

idx <- which(rownames(regulonActivity_byCellType_Scaled) %in% tf_interest)

ha <- rowAnnotation(
  mark = anno_mark(
    at = idx,
    labels = rownames(regulonActivity_byCellType_Scaled)[idx]
  )
)

Heatmap(
  regulonActivity_byCellType_Scaled,
  name = "Regulon activity",
  show_row_names = FALSE,
  right_annotation = ha,
  column_names_rot = 45
)
```

```{r}
rss <- calcRSS(AUC = regulonAUC,cellAnnotation = cellClusters[colnames(regulonAUC), selectedResolution])
## Showing regulons and cell types with any RSS > 0.01 
rssPlot <- plotRSS(rss)
plotly::ggplotly(rssPlot$plot)
options(repr.plot.width=5, repr.plot.height=5) # To set the figure size in Jupyter
plotRSS_oneSet(rss, setName = "Macrophage",n=10) # cluster ID
plotRSS_oneSet(rss, setName = "B cells",n=10) # cluster ID
par(
  cex.axis = 1.5,   # axes ticks
  cex.lab = 1.8,    # titres des axes
  cex.main = 2      # titre principal
)
plotRSS_oneSet(rss, setName = "Intermediate population",n=15) # cluster ID
```

## Patient 3 analysis

```{r}
tf_auc <- read.csv(params$file_step3_p3, row.names = 1, check.names=FALSE)
colnames(tf_auc) <- gsub("[(+)]", "", colnames(tf_auc))
celltypes<-data_p3@meta.data[["bridge_annot"]]
celltypes<-ifelse(celltypes=="B intermediate"|celltypes=="B naive"|celltypes=="B memory"|celltypes=="B","B cells", celltypes)
tf_zscore <- CalcStats(tf_auc, f = celltypes, order = "p", n = 15, t = TRUE) # change the n (n*cluster=TF number)
plot<-Heatmap(tf_zscore, lab_fill = "zscore")+ ggtitle("Transcription factor activity (z-score) across B cells,\n intermediate population, and macrophages, P3")
plot
ggsave("Plots/heatmap_tf_P3.jpeg", plot, width = 1500, height = 2000, units = "px")
```

```{r}
celltypes<-data_p3@meta.data[["bridge_annot"]]
tf_zscore <- CalcStats(tf_auc, f = celltypes, order = "p", n = 6, t = TRUE) # change the n (n*cluster=TF number)
plot<-Heatmap(tf_zscore, lab_fill = "zscore")
plot
```

Waterfall plot (CLL-macro vs B, CLL-macro vs macro)

```{r}
plot<-WaterfallPlot(t(tf_auc), f = celltypes,
  ident.1 = "Intermediate population",      # First group of cells
  ident.2 = "Macrophage",     # Second group of cells
  exp.transform = FALSE,      # Disable transformation of expression data
  top.n = 20,                  # Display the top 20 most differentially active TFs
  length="logFC", 
  log.base="2", 
  color="p"
) + ggtitle("Differential transcription factor activity: \nintermediate population vs macrophages, P3") +theme(axis.text.y = element_text(size = 12))
plot
ggsave("Plots/waterfallplot_macroxbridge_p3.jpeg", plot, width = 2000, height = 2000, units = "px")
```

```{r}
celltypes<-ifelse(celltypes=="B intermediate"|celltypes=="B naive"|celltypes=="B memory"|celltypes=="B","B cells", celltypes)
plot<-WaterfallPlot(t(tf_auc), f = celltypes,
  ident.1 = "Intermediate population",      # First group of cells
  ident.2 = "B cells",     # Second group of cells
  exp.transform = FALSE,      # Disable transformation of expression data
  top.n = 20,                  # Display the top 20 most differentially active TFs
  length="logFC", 
  log.base="2", 
  color="p"
) + ggtitle("Differential transcription factor activity: \nintermediate population vs B cells, P3")+theme(axis.text.y = element_text(size = 12))
plot
ggsave("Plots/waterfallplot_Bxbridge_p3.jpeg", plot, width = 2000, height = 2000, units = "px")
```

TF Regulon Activity (AUCell)

```{r}
# AUC score dataframe
auc_mtx <- read.csv(params$file_step3_p3, row.names = 1, check.names=FALSE)
auc_mtx <- t(auc_mtx)
auc_mtx <- as.data.frame(auc_mtx)
rownames(auc_mtx) <- gsub("[(+)]", "", rownames(auc_mtx))
auc_mtx[] <- lapply(auc_mtx, as.numeric)
auc_mtx <- as.matrix(auc_mtx)

data_p3[['TF']] <- CreateAssayObject(data = auc_mtx)
DefaultAssay(data_p3) <- 'RNA'

plot<-DimPlot2(
  data_p3,
  features = c("tf_CEBPD", "tf_NFIX"),
  ncol = 2,
  cols=list("tf_CEBPD" = "D", "tf_NFIX" = "D"),
  theme = NoAxes(), 
  reduction = "umap",
  label.size=2
)
plot
ggsave("Plots/tf_activity_P3.jpeg", plot, width = 2000, height = 1000, units = "px")
```

Heatmap

```{r}
groups <- data_p3@meta.data[["bridge_annot"]]
identical(colnames(auc_mtx), colnames(data_p3))
annotation_df <- data.frame(Cell_types = data_p3@meta.data[["bridge_annot"]])
rownames(annotation_df) <- colnames(auc_mtx)

# Keep the most variable TFs
var_auc <- apply(auc_mtx, 1, var)
top_var_regulons <- names(sort(var_auc, decreasing = TRUE))[1:20]
auc_mat_top <- auc_mtx[top_var_regulons, ]

my_colors <- list(
  Cell_types = c(B = "darkgreen","B intermediate" = "darkgreen","B memory" = "darkgreen","B naive" = "darkgreen", "CD4 T" = "lightblue", "CD8 T" = "lightblue", "T" = "lightblue", "Macrophage"= "red","Monocyte"="orange", "Neutrophil"="brown", "NK"="darkblue", "Erythroblast"="grey" )
)

# Scaled matrix 
auc_mat_scaled <- t(scale(t(auc_mat_top)))

# Heatmap
pheatmap(
  auc_mat_top,
  cluster_cols = TRUE,
  cluster_rows = TRUE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  annotation_col = annotation_df, 
  annotation_colors = my_colors)
```

```{r}
cell_types <- data_p3@meta.data[["bridge_annot"]]

# Mean per cell type
auc_mtx_by_type <- sapply(unique(cell_types), function(ct) {
  cells_in_type <- which(cell_types == ct)
  rowMeans(auc_mtx[, cells_in_type, drop = FALSE])
})
auc_mtx_by_type <- t(auc_mtx_by_type)

# TF most variable
var_auc <- apply(auc_mtx_by_type, 2, var)
top_var_regulons <- names(sort(var_auc, decreasing = TRUE))[1:30]

auc_mat_top <- auc_mtx_by_type[, top_var_regulons, drop = FALSE]

plot<-pheatmap::pheatmap(
  auc_mat_top,
  cluster_cols = TRUE,
  cluster_rows = TRUE,
  show_colnames = TRUE,
  show_rownames = TRUE,
  color = colorRampPalette(c("white", "blue", "darkblue"))(100), 
  angle_col = "45", 
  fontsize_row = 7,
  fontsize_col = 7, 
  fontsize = 5,
  main="TF most variable depending on cell type")
plot
ggsave("Plots/heatmap_tf_P3.jpeg", plot, width = 2000, height = 800, units = "px")
```

```{r}
TF_list<-tf_zscore[order(tf_zscore$`Intermediate population`, decreasing = T),]
TF_list<-head(TF_list, 15)
TF_list<-rownames(TF_list)

enrich_tf <- enrichGO(
  gene          = TF_list,
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05
)
barplot(enrich_tf, showCategory = 20, title = "Top 20 enriched GO terms")
dotplot(enrich_tf, showCategory = 20, title = "Top 20 enriched GO terms")
emapplot(enrich_tf)
cnetplot(enrich_tf, categorySize="pvalue", foldChange=NULL)  # si foldChange disponible
goplot(enrich_tf)
```

```{r}

```

UMAP based on TF activity

```{r}
DefaultAssay(data_p3) <- 'TF'
all.genes <- rownames(data_p3)
data_p3 <- ScaleData(data_p3, features = all.genes)
data_p3<-RunPCA(data_p3, assay="TF", reduction.name = "pca_tf", features = all.genes)
data_p3 <- RunUMAP(data_p3, dims = 1:10,  reduction.name = "umap_tf")

cell_palette <- c(
  "Intermediate population" = "#1f77b4",
  "B intermediate" = "#ff7f0e",
  "B naive" = "#ffd92f",
  "B memory" = "#582d15ff",
  "Macrophage" = "#984ea3",
  "B" = "#f2a890ff")
plot<-DimPlot(data_p3, reduction = "umap_tf", group.by = "bridge_annot", cols = cell_palette) + ggtitle("UMAP based on TF activity, P3")+ theme(plot.title = element_text(size = 10, hjust = 0.5), legend.text = element_text(size =8 ))
plot
ggsave("Plots/umap_tf_P3.jpeg", plot, width = 2000, height = 1000, units = "px")
```

```{r}
plot<-DimPlot2(
  data_p3,
  features = c("tf_CEBPA", "tf_NFIX"),
  ncol = 2,
  cols=list("tf_CEBPA"="D","tf_NFIX"="D" ),
  theme = NoAxes(), 
  reduction = "umap",
  label.size=2
)
plot
ggsave("Plots/tf_activity_P3.jpeg", plot, width = 2000, height = 1000, units = "px")
```

```{r}
selectedResolution <- "Annotation_B" # select resolution
# Split the cells by cluster:
cellClusters<-data_p3@meta.data %>%
  dplyr::select(Annotation) %>%
  dplyr::mutate(Annotation=dplyr::recode(Annotation, "Bridge" = "Intermediate population")) 

cellClusters$Annotation_B <- ifelse(cellClusters$Annotation %in% c("B memory", "B intermediate", "B naive", "B"),"B cells",cellClusters$Annotation)

cellsPerCluster <- split(rownames(cellClusters), cellClusters[,selectedResolution]) 
regulonAUC <- read.csv("/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/step_3_aucell_p3.csv", row.names = 1, check.names=FALSE)
regulonAUC <-t(regulonAUC)
regulonAUC <- regulonAUC[onlyNonDuplicatedExtended(rownames(regulonAUC)),]
# Calculate average expression:
regulonAUC_mat <- as.matrix(regulonAUC)

regulonActivity_byCellType <- sapply(
  cellsPerCluster,
  function(cells) {
    rowMeans(regulonAUC_mat[, cells, drop = FALSE])
  })

# Scale expression:
regulonActivity_byCellType_Scaled <- t(scale(t(regulonActivity_byCellType), center = T, scale=T))

# Plots
options(repr.plot.width=8, repr.plot.height=10) # To set the figure size in Jupyter
hm <- draw(ComplexHeatmap::Heatmap(regulonActivity_byCellType_Scaled, name="Regulon activity",
                       row_names_gp=grid::gpar(fontsize=6), show_row_names = FALSE, column_names_rot = 45)) # row font size
regulonOrder <- rownames(regulonActivity_byCellType_Scaled)[row_order(hm)] # to save the clustered regulons for later
```

```{r}
rss <- calcRSS(AUC = regulonAUC,cellAnnotation = cellClusters[colnames(regulonAUC), selectedResolution])
## Showing regulons and cell types with any RSS > 0.01 
rssPlot <- plotRSS(rss)
plotly::ggplotly(rssPlot$plot)
options(repr.plot.width=5, repr.plot.height=5) # To set the figure size in Jupyter
plotRSS_oneSet(rss, setName = "Macrophage",n=10) # cluster ID
plotRSS_oneSet(rss, setName = "B cells",n=10) # cluster ID
par(
  cex.axis = 1.5,   # axes ticks
  cex.lab = 1.8,    # titres des axes
  cex.main = 2      # titre principal
)
plotRSS_oneSet(rss, setName = "Intermediate population",n=10) # cluster ID
```

```{r}
tf_auc <- read.csv(params$file_step3_p2, row.names = 1, check.names=FALSE)
colnames(tf_auc) <- gsub("[(+)]", "", colnames(tf_auc))
tf<-tf_auc[,c("NANOG","CEBPA","IRF8","SPI1","BACH2","GATA1","CEBPB","CEBPD")]
tf$TF<- ifelse(tf$NANOG == 0 & tf$CEBPA == 0, "None",ifelse(tf$NANOG != 0 & tf$CEBPA == 0, "NANOG",ifelse(tf$NANOG == 0 & tf$CEBPA != 0, "CEBPA","NANOG+CEBPA")))
# Add in the seurat metadata
data_p2 <- AddMetaData(data_p2, metadata = tf)
DimPlot(data_p2,reduction="umap", group.by = "TF",cols =c("CEBPA"="red", "None"="grey", "NANOG"="blue","NANOG+CEBPA"="darkgreen"))
DimPlot(data_p2,reduction="umap", group.by = "CEBPD")

FeaturePlot(data_p2,features = "CEBPD",reduction = "umap")
```

# CardamomOT

```{r}
data_cardamom<-data_p2
data_cardamom$time<-data_cardamom$timepoint
data_cardamom$cell_type<-data_cardamom$Annotation
data_cardamom@meta.data <- data_cardamom@meta.data[, c("time", "cell_type"), drop = FALSE]
data_cardamom$time <- c(D1 = 1,D4 = 4,D8 = 8,D11 = 11,D14 = 14)[data_cardamom$time]

data_cardamom$time <- recode(
  data_cardamom$time,
  "D1" = 1,
  "D4" = 4,
  "D8" = 8,
  "D11" = 11,
  "D14" = 14
) |> as.numeric()

ConvertSeuratToH5ad <- function(seuratobj, output, raw = FALSE) {
  library(Seurat)
  library(SeuratDisk)
  library(SeuratObject)
  seuratobj$time <- as.character(seuratobj$time)
  # Convert Seurat v5 in V3/4
  seuratobj[["RNA"]] <- as(object = seuratobj[["RNA"]], Class = "Assay")
  # Delete scale data because put in the main layer (adata.X) in scanpy files
  seuratobj[["RNA"]]@scale.data <- matrix() 
  # Raw data save
  if (raw == TRUE) {
    seuratobj[["RNA"]]@data <- seuratobj[["RNA"]]@counts
  }
  # Convert
  SaveH5Seurat(seuratobj, filename = paste0(output, ".h5Seurat"), overwrite = TRUE)
  SeuratDisk::Convert(paste0(output, ".h5Seurat"),dest = "h5ad",overwrite = TRUE)
}

ConvertSeuratToH5ad(data_cardamom,"/home/a.blanc-boekholt/Documents/Singlecell-R/my_project/Data/data",raw=TRUE)
```

```{bash}
conda activate cardamom_env
python -m CardamomOT.cli pipeline -i my_project -s full -r 0.6 -c 1 -m 0.5 # in the folder my_projet in need a Data folder with data.h5ad
```

# SINCERITIES

```{r}
library(kSamples)
library(glmnet)
library(ppcor)

# Data
DATA <- GetAssayData(data_p2, assay = "RNA", layer = "counts")
DATA <- as.data.frame(t(as.matrix(DATA)))
DATA$Time <- data_p2$timepoint
DATA$Time <- as.numeric(sub("^D", "", DATA$Time))
rownames(DATA) <- NULL
write.csv(DATA,"data_p2_sincerities")
uploading <- dget("SINCERITIES functions/uploading.R")
DATA <- uploading("data_p2_sincerities")

# SINCERITIES
SINCERITIES <- dget("SINCERITIES functions/SINCERITIES.R")
SINCERITIES_PLUS <- dget("SINCERITIES functions/SINCERITIES_PLUS.R")
result = SINCERITIES(DATA) 
result = SINCERITIES_PLUS(DATA) # less than 5 timepoint  
result <- SINCERITIES_CROSS_VALIDATION(DATA)

#Final ranked list of regulatory edges
adj_matrix <- result$adj_matrix/max(result$adj_matrix)
final_ranked_predictions <- dget("SINCERITIES functions/final_ranked_predictions.R")
table <- final_ranked_predictions(adj_matrix,DATA$genes,SIGN=1,saveFile = TRUE)
```

Test SINECRITIES on one only branch (no differentiation branches) and keep just some genes (by DEG between timepoint)

```{r}
# keep only B cells cluster 0 and bridge
data_p2<-subset(data_p2, subset = (Annotation %in% c("B intermediate", "B naive", "B memory", "B", "Bridge","Monocyte","Macrophage")))
data_p2<-run_dim_red(data_p2)
data_p2<-FindClusters(data_p2, resolution = 0.1)
DimPlot(data_p2, reduction="umap")
data_p2<-subset(data_p2, subset = (seurat_clusters %in% c("0","3")))
data_p2<-subset(data_p2, subset = (Annotation %in% c("B","B intermediate","B memory","B naive","Bridge")))
DimPlot(data_p2, reduction="umap")

# features selection : DEG between timepoint
markers<-FindAllMarkers(data_p2, group.by="timepoint", only.pos= FALSE)
markers_sig <- subset(markers,p_val_adj < 0.05)
genes_DE <- unique(markers_sig$gene)
data_p2_DE <- subset(data_p2, features = genes_DE)
# keep only TFs
Tf_list<-read.table("/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Matrix/TF_names.txt")
tf_DE<-markers_sig[markers_sig$gene %in% Tf_list$V1, ]
head(genes_DE)
nrow(genes_DE)
write.csv(genes_DE,"genes_de.csv")

# Data
DATA <- GetAssayData(data_p2_DE, assay = "RNA", layer = "counts")
DATA <- as.data.frame(t(as.matrix(DATA)))
DATA$Time <- data_p2_DE$timepoint
DATA$Time <- as.numeric(sub("^D", "", DATA$Time))
rownames(DATA) <- NULL
write.csv(DATA,"data_p2_sincerities_sub")

# Sincerities 
uploading <- dget("SINCERITIES functions/uploading.R")
DATA <- uploading("data_p2_sincerities_sub")

# SINCERITIES
# Pb sincerities : Add sum(apply(DISTANCE_matrix, 2, function(x) length(unique(x)) <= 1))
SINCERITIES <- dget("SINCERITIES functions/SINCERITIES.R")
result = SINCERITIES(DATA) 
```

```{r}
# keep only B cells cluster 0 and bridge
data_p2<-subset(data_p2, subset = (Annotation %in% c("B intermediate", "B naive", "B memory", "B", "Bridge","Monocyte","Macrophage")))
data_p2<-run_dim_red(data_p2)
data_p2<-FindClusters(data_p2, resolution = 0.1)
DimPlot(data_p2, reduction="umap")
data_p2<-subset(data_p2, subset = (seurat_clusters %in% c("0","3")))
data_p2<-subset(data_p2, subset = (Annotation %in% c("B","B intermediate","B memory","B naive","Bridge")))
DimPlot(data_p2, reduction="umap")

# Filtrer et garder que les Tf, puis caluler les variation d'activité a trevers les timepoint 
# AUC score dataframe
auc_mtx <- read.csv(params$file_step3_p2, row.names = 1, check.names=FALSE)
auc_mtx <- as.data.frame(auc_mtx)
colnames(auc_mtx) <- gsub("[(+)]", "", colnames(auc_mtx))

auc_mtx <- auc_mtx[colnames(data_p2),]
auc_mtx$CellID <- rownames(auc_mtx)

# Add timepoint
auc_mtx$timepoint <- data_p2$timepoint[
  match(auc_mtx$CellID, colnames(data_p2))]

# Mean activity per tiempoint
avg_activity <- auc_mtx %>%
  group_by(timepoint) %>%
  summarise(across(-CellID, mean, na.rm = TRUE))
# Variation
variation <- apply(avg_activity[, -1], 2, function(x) max(x) - min(x))
variation <- sort(variation, decreasing = TRUE)
variation_df <- data.frame(
  TF = names(variation),
  variation = as.numeric(variation),
  row.names = NULL)
TF_keep <- names(variation)[1:100]

```

```{r}
# Data
data_p2_TF <- subset(data_p2, features = TF_keep)
DATA <- GetAssayData(data_p2_TF, assay = "RNA", layer = "counts")
DATA <- as.data.frame(t(as.matrix(DATA)))
DATA$Time <- data_p2_TF$timepoint
DATA$Time <- as.numeric(sub("^D", "", DATA$Time))
rownames(DATA) <- NULL
write.csv(DATA,"data_p2_sincerities_TF")

# Sincerities 
uploading <- dget("SINCERITIES functions/uploading.R")
DATA <- uploading("data_p2_sincerities_TF")

# SINCERITIES
# Pb sincerities : Add sum(apply(DISTANCE_matrix, 2, function(x) length(unique(x)) <= 1))
SINCERITIES_PLUS <- dget("SINCERITIES functions/SINCERITIES_PLUS.R")
result = SINCERITIES_PLUS(DATA) 
```

```{r}
adj_matrix<-result[["adj_matrix"]]
diag(adj_matrix) <- 0   # enlever l'auto-régulation

dim(adj_matrix)
library(tidyr)
library(igraph)
edges <- as.data.frame(adj_matrix) %>%
  mutate(regulator = rownames(adj_matrix)) %>%
  pivot_longer(
    cols = -regulator,
    names_to = "target",
    values_to = "weight"
  ) %>%
  filter(weight > 0) %>%
  arrange(desc(weight))

head(edges)
nrow(edges)

hist(log10(edges$weight), breaks = 100,
     main = "Distribution des poids (log10)", xlab = "log10(weight)")

threshold <- quantile(edges$weight, 0.99)
filtered <- edges %>% filter(weight >= threshold)
nrow(filtered)

top_per_target <- edges %>%
  group_by(target) %>%
  slice_max(weight, n = 5) %>%
  ungroup()

g <- graph_from_data_frame(filtered, directed = TRUE)

vcount(g)   # nombre de gènes
ecount(g)   # nombre d'arêtes

# Identifier les hubs = régulateurs avec le plus haut out-degree (pondéré)
hub_scores <- strength(g, mode = "out", weights = E(g)$weight)
sort(hub_scores, decreasing = TRUE)[1:20]

plot(
  g,
  vertex.size = 5,
  vertex.label.cex = 0.6,
  edge.width = E(g)$weight * 2,
  edge.arrow.size = 0.3,
  layout = layout_with_fr(g)
)
```

```{r}
library(dplyr)
library(tidyr)
library(igraph)

diag(adj_matrix) <- 0  # enlever l'auto-régulation

edges <- as.data.frame(adj_matrix) %>%
  mutate(regulator = rownames(adj_matrix)) %>%
  pivot_longer(cols = -regulator, names_to = "target", values_to = "weight") %>%
  filter(weight != 0) %>%
  mutate(
    sign = ifelse(weight > 0, "activation", "repression"),
    color = ifelse(weight > 0, "forestgreen", "firebrick")
  )

top_edges <- edges %>%
  slice_max(abs(weight), n = 100)

g <- graph_from_data_frame(top_edges, directed = TRUE)

# Attribuer la couleur des arêtes selon le signe
E(g)$color <- edges$color
E(g)$width <- abs(edges$weight) * 3   # épaisseur proportionnelle à la confiance

plot(
  g,
  vertex.size = 6,
  vertex.label.cex = 0.6,
  vertex.color = "lightgrey",
  edge.color = E(g)$color,
  edge.width = 1.5,
  edge.arrow.size = 0.3,
  layout = layout_with_fr(g, weights = NA),
  main = "GRN inféré par SINCERITIES\nVert = activation, Rouge = répression"
)

legend("bottomleft",
       legend = c("Activation", "Répression"),
       col = c("forestgreen", "firebrick"),
       lty = 1, lwd = 2, bty = "n")
```

```{r}
# AUC score dataframe
df <- read.csv(params$file_step3_p2, row.names = 1, check.names=FALSE)
df <- as.data.frame(df)
colnames(df) <- gsub("[(+)]", "", colnames(df))

df <- df[colnames(data_p2),]
df$CellID <- rownames(df)

# Add timepoint
df$timepoint <- data_p2$timepoint[
  match(df$CellID, colnames(data_p2))]
tp_order <- c("D1", "D4", "D8", "D11", "D14")
df$Timepoint <- factor(df$timepoint, levels = tp_order)

non_tf_cols <- c("Timepoint", "CellID")
tf_cols <- setdiff(colnames(df), non_tf_cols)

## mean activity

mean_activity <- df %>%
  group_by(Timepoint) %>%
  summarise(across(all_of(tf_cols), mean, na.rm = TRUE)) %>%
  ungroup()

mean_activity_long <- mean_activity %>%
  pivot_longer(cols = all_of(tf_cols), names_to = "TF", values_to = "mean_activity")


## Delta timepoint calculation (D4-D1, D8-D4, D11-D8, D14-D11)
delta_consecutive <- mean_activity_long %>%
  arrange(TF, Timepoint) %>%
  group_by(TF) %>%
  mutate(
    delta = mean_activity - lag(mean_activity),
    transition = paste0(lag(as.character(Timepoint)), "->", as.character(Timepoint))
  ) %>%
  filter(!is.na(delta)) %>%   # enlève la première ligne (pas de delta pour D1)
  ungroup()

# Order
transition_order <- c("D1->D4", "D4->D8", "D8->D11", "D11->D14")
delta_consecutive$transition <- factor(delta_consecutive$transition, levels = transition_order)

## evolution specific TFs
tf_of_interest <- c("CEBPA", "SPI1", "NANOG")

plot_data_1 <- delta_consecutive %>%
  filter(TF %in% tf_of_interest)

p1 <- ggplot(plot_data_1, aes(x = transition, y = delta, color = TF, group = TF)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  labs(
    title = "Change in TF activity between consecutive time points",
    x = "Transition between timepoint",
    y = "TF activity delta (Tn - Tn-1)",
    color = "TF"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 0))

print(p1)
ggsave("tf_specifiques_deltas.png", p1, width = 8, height = 5, dpi = 300)

tf <- df %>%
  dplyr::select(timepoint, SPI1)
tf$timepoint <- factor(
  tf$timepoint,
  levels = c("D1", "D4", "D8", "D11", "D14")
)
ggplot(tf, aes(x = timepoint, y = SPI1, fill = timepoint)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white") +
  theme_classic() +
  labs(
    title = "SPI1 TF activity across timepoints",
    x = "Timepoint",
    y = "SPI1 activity"
  )

tf <- df %>%
  dplyr::select(timepoint, CEBPA)
tf$timepoint <- factor(
  tf$timepoint,
  levels = c("D1", "D4", "D8", "D11", "D14")
)
ggplot(tf, aes(x = timepoint, y = CEBPA, fill = timepoint)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white") +
  theme_classic() +
  labs(
    title = "CEBPA TF activity across timepoints",
    x = "Timepoint",
    y = "CEBPA activity"
  )

## Delta D1 -> D14
delta_total <- mean_activity_long %>%
  filter(Timepoint %in% c("D1", "D14")) %>%
  pivot_wider(names_from = Timepoint, values_from = mean_activity) %>%
  mutate(
    delta_D1_D14 = D14 - D1,
    direction = ifelse(delta_D1_D14 >= 0, "Increase", "Decrease")
  ) %>%
  arrange(delta_D1_D14)

delta_filtered <- delta_total %>%
  filter(abs(delta_D1_D14) > 0.005)

p2 <- ggplot(delta_filtered, aes(x = TF, y = delta_D1_D14, fill = direction)) +
  geom_col() +
  geom_hline(yintercept = 0, color = "grey30") +
  coord_flip() +
  scale_fill_manual(values = c("Increase" = "#d73027", "Decrease" = "#4575b4")) +
  labs(
    title = "Change in TF activity between D1 and D14",
    x = "TF",
    y = "TFs activity delta (D14 - D1)",
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 10))
print(p2)

ggsave("tf_all_delta_D1_D14.png", p2, width = 8, height = 20, dpi = 300, limitsize = FALSE)
```

# CollecTRI reduction

Reduce the size of the Influence graph based on CollecTRI, as its size is causing BoNesis to crash

Selection of TF based on the literature and DEG on the TF activity between macrostates

```{r}
important_tfs = c(
  # B-cell/CLL identity
    "PAX5","EBF1","IKZF1","IKZF3","BCL6","BACH2","TBX21", "ETV6",
    # Myeloid identity
    "SPI1","CEBPA","CEBPB","CEBPD","RUNX1","KLF4","MAFB","CYBB",
    # Inflammatory signaling
    "STAT1","STAT3","STAT5A","STAT6","NFKB1","RELA","IRF1","IRF5","IRF8",
    # Immediate early genes
    "FOS","JUN","ATF3", "NANOG",
    # Proliferation
    "MYC","TP53","FOXO1","FOXO3",
    # Transdifferenciation 
    "IRF4")

# Differential TF activity
# Seurat objet 
data_p2<-readRDS(params$file_p2)
Idents(data_p2) <- data_p2$final_annot
data_p2<-subset(data_p2, subset = (final_annot %in% c("B","B intermediate","B memory","B naive","Macrophage")))
# Macrostates annotation 
annot <- read.csv("~/Documents/Singlecell-R/Scripts/Matrix/cell_annotations_macrostates.csv")
data_p2$macrostates <- "None"
idx <- match(colnames(data_p2), annot$cell_id)
data_p2$macrostates[!is.na(idx)] <- annot$macrostate[idx[!is.na(idx)]]
# AUC score dataframe
auc_mtx <- read.csv(params$file_step3_p2, row.names = 1, check.names=FALSE)
auc_mtx <- t(auc_mtx)
auc_mtx <- as.data.frame(auc_mtx)
rownames(auc_mtx) <- gsub("[(+)]", "", rownames(auc_mtx))
auc_mtx[] <- lapply(auc_mtx, as.numeric)
auc_mtx <- as.matrix(auc_mtx)

data_p2[['TF']] <- CreateAssayObject(data = auc_mtx)
DefaultAssay(data_p2) <- 'RNA'

Idents(data_p2) <- data_p2$macrostates
data_macrostates<-subset(data_p2, subset = (macrostates %in% c("S0","S3","S2","S11")))
markers_tf <- FindAllMarkers(data_macrostates,assay = "TF",slot = "data", test.use = "wilcox",logfc.threshold = 0,min.pct = 0)
markers_tf <- subset(markers_tf, p_val_adj < 0.05)

deg_tfs<-unique(markers_tf$gene)

# Union 
tfs<-union(important_tfs, deg_tfs)

# Reduce CollecTRI
collectri <- read.table("/home/a.blanc-boekholt/Documents/Singlecell-R/Scripts/Data/CollecTRI_regulons.csv",sep=",", header=TRUE)
collectri_filter<-subset(collectri, source %in% tfs)
```
