---
title: "Copy number alteration"
params: 
  patient: "P2"
  file: "~/Documents/Singlecell-R/Scripts/Matrix/CLL_Seurat_P2.rds"
  filebis: "~/Documents/Singlecell-R/Scripts/Matrix/CLL_Seurat_P3"
format: html
editor: visual
---

## Clustering and GeneOntology on the sub population

## CLL and myeloid sub population

```{r}
#| message: false
library(Seurat)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggplot2)
```

```{r}
data_p2<-readRDS(params$file)
```

Define sub population in CLL population

```{r}
data_cll<-subset(data_p2, subset = (Annotation %in% c("B intermediate", "B naive", "B memory", "B", "Bridge")))
```

```{r}
data_cll<-NormalizeData(data_cll)
data_cll <- FindVariableFeatures(data_cll, selection.method = "vst", nfeatures = 2000)
all.genes <- rownames(data_cll)
data_cll <- ScaleData(data_cll, features = all.genes)
data_cll <- RunPCA(data_cll, features = VariableFeatures(object = data_cll))
data_cll <- FindNeighbors(data_cll, dims = 1:10)
data_cll <- FindClusters(data_cll, resolution = 0.1)
data_cll <- RunUMAP(data_cll, dims = 1:10)
data_cll <- RunTSNE(data_cll, dims = 1:10)
```

```{r}
data_cll$timepoint <- factor(
  data_cll$timepoint,
  levels = c("D1", "D4", "D8", "D11", "D14")
)

png("~/Documents/Singlecell-R/Scripts/Plots/CLL_umap_cluster.jpeg")
DimPlot(data_cll, reduction = "umap")
dev.off()
png("~/Documents/Singlecell-R/Scripts/Plots/CLL_tsne.jpeg")
DimPlot(data_cll, reduction = "tsne")
dev.off()
png("~/Documents/Singlecell-R/Scripts/Plots/CLL_umap_celltype.jpeg")
DimPlot(data_cll, reduction = "umap",group.by = "Annotation")
dev.off()
png("~/Documents/Singlecell-R/Scripts/Plots/CLL_tsne_celltype.jpeg")
DimPlot(data_cll, reduction = "tsne",group.by = "Annotation")
dev.off()
png("~/Documents/Singlecell-R/Scripts/Plots/CLL_umap_clone.jpeg")
DimPlot(data_cll, reduction = "umap",group.by = "clone_opt")
dev.off()
png("~/Documents/Singlecell-R/Scripts/Plots/CLL_umap_time.jpeg")
DimPlot(data_cll, reduction = "umap",group.by = "timepoint")
dev.off()
```

Quantification : proportion

```{r}
df <- data_cll@meta.data

df$timepoint <- factor(
  df$timepoint,
  levels = c("D1", "D4", "D8", "D11", "D14")
)

df$clone_opt <- as.factor(df$clone_opt)

tumor_prop <- df %>%
  group_by(seurat_clusters, compartment_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(tumor_prop, aes(x = seurat_clusters, y = prop, fill = compartment_opt)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Cluster") + labs(fill = "Cell annotation\n(tumor/normal)")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/propor_tumor_cll_p2.jpeg", plot, width = 3000, height = 2000, units = "px")

df %>%
  group_by(compartment_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(prop = n / sum(n))


clone_prop <- df %>%
  group_by(seurat_clusters, clone_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(clone_prop, aes(x = seurat_clusters, y = prop, fill = clone_opt)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Cluster") + labs(fill = "Clones")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/propor_clones_cll_p2.jpeg", plot, width = 3000, height = 2000, units = "px")

time_prop <- df %>%
  group_by(seurat_clusters, timepoint) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(time_prop, aes(x = seurat_clusters, y = prop, fill = timepoint)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Clusters") + labs(fill = "Timepoint")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/propor_time_cll_p2.jpeg", plot, width = 3000, height = 2000, units = "px")
```

Differential gene analysis and Gene Ontology

```{r}
res <- FindAllMarkers(data_cll, only.pos = TRUE)
```

```{r}
for (clus in 0:5) {

  res_clus <- subset(res, res$cluster == as.character(clus))
  genes <- rownames(res_clus)

  Go_results <- enrichGO(
    gene = genes, OrgDb = org.Hs.eg.db, keyType = "SYMBOL", ont = "BP")
  
  plot_go <- barplot(Go_results, showCategory = 20) +
    ggtitle(paste("GO Enrichment Analysis CLL cells - Cluster", clus)) +
    theme_bw() + theme(plot.title = element_text(hjust = 0.5, face = "bold"), axis.text.y = element_text(size = 14))
  
  print(plot_go)

  ggsave(
    filename = paste0("~/Documents/Singlecell-R/Scripts/Plots/GO_cluster_cll", clus, ".jpeg"), plot = plot_go,
    width = 4000, height = 3000, units = "px"
  )
}
```

Define sub population in myeloid population (mono+macro+bridge)

```{r}
data_myel<-subset(data_p2, subset = (Annotation %in% c("Monocyte", "Macrophage", "Bridge")))
data_myel<-NormalizeData(data_myel)
data_myel <- FindVariableFeatures(data_myel, selection.method = "vst", nfeatures = 2000)
all.genes <- rownames(data_myel)
data_myel <- ScaleData(data_myel, features = all.genes)
data_myel <- RunPCA(data_myel, features = VariableFeatures(object = data_myel))
data_myel <- FindNeighbors(data_myel, dims = 1:10)
data_myel <- FindClusters(data_myel, resolution = 0.2)
data_myel <- RunUMAP(data_myel, dims = 1:10)
data_myel <- RunTSNE(data_myel, dims = 1:10)
```

```{r}
png("~/Documents/Singlecell-R/Scripts/Plots/myel_umap.jpeg")
DimPlot(data_myel, reduction = "umap")
dev.off()

png("~/Documents/Singlecell-R/Scripts/Plots/myel_tsne.jpeg")
DimPlot(data_myel, reduction = "tsne")
dev.off()

png("~/Documents/Singlecell-R/Scripts/Plots/myel_umap_celltype.jpeg")
DimPlot(data_myel, reduction = "umap",group.by = "Annotation")
dev.off()

png("~/Documents/Singlecell-R/Scripts/Plots/myel_tsne_celltype.jpeg")
DimPlot(data_myel, reduction = "tsne",group.by = "Annotation")
dev.off()

png("~/Documents/Singlecell-R/Scripts/Plots/myel_umap_clone.jpeg")
DimPlot(data_myel, reduction = "umap",group.by = "clone_opt")
dev.off()

png("~/Documents/Singlecell-R/Scripts/Plots/myel_umap_time.jpeg")
DimPlot(data_myel, reduction = "umap",group.by = "timepoint")
dev.off()
```

```{r}
res <- FindAllMarkers(data_myel, only.pos = TRUE)
```

```{r}
for (clus in 0:5) {

  res_clus <- subset(res, res$cluster == as.character(clus))
  genes <- rownames(res_clus)

  Go_results <- enrichGO(
    gene = genes, OrgDb = org.Hs.eg.db, keyType = "SYMBOL", ont = "BP")
  
  plot_go <- barplot(Go_results, showCategory = 20) +
    ggtitle(paste("GO Enrichment Analysis Myeloid cells - Cluster", clus)) +
    theme_bw() + theme(plot.title = element_text(hjust = 0.5, face = "bold"), axis.text.y = element_text(size = 14))
  
  print(plot_go)

  ggsave(
    filename = paste0("~/Documents/Singlecell-R/Scripts/Plots/GO_cluster_myel", clus, ".jpeg"), plot = plot_go,
    width = 4000, height = 3000, units = "px"
  )
}
```

Quantification : proportion

```{r}
df <- data_myel@meta.data

df$timepoint <- factor(
  df$timepoint,
  levels = c("D1", "D4", "D8", "D11", "D14")
)

df$clone_opt <- as.factor(df$clone_opt)

tumor_prop <- df %>%
  group_by(seurat_clusters, compartment_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(tumor_prop, aes(x = seurat_clusters, y = prop, fill = compartment_opt)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Cluster myeloid + intermediate population") + labs(fill = "Cell annotation\n(tumor/normal)")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/propor_tumor_myel_p2.jpeg", plot, width = 3000, height = 2000, units = "px")

df %>%
  group_by(compartment_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(prop = n / sum(n))


clone_prop <- df %>%
  group_by(seurat_clusters, clone_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(clone_prop, aes(x = seurat_clusters, y = prop, fill = clone_opt)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Cluster myeloid + intermediate population") + labs(fill = "Clones")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/propor_clones_myel_p2.jpeg", plot, width = 3000, height = 2000, units = "px")

time_prop <- df %>%
  group_by(seurat_clusters, timepoint) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(time_prop, aes(x = seurat_clusters, y = prop, fill = timepoint)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Clusters myeloid + intermediate population") + labs(fill = "Timepoint")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/propor_time_myel_p2.jpeg", plot, width = 3000, height = 2000, units = "px")
```

Compare cluster 0 and 1 (macrophages), in order to define M1 and M2

```{r}
res_cluster0 <- FindMarkers(data_myel, only.pos = TRUE, ident.1 = 0, ident.2=1)
res_cluster1 <- FindMarkers(data_myel, only.pos = TRUE, ident.1 = 1, ident.2=0)

res_cluster0 <- res_cluster0 %>%
  arrange(desc(avg_log2FC))
genelist<-head(rownames(res_cluster0),n=2000)
genelist<- data.frame(gene = genelist)
write.table(genelist, file = "~/Documents/Singlecell-R/Scripts/Matrix/res_clu0.tsv", sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)

res_cluster1 <- res_cluster1 %>%
  arrange(desc(avg_log2FC))
genelist<-head(rownames(res_cluster1),n=2000)
genelist<- data.frame(gene = genelist)
write.table(genelist, file = "~/Documents/Singlecell-R/Scripts/Matrix/res_clu1.tsv", sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
```

Surface proteins

```{r}
adt=unique(data_p2@assays[["ADT"]]@var.features)
plot <- DoHeatmap(data_myel, features = features, size = 4, assay = "ADT") +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16)
  )
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/heatmap_adt_p2.jpeg", plot, width = 3000, height = 2000, units = "px")
```

Sub population macro-mono

```{r}
data_mono<-subset(data_p2, Annotation %in% c("Monocyte", "Macrophage"))
data_mono<-run_dim_red(data_mono)
data_mono <- FindClusters(data_mono, resolution=0.4)
DimPlot(data_mono, reduction = "umap")
# Explore the 2 monocytes types
res_cluster6 <- FindMarkers(data_mono, only.pos = TRUE, ident.1 = 6, ident.2=7)
res_cluster7 <- FindMarkers(data_mono, only.pos = TRUE, ident.1 = 7, ident.2=6)
```

### Patient 3

```{r}
data_p3<-readRDS(params$filebis)
```

Define sub population in CLL population

```{r}
data_cll<-subset(data_p3, subset = (Annotation %in% c("B intermediate", "B naive", "B memory", "B", "Bridge")))
```

```{r}
data_cll<-NormalizeData(data_cll)
data_cll <- FindVariableFeatures(data_cll, selection.method = "vst", nfeatures = 2000)
all.genes <- rownames(data_cll)
data_cll <- ScaleData(data_cll, features = all.genes)
data_cll <- RunPCA(data_cll, features = VariableFeatures(object = data_cll))
data_cll <- FindNeighbors(data_cll, dims = 1:10)
data_cll <- FindClusters(data_cll, resolution = 0.1)
data_cll <- RunUMAP(data_cll, dims = 1:10)
data_cll <- RunTSNE(data_cll, dims = 1:10)
```

```{r}
png("~/Documents/Singlecell-R/Scripts/Plots/CLL_umap_cluster_p3.jpeg")
DimPlot(data_cll, reduction = "umap")
dev.off()
png("~/Documents/Singlecell-R/Scripts/Plots/CLL_tsne_p3.jpeg")
DimPlot(data_cll, reduction = "tsne")
dev.off()
png("~/Documents/Singlecell-R/Scripts/Plots/CLL_umap_celltype_p3.jpeg")
DimPlot(data_cll, reduction = "umap",group.by = "Annotation")
dev.off()
png("~/Documents/Singlecell-R/Scripts/Plots/CLL_tsne_celltype_p3.jpeg")
DimPlot(data_cll, reduction = "tsne",group.by = "Annotation")
dev.off()
png("~/Documents/Singlecell-R/Scripts/Plots/CLL_umap_clone_p3.jpeg")
DimPlot(data_cll, reduction = "umap",group.by = "clone_opt")
dev.off()
png("~/Documents/Singlecell-R/Scripts/Plots/CLL_umap_time_p3.jpeg")
DimPlot(data_cll, reduction = "umap",group.by = "timepoint")
dev.off()
```

Quantification : proportion

```{r}
df <- data_cll@meta.data

df$timepoint <- factor(
  df$timepoint,
  levels = c("D1", "D4", "D8", "D11", "D14")
)

df$clone_opt <- as.factor(df$clone_opt)

tumor_prop <- df %>%
  group_by(seurat_clusters, compartment_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(tumor_prop, aes(x = seurat_clusters, y = prop, fill = compartment_opt)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Cluster") + labs(fill = "Cell annotation\n(tumor/normal)")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/propor_tumor_cll_p3.jpeg", plot, width = 3000, height = 2000, units = "px")

df %>%
  group_by(compartment_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(prop = n / sum(n))

clone_prop <- df %>%
  group_by(seurat_clusters, clone_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(clone_prop, aes(x = seurat_clusters, y = prop, fill = clone_opt)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Cluster") + labs(fill = "Clones")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/propor_clones_cll_p3.jpeg", plot, width = 3000, height = 2000, units = "px")

time_prop <- df %>%
  group_by(seurat_clusters, timepoint) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(time_prop, aes(x = seurat_clusters, y = prop, fill = timepoint)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Clusters") + labs(fill = "Timepoint")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/propor_time_cll_p3.jpeg", plot, width = 3000, height = 2000, units = "px")
```

Differential gene analysis and Gene Ontology

```{r}
res <- FindAllMarkers(data_cll, only.pos = TRUE)
```

```{r}
for (clus in 0:5) {

  res_clus <- subset(res, res$cluster == as.character(clus))
  genes <- rownames(res_clus)

  Go_results <- enrichGO(
    gene = genes, OrgDb = org.Hs.eg.db, keyType = "SYMBOL", ont = "BP")
  
  plot_go <- barplot(Go_results, showCategory = 20) +
    ggtitle(paste("GO Enrichment Analysis CLL cells - Cluster", clus)) +
    theme_bw() + theme(plot.title = element_text(hjust = 0.5, face = "bold"), axis.text.y = element_text(size = 14))
  
  print(plot_go)

  ggsave(
    filename = paste0("~/Documents/Singlecell-R/Scripts/Plots/GO_cluster_cll_p3", clus, ".jpeg"), plot = plot_go,
    width = 4000, height = 3000, units = "px"
  )
}
```

Define sub population in myeloid population (mono+macro+bridge)

```{r}
data_myel<-subset(data_p3, subset = (Annotation %in% c("Monocyte", "Macrophage", "Bridge")))
data_myel<-NormalizeData(data_myel)
data_myel <- FindVariableFeatures(data_myel, selection.method = "vst", nfeatures = 2000)
all.genes <- rownames(data_myel)
data_myel <- ScaleData(data_myel, features = all.genes)
data_myel <- RunPCA(data_myel, features = VariableFeatures(object = data_myel))
data_myel <- FindNeighbors(data_myel, dims = 1:10)
data_myel <- FindClusters(data_myel, resolution = 0.08)
data_myel <- RunUMAP(data_myel, dims = 1:10)
data_myel <- RunTSNE(data_myel, dims = 1:10)
```

```{r}
png("~/Documents/Singlecell-R/Scripts/Plots/myel_umap_p3.jpeg")
DimPlot(data_myel, reduction = "umap")
dev.off()

png("~/Documents/Singlecell-R/Scripts/Plots/myel_tsne_p3.jpeg")
DimPlot(data_myel, reduction = "tsne")
dev.off()

png("~/Documents/Singlecell-R/Scripts/Plots/myel_umap_celltype_p3.jpeg")
DimPlot(data_myel, reduction = "umap",group.by = "Annotation")
dev.off()

png("~/Documents/Singlecell-R/Scripts/Plots/myel_tsne_celltype_p3.jpeg")
DimPlot(data_myel, reduction = "tsne",group.by = "Annotation")
dev.off()

png("~/Documents/Singlecell-R/Scripts/Plots/myel_umap_clone_p3.jpeg")
DimPlot(data_myel, reduction = "umap",group.by = "clone_opt")
dev.off()

png("~/Documents/Singlecell-R/Scripts/Plots/myel_umap_time_p3.jpeg")
DimPlot(data_myel, reduction = "umap",group.by = "timepoint")
dev.off()
```

```{r}
res <- FindAllMarkers(data_myel, only.pos = TRUE)
```

```{r}
for (clus in 0:5) {

  res_clus <- subset(res, res$cluster == as.character(clus))
  genes <- rownames(res_clus)

  Go_results <- enrichGO(
    gene = genes, OrgDb = org.Hs.eg.db, keyType = "SYMBOL", ont = "BP")
  
  plot_go <- barplot(Go_results, showCategory = 20) +
    ggtitle(paste("GO Enrichment Analysis Myeloid cells - Cluster", clus)) +
    theme_bw() + theme(plot.title = element_text(hjust = 0.5, face = "bold"), axis.text.y = element_text(size = 14))
  
  print(plot_go)

  ggsave(
    filename = paste0("~/Documents/Singlecell-R/Scripts/Plots/GO_cluster_myel_p3", clus, ".jpeg"), plot = plot_go,
    width = 4000, height = 3000, units = "px"
  )
}
```

Quantification : proportion

```{r}
df <- data_myel@meta.data

df$timepoint <- factor(
  df$timepoint,
  levels = c("D1", "D4", "D8", "D11", "D14")
)

df$clone_opt <- as.factor(df$clone_opt)

tumor_prop <- df %>%
  group_by(seurat_clusters, compartment_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(tumor_prop, aes(x = seurat_clusters, y = prop, fill = compartment_opt)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Cluster myeloid + intermediate population") + labs(fill = "Cell annotation\n(tumor/normal)")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/propor_tumor_myel_p3.jpeg", plot, width = 3000, height = 2000, units = "px")

df %>%
  group_by(compartment_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(prop = n / sum(n))


clone_prop <- df %>%
  group_by(seurat_clusters, clone_opt) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(clone_prop, aes(x = seurat_clusters, y = prop, fill = clone_opt)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Cluster myeloid + intermediate population") + labs(fill = "Clones")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/propor_clones_myel_p3.jpeg", plot, width = 3000, height = 2000, units = "px")

time_prop <- df %>%
  group_by(seurat_clusters, timepoint) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters) %>%
  mutate(prop = n / sum(n))

plot<-ggplot(time_prop, aes(x = seurat_clusters, y = prop, fill = timepoint)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Cell proportion") + xlab("Clusters myeloid + intermediate population") + labs(fill = "Timepoint")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/propor_time_myel_p3.jpeg", plot, width = 3000, height = 2000, units = "px")
```

Surface proteins

```{r}
adt=unique(data_p3@assays[["ADT"]]@var.features)
plot <- DoHeatmap(data_myel, features = features, size = 4, assay = "ADT") +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16)
  )
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/heatmap_adt_p3.jpeg", plot, width = 3000, height = 2000, units = "px")
```

## Annotated B cells in T cells population

```{r}
plot<-DimPlot(data_p3, reduction = 'wnn.umap', label = TRUE, repel = TRUE, label.size = 2.5, group.by = "Annotation")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/BinT_wnn_p3.jpeg", plot, width = 3000, height = 2000, units = "px")
```

```{r}
data_sub<-subset(data_p3, subset=final_annot %in% c("B","B intermediate","B memory","B naive","Macrophage", "Monocyte"))
plot<-DimPlot(data_sub, reduction = 'wnn.umap', label = TRUE, repel = TRUE, label.size = 2.5, group.by = "final_annot")
plot
ggsave("~/Documents/Singlecell-R/Scripts/Plots/BinT_wnn_sub_p3.jpeg", plot, width = 3000, height = 2000, units = "px")
```

Calculate new subset wnn umap coordinates

```{r}
data_sub<- FindMultiModalNeighbors(
  data_sub, reduction.list = list("pca", "apca"), 
  dims.list = list(1:30, 1:18), modality.weight.name = "RNA.weight"
)
data_sub<-RunUMAP(data_sub, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
DimPlot(data_sub, reduction = 'wnn.umap', label = TRUE, repel = TRUE, label.size = 2.5, group.by = "final_annot")
```

M1 and M2 annotation

```{r}
cell_palette <- c(
  "Macrophages M1" = "#1f77b4",
  "Macrophages M2" = "#93f290ff",
  "Monocytes" = "#e41a1c",
  "Macrophages" = "#984ea3")
ref <- fetchReference("blueprint_encode", "2024-02-26")
ref <- ref[, grepl("Monocytes|Macrophages|Myelocytes", ref$label.fine, ignore.case = TRUE)]
norm_counts <- LayerData(data_myel, assay = "RNA", layer = 'data')
annotation<-SingleR(test = norm_counts,
                  ref = ref, 
                  labels = ref$label.fine,
                  de.method = 'wilcox')
data_myel <- AddMetaData(data_myel, annotation$pruned.labels, col.name = 'Macrophages')

title="UMAP annotated by SingleR, patient 2"
plotumap<-DimPlot(data_myel, group.by="Macrophages", reduction = "umap", label = FALSE, repel = TRUE, label.size = 4, cols = cell_palette) + ggtitle(str_wrap(title, width = 100)) + theme(plot.title = element_text(size = 15, hjust = 0.5), legend.text = element_text(size = 12))
print(plotumap)
ggsave(paste0("bridge_SingleR_", params$patient, ".jpeg"), plotumap, width = 3000,
  height = 2000, units = "px")

table(data_p2$SingleR_fine)
 
```

Myeloid + CLL population

```{r}
data_sub<-subset(data_p2, subset= final_annot %in% c("B intermediate", "B naive", "B memory", "B", "Monocyte", "Macrophage"))
data_sub<-run_dim_red(data_sub)
data_sub <- FindClusters(data_sub, resolution = 0.1)
```

```{r}
# Selected only the main CLL cloud related to the bridge (cluster 0)
data_sub<-subset(data_sub, subset= seurat_clusters %in% c("0", "3"))
data_sub<-run_dim_red(data_sub)
data_sub <- FindClusters(data_sub, resolution = 0.1)
data_sub<-subset(data_sub, subset= seurat_clusters %in% c("0", "1"))
data_sub<-run_dim_red(data_sub)
data_sub <- FindClusters(data_sub, resolution = 0.1)
data_sub<-subset(data_sub, subset= seurat_clusters %in% c("0", "2","3"))
data_sub<-run_dim_red(data_sub)
data_sub@meta.data$global <- ifelse(
  data_sub@meta.data[["Annotation"]] %in% c("B memory", "B intermediate", "B naive", "B"), 
  "B cells", 
  data_sub@meta.data[["Annotation"]]
)

data_sub<-subset(data_sub, subset= seurat_clusters %in% c("2","3","4","5","6","7","8"))
data_sub<-run_dim_red(data_sub)

# Visualize the CLL "initial" short in the complete data 
Idents(data_sub) <- "seurat_clusters"
cells_B_initial <- colnames(data_sub)[Idents(data_sub) %in% c(2, 3)]
data_p2$macrostate <- data_p2$global
data_p2$macrostate[cells_B_initial] <- "B_initial"
```
