---
title: "Annotation Azimuth"
params: 
  patient: "P1"
  file: "CLL_P1_seurat_reduit.rds"
  type: "DDQC"
format: html
editor: visual
---

Automatic annotation with Azimuth. This script and Azimuth needs Seurat 5.2.1.

```{r}
#renv::remove("Seurat")
#devtools::install_github("satijalab/seurat", ref = "v5.2.1")
```

## Data and libraries initialization

```{r}
#| message: false
library(SeuratObject)
library(sp)
library(Seurat)
library(hdf5r)
library(dplyr)
library(patchwork)
library(ggplot2)
library(Azimuth)
library(stringr)
library(SeuratData)
library(celldex)
source("functions.R")
```

Read the data pre-processed

```{r}
data<-readRDS(params$file)
```

### Azimuth

Retrieve the Seurat object (containing both RNA and ADT data), but perform cell type annotation using only the RNA modality. The object must already be filtered (cells and genes), and must contain raw RNA counts, as Azimuth will apply its own normalization consistent with the reference dataset used (Hao et al., 2021).

Create an Azimuth reference using the BlueprintEncode dataset.

```{r}
bp <- celldex::BlueprintEncodeData()
logcounts <- assay(ref, "logcounts")
seurat_ref <- CreateSeuratObject(counts = logcounts)

seurat_ref <- NormalizeData(seurat_ref)
seurat_ref <- FindVariableFeatures(seurat_ref)
seurat_ref <- ScaleData(seurat_ref)
seurat_ref <- RunPCA(seurat_ref)
seurat_ref <- RunUMAP(seurat_ref, dims = 1:50, return.model = TRUE )

# Add labels
seurat_ref$celltype <- bp$label.fine

# Azimuth ref
azimuth_ref <- AzimuthReference(seurat_ref, refDR="pca",refUMAP = "umap", refAssay="RNA", metadata = "celltype")
```

```{r}
#| include: false
#InstallData("pbmcref")
data_annotated <- RunAzimuth(data, reference = "pbmcref")
```

## Visualization

UMAP

```{r}
title=paste("UMAP annotated by Azimuth, pre-processing type :", params$type, ", patient", params$patient)
plotumap<-DimPlot(data_annotated, group.by = "predicted.celltype.l1", reduction = "umap", label = TRUE, repel = TRUE, label.size = 2.5) + ggtitle(str_wrap(title, width = 50)) + theme(plot.title = element_text(size = 10, hjust = 0.5))
plotumap
ggsave(paste0(params$type,"_umap_Azimuth_", params$patient, ".jpeg"), plotumap)
```

UMAP by timepoint

```{r}
data_annotated$timepoint <- factor(
  data_annotated$timepoint,
  levels = c("D1", "D4", "D8", "D11", "D14")
)
title=paste("UMAP annotated by Azimuth by timepoint, pre-processing type :", params$type, ", patient", params$patient)
plot1<-DimPlot(data_annotated, group.by = "predicted.celltype.l1", reduction = "umap", split.by = "timepoint", label = TRUE, repel = TRUE, label.size = 2.5) + ggtitle(str_wrap(title, width = 50)) + theme(plot.title = element_text(size = 10, hjust = 0.5))
plot1
ggsave(paste0(params$type, "_umap_timepoint_Azimuth_", params$patient, ".jpeg"), plot1)
```

wnnUMAP

```{r}
#| message: false
#| warning: false
data_annotated <-run_adt_umap(data_annotated)
title=paste("wnnUMAP annotated by Azimuth, pre-processing type :", params$type, ", patient", params$patient)
plotwnn <- DimPlot(data_annotated, group.by = "predicted.celltype.l1", reduction = 'wnn.umap', label = TRUE, repel = TRUE, label.size = 2.5) + 
  NoLegend() + ggtitle(str_wrap(title, width = 50)) + theme(plot.title = element_text(size = 10, hjust = 0.5))
plotwnn
ggsave(paste0(params$type,"_wnnumap_Azimuth_", params$patient, ".jpeg"), plotwnn)
DefaultAssay(data_annotated) <- 'RNA'
```

t-SNE

```{r}
title=paste("t-SNE annotated by Azimuth, pre-processing type :", params$type, ", patient", params$patient)
plot1<-DimPlot(data_annotated, group.by="predicted.celltype.l1", reduction = 'tsne', label = TRUE, repel = TRUE, label.size = 2.5) + ggtitle(str_wrap(title, width = 50)) + theme(plot.title = element_text(size = 10, hjust = 0.5))
plot1
ggsave(paste0(params$type,"_tsne_Azimuth_", params$patient, ".jpeg"), plot1)
```

Bonsai

In this analysis, we use the single-cell Bonsai pipeline available at <https://bonsai.unibas.ch/bonsai/>. The TSV matrix is exported in the pre-processing script, and we supplemented it with an annotation TSV file containing cell barcodes and their corresponding cell type annotations.

```{r}
barcodes <- colnames(data_annotated)
annotations <- data_annotated$predicted.celltype.l1

df <- data.frame(
  barcode = barcodes,
  predicted_celltype = annotations
)
namefile=paste0("azimuth_annotation_", params$type, "_", params$patient, ".tsv")
write.table(df, file = namefile, sep = "\t", row.names = FALSE, quote = FALSE)
```

Isomap

```{r}
pc_mat <- Embeddings(data_annotated, reduction = "pca")[, 1:10]

iso <- isomap(dist(pc_mat), ndim = 2, k = 10)  # k = nb voisins, à ajuster

# Add Isomap reduction in the Seurat Object 
data_annotated[["isomap"]] <- CreateDimReducObject(
  embeddings = iso$points,
  key = "ISO_",
  assay = DefaultAssay(data_annotated)
)

title=paste("Isomap annotated by Azimuth, pre-processing type :", params$type, ", patient", params$patient)
DimPlot(data_annotated, reduction = "isomap") + ggtitle(str_wrap(title, width = 80)) + theme(plot.title = element_text(size = 10, hjust = 0.5))
```

## Save the annotation file

```{r}
namepath=paste0("CLL_azimuth_annotated_", params$type, "_", params$patient, ".rds")
saveRDS(data_annotated, file = namepath)
```
