---
title: "Annotation"
params: 
  patient: "P1"
  file: "CLL_P1_seurat.rds"
format: html
editor: visual
---

# Cell types annotation

This script aims to annotate cell clusters in order to identify the corresponding cell types. It combines both automatic and manual annotation approaches to improve accuracy and biological relevance.

## Data and libraries initialization

```{r}
#| message: false
library(SeuratObject)
library(sp)
library(Seurat)
library(hdf5r)
library(dplyr)
library(patchwork)
library(ggplot2)
library(Azimuth)
source("functions.R")
```

Read the data pre-processed

```{r}
data<-readRDS(params$file)
```

## Manual annotation

With WebCSEA and Seurat

```{r}
cluster6.markers <- FindMarkers(data, ident.1 = 6, ident.2 = c(0, 1, 2, 3, 4, 5))
```

```{r}
cluster6.markers <- FindMarkers(data, ident.1 = 6, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)
```

```{r}
VlnPlot(data, features = c("MS4A1", "CD79A"))
```

```{r}
FeaturePlot(data, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A", "FCGR3A", "LYZ", "PPBP",
    "CD8A"))
```

Extraction des marqueurs différencitiellement exprimés puis injectuon de ces mraqueurs l'outil WebCSEA (Dai Y, Hu R, Liu A, Cho KS, Manuel AM, Li X, Dong X, Jia P, Zhao Z, 2022).

```{r}
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", "CD8 T", "FCGR3A+ Mono",
    "NK", "DC", "Platelet")
names(new.cluster.ids) <- levels(data)
data <- RenameIdents(data, new.cluster.ids)
DimPlot(data, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```

```{r}
plot <- DimPlot(data, reduction = "umap", label = TRUE, label.size = 4.5) + xlab("UMAP 1") + ylab("UMAP 2") +
    theme(axis.title = element_text(size = 18), legend.text = element_text(size = 18)) + guides(colour = guide_legend(override.aes = list(size = 10)))
ggsave(filename = "./Plots/umap_annotated.jpg", height = 7, width = 12, plot = plot, quality = 50)
```

```{r}
saveRDS(data, file = "CLL_annoted.rds")
```

## Automatic annotation

With Azimuth

Récupérer le seurat object (ARN et ADT) mais annotation en utilisant seulement les ARN. l'object doit avoir été filtrés (cellules et gènes). Il doit contenir les RNA count. Car Azimuth va appliquer une normalization comme celui de la référence utilisée (Hao, Y., Hao, S., Andersen-Nissen, E., Mauck, W. M., Zheng, S., Butler, A., Lee, M. J., Wilk, A. J., Darby, C., Zager, M., Hoffman, P., Stoeckius, M., Papalexi, E., Mimitou, E. P., Jain, J., Srivastava, A., Stuart, T., Fleming, L. M., Yeung, B., … Satija, R., 2021).

```{r}
DefaultAssay(data) <- "RNA"
data <- RunAzimuth(data, reference = "pbmcref")
```

```{r}
DimPlot(data, group.by = "predicted.celltype.l2", label = TRUE, label.size = 3) + NoLegend()
```

check the score prediction

With SingleR
